Decapping activators Edc3 and Scd6 act redundantly with Dhh1 in post-transcriptional repression of starvation-induced pathways

Version 1

Abstract

Abstract

Degradation of many yeast mRNAs involves decapping by the Dcp1:Dcp2 complex. Previous studies on decapping activators Edc3 and Scd6 suggested their limited roles in mRNA decay. RNA-seq analysis of mutants lacking one or both proteins revealed that Scd6 and Edc3 have largely redundant activities in targeting numerous mRNAs for degradation that are masked in the single mutants. These transcripts also are frequently targeted by decapping activators Dhh1 and Pat1, and the collective evidence suggests that Scd6/Edc3 act interchangeably to recruit Dhh1 to Dcp2. Ribosome profiling shows that redundancy between Scd6 and Edc3 and their functional interactions with Dhh1 and Pat1 extend to translational repression of particular transcripts, including a cohort of poorly translated mRNAs displaying interdependent regulation by all four factors. Scd6/Edc3 also participate with Dhh1/Pat1 in post-transcriptional repression of proteins required for respiration and catabolism of alternative carbon sources, which are normally expressed only in limiting glucose. Simultaneously eliminating Scd6/Edc3 increases mitochondrial membrane potential and elevates metabolites of the tricarboxylic acid and glyoxylate cycles typically observed only during growth in low glucose. Thus, Scd6/Edc3 act redundantly, in parallel with Dhh1 and in cooperation with Pat1, to adjust gene expression to nutrient availability by controlling mRNA decapping and decay.

Introduction

Degradation of mRNA is a key aspect of gene expression that can be regulated in response to nutrient availability, cell stress, and developmental pathways in eukaryotic cells and also serves to eliminate defective mRNAs. A major pathway of cytoplasmic mRNA turnover involves truncating the poly(A) tail by the Ccr4/Not and Pan2/Pan3 complexes, followed by removal of the m7G cap by the Dcp1/Dcp2 decapping complex and 5’ to 3’ exonucleolytic degradation by Xrn1. The decapping complex is activated by factors that interact with low-complexity sequence motifs in the C-terminal tail (CTT) of the catalytic subunit Dcp2, including Edc3, Scd6, DEAD-box helicase Dhh1, and Pat1, which also interact with one another extensively. There is evidence that Pat1 is recruited to oligoadenylate tails remaining on mRNAs following partial deadenylation, in association with the Lsm1-Lsm7 complex, and activates decapping via interactions with other decapping activators and with the Dcp2 CTT itself (Parker 2012) (He and Jacobson 2023)

Genome-wide analysis of mRNA abundance (RNA-seq) in yeast mutants lacking Dhh1 or Pat1 suggested that only ∼40-50% of mRNAs up- or down-regulated in one mutant were also similarly dysregulated in the other mutant, suggesting that many mRNAs are targeted for degradation exclusively by Pat1 or Dhh1 (He et al. 2018). By also determining mRNA changes in a pat1Δdhh1Δ double mutant, we uncovered more extensive functional cooperation by Pat1 and Dhh1 in targeting ∼75% of mRNAs up-regulated in the single mutants, as a fraction of mRNAs were derepressed only in the double mutant (Vijjamarri et al. 2023a). The majority of mRNAs derepressed by dhh1Δ or pat1Δ mutations are likewise up-regulated in dcp2Δ cells (He et al. 2018), their derepression on eliminating Dhh1 is largely dependent on Dcp2, and they exhibit greater than average proportions of decapped mRNAs in WT cells but not in the dhh1Δ or pat1Δ mutant (Vijjamarri et al. 2023a)—all consistent with Dhh1/Pat1 targeting mRNAs for degradation via decapping. Moreover, more than half of the mRNAs up-regulated by dcp2Δ are likewise up-regulated by dhh1Δ and/or pat1Δ, whereas most of the remaining 45% of Dcp2-repressed transcripts are targeted primarily by the Upf factors responsible for nonsense-mediated mRNA decay (NMD) (Vijjamarri et al. 2023b), which represses a large cohort of natural NMD targets in WT cells (Celik et al. 2017).

The cumulative contributions of Dhh1 and Pat1 to mRNA decay via decapping is consistent with their independent interactions with distinct segments of the Dcp2 CTT (He et al. 2018), which appears to be direct for Pat1 but bridged by Edc3 or Scd6 for Dhh1. There is also evidence for distinct decapping complexes containing either Dhh1 or Pat1 in addition to Xrn1, Edc3 or Scd6 (He et al. 2022). However, Dhh1 occupancy determined by RIP-Seq analysis (Miller et al. 2018), tends to be elevated for mRNAs derepressed by either dhh1Δ (23) or pat1Δ (Vijjamarri et al. 2023a), consistent with Dhh1 contributing to degradation of many mRNAs targeted by Pat1. In accordance with overlapping roles for Pat1 and Dhh1, Pat1 binding to the Dcp2 CTT was required for degradation of certain Dhh1-targeted mRNAs only when Dhh1 recruitment to the CTT was compromised (He et al. 2022).

There is evidence that Edc3 is a common constituent of decapping complexes containing Xrn1 and one or more of the decapping activators Dhh1, Pat1, Scd6, and Upf1 (He et al. 2022). It is surprising therefore that Edc3 has been implicated in targeting only two transcripts for degradation, YRA1 and RPS28B (Badis et al. 2004; Dong et al. 2007). Edc3 shares sequence similarity with decapping activator Scd6 and both proteins contain an FDF motif shown to interact competitively with the Dhh1 homolog in animal systems (Tritschler et al. 2008; Tritschler et al. 2009). Yeast Edc3 contains two other segments besides the FDF motif that likely comprise a tripartite interaction surface for Dhh1 (Sharif et al. 2013). Edc3 and Scd6 also share N-terminal LSm domains that compete for binding to helical-leucine-rich (HLM) motifs in the yeast Dcp2 CTT (Fromm et al. 2012), an interaction that appears to activate decapping by overcoming an autoinhibitory element in Dcp2 (He and Jacobson 2015) (Paquette et al. 2018). Analyzing effects of deleting the Edc3 interaction site in the Dcp2-CTT on levels of several Dhh1-repressed mRNAs suggested that recruitment of Dhh1 to Dcp2 can be mediated interchangeably by Edc3 or Scd6 bound to the same site in the Dcp2 CTT (He et al. 2022). These findings, plus the fact that deleting SCD6 and EDC3 simultaneously confers a synthetic growth defect (Decourty et al. 2008), suggest that Scd6 and Edc3 might function redundantly in targeting specific mRNAs for decapping and degradation.

Although tethering Dhh1 or Scd6 enhances degradation of reporter mRNAs via Dcp1/Dcp2 (Carroll et al. 2011; Sweet et al. 2012; Zeidan et al. 2018), it is not well understood how these factors are targeted to specific native mRNAs. Dhh1 has been implicated in accelerating degradation associated with non-optimal codons in yeast mRNAs (Presnyak et al. 2015), being required for the rapid turnover conferred by suboptimal codons inserted in reporter mRNAs (Sweet et al. 2012). A queue of slowly elongating ribosomes upstream from non-optimal codons can be recognized by Dhh1, and overexpressing or tethering Dhh1 evokes ribosome stalling at non-optimal codons. Dhh1 association and Dhh1-dependent repression of mRNA abundance both correlate with codon non-optimality across the yeast transcriptome (Radhakrishnan et al. 2016). Structural evidence suggests that ribosomes stalled at suboptimal codons with an empty A site are recognized by Not5 binding to the ribosomal E-site to elicit Dhh1 recruitment and increased mRNA decapping and degradation (Buschauer et al. 2020). Dcp2, Pat1, and the deadenylase subunits of the Ccr4/Not complex also contribute to this process (Radhakrishnan et al. 2016; Webster et al. 2018). Despite these findings, the sets of mRNAs found to be most highly up-regulated in dhh1Δ and pat1Δ cells are not enriched for suboptimal codons and are generally well-translated in WT cells, suggesting that other features of these transcripts are responsible for their preferential targeting by Dhh1 or Pat1 for decapping/decay (Vijjamarri et al. 2023a).

In addition to regulating mRNA degradation, there is evidence that Pat1, Dhh1, and Scd6 can repress mRNA translation. Tethering Dhh1 or Scd6 represses translation of reporter mRNAs in dcp2Δ mutant cells where the tethered transcripts cannot be decapped and degraded (Carroll et al. 2011; Sweet et al. 2012; Zeidan et al. 2018), functioning either during elongation (Dhh1) or the initiation stage (Scd6) of translation. Deletion of Dhh1 and Pat1 simultaneously eliminated loss of bulk polysomes evoked by nutrient starvation and also increased initiation rates of certain mRNAs (Holmes et al. 2004) (Coller and Parker 2005) (Arribere et al. 2011). Supporting a direct role in repressing translation, overexpressing Dhh1 or Pat1 in nonstarved cells evoked polysome disassembly and reduced the initiation rate of specific mRNAs; and addition of Dhh1 (Coller and Parker 2005) or N-terminally truncated Pat1 (Nissan et al. 2010) to yeast extracts inhibited bulk translation and 48S preinitiation complex (PIC) assembly in vitro. Ribosome profiling studies of dhh1Δ, pat1Δ and pat1Δdhh1Δ mutants identified hundreds of mRNAs for particular genes that appear to be translationally repressed or activated by Dhh1 or Pat1 in nutrient-replete cells (Jungfleisch et al. 2017) (Radhakrishnan et al. 2016; Zeidan et al. 2018), which frequently involves cooperation between Dhh1 and Pat1 (Vijjamarri et al. 2023a). Interestingly, a fraction of mRNAs showed translational repression by Dhh1 that required Dcp2 and exhibit high proportions of decapped degradation intermediates in WT cells, suggesting translational repression via Dhh1-stimulated decapping that precedes mRNA turnover (Vijjamarri et al. 2023b).

Recently, we showed that Pat1 and Dhh1 function with the decapping enzyme in rich medium to repress the abundance or translation of numerous mRNAs encoding proteins required specifically in media containing an alternative carbon or nitrogen source (Vijjamarri et al. 2023a; Vijjamarri et al. 2023b). These include many of the mitochondrial proteins involved in oxidative phosphorylation (Ox. Phos.), and Pat1, Dhh1 and Dcp2 are required for the low mitochondrial membrane potential in glucose-replete cells expected from low electron transport chain (ETC) activity. Mutants lacking Dcp2, Dhh1, or Pat1 also display impaired mRNA turnover or elevated translation of mRNAs subject to carbon or nitrogen catabolite transcriptional repression in WT cells in rich medium. Related to this, Pat1 cooperates with Dhh1 (Hu et al. 2015) in repressing mRNAs encoding factors for autophagy and helps to suppress this pathway in rich medium. Pat1 further assists Dcp2 in post-transcriptional repression of mRNAs encoding cell adhesion proteins, critical for forming adhesion-linked chains of elongated cells able to penetrate substrates, thereby repressing invasive cell growth on nutrient-rich agar medium (Vijjamarri et al. 2023a; Vijjamarri et al. 2023b).

In this study, we employed RNA-Seq and ribosome profiling to determine whether Edc3 and Scd6 have largely redundant functions in targeting mRNAs for decapping and attendant degradation, and whether they functionally cooperate with Pat1 or Dhh1 in repressing the abundance or translation of particular mRNAs. We also sought to determine whether Edc3/Scd6 participate with Pat1/Dhh1 and the decapping enzyme in conferring post-transcriptional repression in rich medium of genes normally well-expressed only in nutrient-deprived cells. We find that a mutant lacking both Edc3 and Scd6 exhibits increased abundance of a cohort of mRNAs that are up-regulated only when both factors are eliminated, and which occurs without increased transcription of the cognate genes, signifying reduced decapping and degradation in mutant cells lacking both factors. These transcripts are also generally derepressed in dcp2Δ cells, and in cells lacking Pat1 or Dhh1, and show enrichment for Dhh1 occupancy. A strikingly similar pattern of mRNA derepression observed in the dhh1Δ and scd6Δedc3Δ mutants supports the notion that Edc3/Scd6 act redundantly to recruit Dhh1 to Dcp2 for activation of decapping (He et al. 2022). We also observed functional redundancy between Scd6 and Edc3 and cooperation with Dhh1/Pat1 in repressing translation of particular mRNAs, with evidence for interdependent repression by all four decapping factors. Importantly, Edc3/Scd6 also contribute to the post-transcriptional repression of proteins required for catabolism of non-preferred carbon or nitrogen sources on rich medium, including Ox. Phos. proteins, acting collectively to enhance glucose repression, maintain low-level mitochondrial electron transport and reduce levels of tricarboxylic acid (TCA) and glyoxylate cycle intermediates in glucose-replete cells. Together, our findings indicate that Edc3 and Scd6 have redundant functions in stimulating degradation or translational repression of transcripts targeted in parallel by Dhh1 and Pat1 to promote post-transcriptional repression in rich medium of factors normally up-regulated only in nutrient-deprived cells.

Methods

Yeast strains and plasmids

The following yeast strains were employed for all experiments: WT strain HFY114 (W303: MATa ade2-1 ura3-1 his3-11,15 trp1-1 leu2-3,112 can1-100) (He et al. 2003), scd6Δ strain SYY2352 (MATa ade2-1 ura3-1 his3-11,15 trp1-1 leu2-3,112 can1-100 scd6Δ::kanMX6) (He and Jacobson 2015), edc3Δ strain FZY862 (MATa ade2-1 ura3-1 his3-11,15 trp1-1 leu2-3,112 can1-100 edc3Δ::kanMX4), and scd6Δedc3Δ strain FZY858 (MATa ade2-1 ura3-1 his3-11,15 trp1-1 leu2-3,112 can1-100 scd6Δ::hphMX4 edc3Δ::kanMX4).

Strains FZY858 and FZY862 were generated by replacing chromosomal EDC3 in strains FZY855 and HFY114, respectively, with the edc3Δ::kanMX4 allele amplified by PCR from the chromosomal DNA of strain 255 (MATa his3Δ1 leu2Δ0 met15Δ0 ura3Δ0 edc3Δ::kanMX4) obtained from Research Genetics. DNA sequences up to 310 bp upstream and 340 bp downstream of the EDC3 coding sequence were included in the amplified fragment used for transformation of FZY855 and HFY114 to G418-resistance. FZY855 was derived from SYY2352 by transforming with marker-swap plasmid pAG32 (Goldstein and McCusker 1999) to replace scd6Δ::kanMX6 with scd6Δ::hphMX6, selecting for resistance to hygromycin and screening for loss of G418-resistance. The presence of all deletion alleles was verified by PCR analysis of chromosomal DNA. A complete list of yeast strains employed is given in Table S1.

Plasmids pLfz614-7 and pLfz635-5 contain the EDC3 CDS with 420 bp upstream and 490 bp downstream, on a ∼2.6 kb fragment, and plasmids pL615-5 and pLfz636-1 contain the SCD6 CDS with 500 bp upstream and 180 bp downstream on a ∼1.7 kb fragment, amplified by PCR from yeast genomic DNA and inserted between the XhoI and EcoRI sites of YCplac33 or YCplac111, respectively. The inserted yeast DNA fragments were verified by sequencing in their entirety. A complete list of plasmids employed is given in Table S2.

Cell spotting growth assays

Yeast transformants harboring plasmids containing EDC3, SCD6, or empty vector were grown to mid-logarithmic phase at 30°C in liquid synthetic complete medium (SC) without uracil (SC-U). Cultures were diluted to OD600 of 1.0 and 10-fold serial dilutions were spotted on agar medium of the same composition and incubated at 30°C or 37°C for 2 days.

Polysome profiling

Polysome profiling was conducted as described previously (Zeidan et al. 2018). In brief, strains were cultured in YPD medium at 30°C to mid-logarithmic phase (OD600 of ∼0.5-0.6). Fifteen A260 units of WCEs were resolved on a 10–50% sucrose gradient by centrifugation at 35,000 rpm. Gradients were fractionated with continuous scanning at 260 nm. Areas under the A260 tracings of polysome and monosome peaks were calculated using ImageJ software and used to calculate polysome to monosome (P/M) ratios in mutants and WT.

Ribosome footprint profiling

Duplicate cultures of yeast strains HFY114 (WT W303), SYY2352 (scd6Δ), FZY862 (edc3Δ), and FZY858 (scd6Δedc3Δ) were cultured in YPD medium at 30°C to OD600 ∼0.6-0.7. Yeast cells were quickly filtered, frozen in liquid nitrogen and stored at -80°C. Cells were lysed in a freezer mill and ribosome protected mRNA fragments (RPF) were isolated and used for cDNA library preparation as described previously (Vijjamarri et al. 2023b). Single-end 100 bp Illumina sequencing was performed by the National Heart, Lung and Blood Institute (NHLBI) DNA Sequencing and Genomics Core facility (Bethesda, MD).

RNA sequencing with spike-in normalization

The same lysates used for RPF library preparation were used to prepare RNA-Seq libraries after adding ERCC RNA Spike-In Control Mix 1 (Thermo Fisher Scientific, Cat. # 4456740). Total RNA was extracted from cell lysates using QIAzol Lysis reagent (Qiagen Cat. # 79306) and miRNeasy Mini Kit (Qiagen, Cat. 217004). Twenty µg of total RNA of each sample was subjected to RNase-free DNase I (Roche, Cat. # 04716728001) treatment and then processed with the RNA Clean and Concentrator Kit (Zymo, Cat. R1018). 2.4 µl of 1:100 diluted ERCC RNA Spike-In Control Mix 1 was added to 1.2 µg of each RNA sample and submitted to the NHLBI DNA Sequencing and Genomics Core facility (Bethesda, MD) for cDNA library preparation and Illumina sequencing.

Rpb1 ChIP-Seq with spike-in normalization

Triplicates of wild type and scd6Δedc3Δ yeast strains were cultured in rich YPD medium at 30°C to OD600=0.6∼0.8. Chromatin extracts were prepared from formaldehyde cross-linked cells as described previously (Qiu et al. 2016). S. pombe chromatin was added to each chromatin sample as a spike-in control prior to immunoprecipitation with Rpb1 antibodies, and ChIP-Seq DNA libraries were prepared as described previously (Zheng et al. 2023) subjected to 50 bp paired-end Illumina sequencing by the NHLBI DNA Sequencing and Genomics Core facility (Bethesda, MD) and analyzed as previously described (Zheng et al. 2023).

Western blot analysis

For the results in Fig. 6B-C, WCEs were prepared by trichloroacetic acid (TCA) extraction as previously described (Reid and Schatz 1982) and immunoblot analysis was conducted as described previously (Nanda et al. 2009). After electroblotting to nitrocellulose membranes (Bio-Rad 1620094), membranes were probed with antibodies against Atp20, Cox14, Pet10, Qcr8, Sdh4 (kindly provided by Dr. Nikolaus Pfanner), Cyb2 (kindly provided by Dr. Thomas Fox), Idh1 (Abnova, PAB19472), Cit2 (Antibodies-online.com, ABIN4889057), and Gcd6 (Bushman et al. 1993). Secondary antibodies employed were HRP-conjugated anti-rabbit (Cytiva, NA9340V), anti-mouse IgG (Cytiva, NA931V) and anti-goat IgG (Abnova, PAB29101). Detection was performed using enhanced chemiluminescence (ECL) Western Blotting Detection Reagent (Cytiva, RPN2016) and the Azure 200 gel imaging biosystem. NIH Image J was employed to analyze images for quantification. For analysis of Cox2 levels in Fig. 6D, total proteins were TCA-extracted as described previously (Rashida et al. 2021) and subjected to Western blot analysis as recently described (Niphadkar et al. 2024) using mouse monoclonal antibody MTCO2 (4B12A5) from Invitrogen.

TMT-MS analysis of global protein abundance

TMT-MS analysis was conducted as described previously (Vijjamarri et al. 2023a) with the following modifications. Replicate cultures of WT, edc3Δ, scd6Δ and scd6Δedc3Δ strains were cultured in YPD medium for ∼3 doublings to OD600 of ∼0.6, and harvested by centrifugation for 5 min at 3000 x g. Cells were resuspended in nuclease-free water, collected by centrifugation, and stored at -80°C. WCEs were prepared in freshly prepared 8M Urea, 25 mM triethylammonium-bicarbonate (TEAB; Thermo Scientific, 90114) by washing the cell pellets once and resuspending again in the extraction buffer, then vortexing with glass beads in the cold room. Lysates were clarified by centrifugation at 13,000 x g for 30 min and the quality of extracted proteins was assessed following SDS-PAGE using GelCode™ Blue Stain (Thermo Scientific, 24592) and quantified with the Pierce™ BCA Protein Assay Kit (Thermo Scientific, 23225). Lysates were stored at -80°C. Sample preparation, TMT-MS/MS (Zecha et al. 2019) and data analysis were performed at the IDeA National Resource for Quantitative Proteomics.

Measuring mitochondrial membrane potential

Precultures were grown in SC-Ura (to select for the URA3 plasmids) to OD600 of ∼3.0 and used to inoculate YPD medium at OD600 of 0.2. Cells were grown to OD600 of ∼0.6-0.8, incubated with 500 nM TMRM for 1 h and subjected to flow cytometry to measure dye fluorescence in individual cells as described previously (Vijjamarri et al. 2023b). In control samples, 50 µM FCCP was added to dissipate the membrane potential and reveal non-specific background fluorescence. Results are presented in arbitrary fluorescence units normalized to OD600 of the cultures.

Analysis of polar metabolites of intermediary metabolism

Yeast cell culturing. Four replicate 15 mL cultures in YEPD medium were prepared for each strain by inoculating with saturated overnight cultures to OD600 of 0.1, culturing with shaking at 30°C, and harvesting at OD600 of 0.6 by centrifugation in conical 15mL tubes for 3 min at 3000 x g in an Avanti J-HC centrifuge pre-cooled to -10°C. Cell pellets were resuspended in 1.8 mL of ice-cold PBS and centrifuged in 2 mL screw-cap tubes in a refrigerated microfuge for 30 s. Supernatants were decanted, tube rims blotted on tissue paper, and cell pellets frozen in dry ice/ethanol for 5 min and stored at -80C. Frozen cell pellets were shipped on dry ice to the NYU Metabolomics Core Resource Laboratory.

Extraction of metabolites. Frozen cell pellets were thawed on wet ice. Extraction buffer, consisting of 80% methanol (Fisher Scientific) and 500 nM metabolomics amino acid mix standard (Cambridge Isotope Laboratories, Inc.), was prepared and placed on dry ice. Samples were extracted by mixing cell pellets with extraction buffer at 10mg/mL (determined by sample OD measurements) in 2.0 mL screw-cap vials containing ∼100 µL of disruption beads (Research Products International, Mount Prospect, IL). Each sample was homogenized for 10 cycles on a bead blaster homogenizer (Benchmark Scientific, Edison, NJ), with each cycle consisting of 30 sec homogenization at 6 m/s followed by a 30 sec pause. Samples were centrifuged at 21,000 g for 3 min at 4°C. Aliquots of 450 µL were transferred to 1.5 mL tubes, dried in a Speedvac (Thermo Fisher, Waltham, MA), reconstituted in 50 µL of Optima LC/MS grade water (Fisher Scientific, Waltham, MA), sonicated for 2 min, and centrifuged at 21,000 g for 3 min at 4°C. Aliquots of 20 µL were transferred to LC vials containing glass inserts for analysis, and the remaining samples stored at -80°C.

LC-MS/MS with the hybrid metabolomics method. Samples were subjected to an LC-MS analysis to detect and quantify known peaks. Extraction of polar metabolites was carried out on each sample based on a previously described method (Jones et al. 2014). LC was conducted using a Millipore™ ZIC-pHILIC (2.1 x150 mm, 5 μm) column coupled to a Dionex Ultimate 3000™ system with gradient elution conducted at 25°C with a flow rate of 100 μL/min using the following buffers A) 10 mM ammonium carbonate in water, pH 9.0, and B) neat acetonitrile. The gradient profile was as follows; 80-20% B (0-30 min), 20-80% B (30-31 min), 80-80% B (31-42 min). Injection volume was set to 2 μL for all analyses with a 42 min total run time per injection.

MS was conducted by coupling the LC system to a Thermo Q Exactive HF™ mass spectrometer operating in heated electrospray ionization mode (HESI). Method duration was 30 min with a polarity switching data-dependent Top 5 method for both positive and negative modes. Spray voltage for both positive and negative modes was 3.5kV and capillary temperature was set to 320 °C with a sheath gas rate of 35, aux gas of 10, and max spray current of 100 μA. The full MS scan for both polarities utilized 120,000 resolution with an AGC target of 3e6 and a maximum IT of 100 ms, and the scan range was from 67-1000 m/z. Tandem MS spectra for both positive and negative mode used a resolution of 15,000, AGC target of 1e5, maximum IT of 50 ms, isolation window of 0.4 m/z, isolation offset of 0.1 m/z, fixed first mass of 50 m/z, and 3-way multiplexed normalized collision energies (nCE) of 10, 35, 80. The minimum AGC target was 1e4 with an intensity threshold of 2e5. All data were acquired in profile mode.

Relative quantification and statistical analyses of polar metabolites. The resulting ThermoTM RAW files were converted to SQLite format using an in-house python script to enable downstream peak detection and quantification. The available MS/MS spectra were first searched against the NIST17 MS/MS (Simon-Manso et al. 2013), METLIN (Smith et al. 2005) and respective Decoy spectral library databases using an in-house data analysis python script adapted from our previously described approach for metabolite identification false discovery rate control (FDR) (Wang et al. 2018; Wang et al. 2020). Metabolite peaks were extracted based on the theoretical m/z of the expected ion type, e.g., [M+H]+, with a 15 part-per-million (ppm) tolerance and a ± 0.2 min peak apex retention time tolerance within an initial retention time search window of ±0.5 min. For all the group-wise comparisons, t-tests were performed using the Python SciPy (1.5.4) (Virtanen et al. 2020) library to test for differences and generate statistics for the downstream analyses. For the pairwise t-tests, any metabolite with a p-value < 0.01 was considered significantly regulated (up-or down-) for prioritization in the subsequent analyses.

Coverage of a library of polar metabolites of major pathways of intermediary metabolism allowed detection of 130 of the 147 metabolites examined in at least 4 samples, and 91 detected in all 20 samples after background threshold correction. Instrument performance was assessed using the internal standards added to the samples during metabolite extraction and instrument mass accuracy was within tolerance (-2.0 ppm), LC column performance was stable (0.16 min RT range) and internal standard response variability was 13% across the samples. The resulting data were analyzed by principal components analysis (Fig. 7A), by constructing heatmaps with unsupervised hierarchical clustering of the imputed matrix values utilizing the R library pheatmap (1.0.12). GraphPad Prism 9 (9.4.1, GraphPad Software, San Diego, CA), and volcano plots (generated utilizing R library script Manhattanly (0.2.0)), in addition to the statistical comparisons summarized in Supplementary File S7.

13C6-glucose metabolic flux measurements

Three replicate cultures of each strain were grown in YP medium with 2% unlabeled glucose to OD600 of 0.6-0.7 and then shifted to YP with 1% unlabeled glucose and cultured for 20 min. 13C6-labelled glucose (1%) was then added and growth was continued for 8 min. Cells were collected, and metabolites were extracted, derivatized and subjected to mass spectrometry as described previously, with the addition of additional masses for assessing the label incorporation into the specific labelled intermediates that were analyzed (Walvekar et al. 2018).

ATP Measurements

Cells were cultured in YPD to OD600 of 0.6-0.7 and treated or untreated with 5 mM sodium azide for 30 min at 30°C. Five OD600 units of cells were harvested and resuspended in 300 µL ice-cold 5% trichloroacetic acid (TCA), incubated on ice for 15 min, and diluted 50-fold in 20 mM Tris-HCl (pH 7.5) to 0.1 % TCA, after which 10 µL was mixed with 90 µL of ATP Determination reaction mixture (Cat. # A22066, Thermo-Fisher Scientific) and incubated at room temperature for 5 min. Units of Firefly Luciferase were measured in a Berthold Centro XS3 LB 960 Luminometer, and ATP levels were calculated from an ATP standard curve generated using the same assay.

Additional data visualization and statistical analyses

Notched box-plots were constructed using a web-based tool at http://shiny.chemgrid.org/boxplotr/. In all such plots, the upper and lower boxes contain the 2nd and 3rd quartiles and the band gives the median. If the notches in two plots do not overlap, there is roughly 95% confidence that their medians differ. Density scatter plots, Venn diagrams, significance testing of gene set overlaps in Venn diagrams using the hypergeometric distribution, hierarchical clustering analysis, and gene ontology (GO) analysis all were conducted as described previously (Vijjamarri et al. 2023a).

Results

Evidence that Scd6 and Edc3 functionally cooperate to control the abundance of many individual mRNAs

To determine whether Scd6 and Edc3 function redundantly in post-transcriptional control of gene expression, we constructed a scd6Δedc3Δ double mutant isogenic to the scd6Δ and edc3Δ single mutants we examined previously (Zeidan et al. 2018). Only the double mutant exhibits a marked slow-growth (Slg-) phenotype on synthetic complete medium (SC), which was largely complemented by introducing either SCD6 or EDC3 on a single copy plasmid (Supplemental Fig. S1A). Analysis of polysome assembly revealed a ∼40% reduction in ratio of polysomes to monosomes (P/M) in the scd6Δedc3Δ double mutant, whereas the single mutants showed little (scd6Δ) or no (edc3Δ) reduction in bulk translation by this assay (Fig. S1B). These results suggest that Scd6 and Edc3 function redundantly to carry out one or more functions required for WT levels of bulk translation and cell growth in nutrient replete cells.

To examine the effects of the scd6Δ, edc3Δ, and scd6Δedc3Δ mutations on the abundance and translation of individual mRNAs, we conducted RNA-Seq and ribosome profiling (Ribo-Seq) of the mutant and WT strains following growth in liquid rich medium (YPD) at 30°C (processed data compiled in supplementary File S1). Ribo-Seq entails deep-sequencing of ribosome-protected fragments (RPFs, or ribosome footprints), and cycloheximide was added to the lysates to arrest elongating ribosomes on the mRNA following cell breakage. The ratio of RPF sequencing reads summed over the coding sequences (CDS) to the total mRNA reads from RNA-Seq for the corresponding transcript provides a measure of translational efficiency (TE) for each mRNA (Ingolia et al. 2009). The ribosome profiling and RNA-Seq results between two biological replicates for each strain were highly reproducible with Pearson correlation coefficients (r) ranging between 0.95-1.0 for different pairwise comparisons of replicates (Fig. S2A-B). We employed DESeq2 (Love et al. 2014) to identify statistically significant differences in relative mRNA abundance, RPF abundance, or TE for all expressed mRNAs between WT and mutant strains (see Methods for details).

Analysis of the RNA-Seq results identified 81 mRNAs that were significantly derepressed in the edc3Δ mutant vs. WT by >1.5-fold at a false discovery rate (FDR) of <0.05 (mRNA_up_e3; Fig. S3A), and 123 mRNAs reduced in abundance by edc3Δ by >1.5-fold at the same FDR (mRNA_dn_e3, Fig. S3B). Applying the same criteria, only 14 mRNAs were up-regulated and only 34 down-regulated by the scd6Δ single mutation (mRNA_up_s6 & mRNA_dn_s6, Fig. S3A-B). Importantly, much greater numbers of mRNAs were dysregulated by the scd6Δedc3Δ double mutation: 741 in the mRNA_up_s6,e3 group and 793 in the mRNA_dn_s6,e3 (Fig. S3A-B), indicating that the two factors have highly overlapping functions in controlling mRNA abundance with either one sufficient for nearly WT levels of most mRNAs dysregulated in the double mutant. CIT1 mRNA, encoding a mitochondrial enzyme of the TCA cycle is representative of a transcript significantly up-regulated in both mRNA and RPF abundance only in the scd6Δedc3Δ double mutant (Fig. S3C).

Many yeast mutants with Slg- phenotypes, including pat1Δ, dhh1Δ, and dcp2Δ deletion mutants, exhibit altered expression of most mRNAs belonging to the Environmental Stress Response (ESR) (O’Duibhir et al. 2014), which includes ∼300 induced (iESR) and ∼600 repressed (rESR) mRNAs dysregulated in WT cells by various stresses (Gasch et al. 2000). In keeping with its Slg- phenotype, the scd6Δedc3Δ mutation conferred a marked reduction in median expression of rESR mRNAs, and increased expression of the iESR mRNAs, which exceeded in magnitude the changes observed for the slowest growing yeast deletion mutants analyzed previously (O’Duibhir et al. 2014) (Fig. S3D-E). The two single mutations, by contrast, conferred much smaller changes in ESR mRNAs (Fig. S3D-E). (In all box plots, when notches do not overlap between adjacent boxes, their two medians differ with 95% confidence; and when notches do not overlap 0 in log2 plots, the median differs significantly from that of all mRNAs, which is invariably close to 1.0.) Consistent with these results, the transcripts up-regulated in the double mutant are enriched for iESR mRNAs (Fig. 1A), making it possible that the 187 iESR transcripts up-regulated in this strain are responding indirectly to cell stress. The remaining 75% of the mRNAs up-regulated in scd6Δedc3Δ cells are not iESR mRNAs however (Fig. 1A), suggesting that the increased abundance of these 554 transcripts arises from eliminating Edc3/Scd6 functions in mRNA decay. Below, we excluded the ESR mRNAs from analyses of mRNA changes in an effort to focus on mRNAs whose abundance is controlled directly by Scd6 and Edc3.

Most mRNAs up-regulated in the scd6Δedc3Δ mutant are not iESR transcripts and exhibit Scd6/Edc3 functional redundancy in repression of transcript abundance.

(A) Venn diagram of overlap between the 741 mRNAs up-regulated in the scd6Δedc3Δ mutant vs. WT and the 283 induced ESR mRNAs, indicating fold-enrichment and P value of overlap determined by the hypergeometric distribution. (B) Venn diagram of overlaps involving all 591 non-iESR mRNAs derepressed in abundance by either the scd6Δ, edc3Δ, or scd6Δedc3Δ mutations. (C) Notched box-plot analyses of log2 changes in mRNA abundance (log2ΔmRNA) determined by DESeq2 analysis between the indicated mutants vs. WT for mRNAs belonging to the specified sectors of the Venn diagram in (B). Un-logged median values are indicated at the bottom. (D) Gene browser image for MDH2 presented as in Fig. S3C.

Considering only mRNAs not governed by the ESR, we identified 591 non-iESR mRNAs significantly up-regulated in any of the three mutants (Fig. 1B). Examining the ΔRNA values for the majority fraction of 521 transcripts up-regulated only in the double mutant confirms little change in median abundance in each single mutant and a strong synthetic derepression in the double mutant (Figs. 1B & 1C sector (iii))—as expected for redundant repressive functions of Scd6 and Edc3. MDH2, encoding an enzyme of the glyoxylate cycle, exemplifies a non-iESR transcript up-regulated in both mRNA and RPF abundance exclusively in the double mutant (Fig. 1D). The small fraction of 27 mRNAs significantly derepressed in both the edc3Δ single mutant and double mutant exhibits only slightly elevated median abundance in the scd6Δ single mutant, and only slightly greater derepression in the double mutant vs. edc3Δ single mutant (Fig. 1B & 1C sector (ii)), indicating a minimal repressive contribution by Scd6. Interestingly, the small set of 37 mRNAs significantly derepressed only in the edc3Δ single mutant shows reduced rather than increased abundance in the scd6Δ single mutant, and lower derepression in the double mutant vs. the edc3Δ single mutant (Fig. 1B & C sector (i)), suggesting that Scd6 enhances rather than represses these mRNAs, especially in edc3Δ cells.

Evidence that Dhh1 and Pat1 functionally cooperate with Scd6/Edc3 in repressing mRNA abundance

We recently identified a large group of 1018 non-iESR mRNAs derepressed by either the pat1Δ, dhh1Δ, or pat1Δdhh1Δ mutations (Vijjamarri et al. 2023a). Importantly, this group is highly enriched for the 591 mRNAs derepressed by the scd6Δ, edc3Δ, or scd6Δedc3Δ mutations (Fig. 2A). Indeed, ∼70% of the transcripts derepressed by scd6Δ/edc3Δ are also derepressed in one of the three pat1Δ/dhh1Δ mutants (Fig. 2A, sector (ii)), showing comparable up-regulation in the scd6Δedc3Δ, pat1Δ, and dhh1Δ mutants, but little change in the scd6Δ and edc3Δ single mutants (Fig. 2B(ii)). These mRNAs generally exhibit the greatest derepression in the pat1Δdhh1Δ double mutant, indicating cumulative contributions of Dhh1 and Pat1 to their repression, in contrast to the largely redundant roles played by Scd6 and Edc3 in repressing these mRNAs. Thus, efficient repression of these 411 mRNAs requires a combination of Pat1, Dhh1 and either Edc3 or Scd6. As expected, the smaller group of 180 mRNAs significantly derepressed exclusively by scd6Δ/edc3Δ (sector (i) of Fig. 2A) show strong derepression only in the scd6Δedc3Δ double mutant; but they are appreciably up-regulated by dhh1Δ while being largely unaffected by pat1Δ (Fig. 2B, sector (i)). Finally, the majority group of 607 mRNAs significantly derepressed by only pat1Δ or dhh1Δ display the largest increases in the two mutants lacking Pat1, but display smaller and similar increases in response to the scd6Δ/edc3Δ and dhh1Δ mutations (Figs. 2A & 2B, sector (iii)). Overall, these findings suggest that Dhh1 and Pat1 both contribute to repressing the majority of mRNAs repressed by Scd6/Edc3, but with Dhh1 contributing more extensively than Pat1.

Most mRNAs up-regulated in the scd6Δedc3Δ mutant are also derepressed by dhh1Δ and pat1Δ.

(A) Venn diagram of overlap between all 591 non-iESR mRNAs derepressed in abundance by the scd6Δ, edc3Δ, or scd6Δedc3Δ mutations (from Fig. 1B) and 1018 non-iESR mRNAs up-regulated by the dhh1Δ, pat1Δ, or pat1Δdhh1Δ mutations identified previously (Vijjamarri et al. 2023a). (B) Notched box-plots of log2ΔmRNA between the indicated mutants vs. WT for mRNAs in the three sectors specified in (A). (C) Hierarchical clustering analysis of log2ΔmRNA values conferred by the indicated mutations vs. WT for 784 of the 794 mRNAs up-or down-regulated in dhh1Δ vs. WT cells for which RNA-Seq data was obtained in all four strains and with log2ΔmRNA values >-5 and <5, conducted with R heatmap.2 function from R ‘gplots’ library, using default hclust hierarchical clustering algorithm. Spearman coefficients (ρ) and associated P values are given for the indicated correlation analyses. (D) Venn diagram of overlaps between the 591 non-iESR mRNAs up-regulated by the scd6Δ/edc3Δ mutations vs. WT (from Fig. 1B) and the indicated 431 and 590 non-iESR mRNAs up-regulated by dhh1Δ or pat1Δ vs. WT, respectively, identified previously (Vijjamarri et al. 2023a). (E) Notched box-plots of log2ΔmRNA between the indicated mutants vs. WT for the 431 non-iESR mRNAs up-regulated by dhh1Δ vs. WT shown in panel (D).

Further evidence for this last point comes from a k-means cluster analysis of mRNA changes in different mutants for the group of 431 non-iESR mRNAs derepressed by dhh1Δ, which revealed a stronger correlation between the changes conferred by scd6Δ/edc3Δ and dhh1Δ (ρ = 0.90) than between pat1Δ with either dhh1Δ (ρ = 0.74) or scd6Δ/edc3Δ (ρ = 0.71) (Fig. 2C). Consistent with this, the scd6Δ/edc3Δ mutations up-regulate a considerably larger proportion of the mRNAs significantly derepressed by dhh1Δ (∼70%) vs. those derepressed by pat1Δ (∼41%) (Fig. 2D). Furthermore, the majority of mRNAs up-regulated by dhh1Δ exhibit redundant repression by Scd6 and Edc3, showing marked derepression comparable to that given by dhh1Δ itself only in the scd6Δ/edc3Δ double mutant (Fig. 2E). These findings are consistent with the involvement of Edc3/Scd6 in the degradation of most mRNAs targeted by Dhh1, which are more variably and less extensively repressed by Pat1.

Evidence that Scd6/Edc3 repress mRNA abundance by enhancing decapping and degradation rather than suppressing transcription

We next examined whether the mRNAs significantly repressed by Edc3/Scd6 are also regulated by the decapping enzyme Dcp1/Dcp2. Supporting this, cluster analysis of mRNA changes revealed that the majority of non-iESR mRNAs derepressed in the scd6Δedc3Δ double mutant are also derepressed by dcp2Δ (Fig. 3A, blue colors in cols. 1 & 3), with a strong correlation between the abundance changes conferred by these mutations relative to WT (ρ=0.74, P≈0). In addition, dcp2Δ increased the median abundance of this group of mRNAs similarly to that given by scd6Δedc3Δ (Fig. 3B, cols. 1 & 4). These findings implicate decapping by Dcp1:Dcp2 as an important driver of the repression of mRNA levels directed by Scd6/Edc3. The cluster analysis in Fig. 3A once again reveals greater similarity between the mRNA changes conferred by scd6Δedc3Δ and those given by dhh1Δ (ρ = 0.90) versus pat1Δ (ρ = 0.69); and pat1Δ also confers a smaller median reduction than dhh1Δ of the Scd6/Edc3-repressed mRNAs (Fig. 3B).

Evidence that impaired decapping rather than increased transcription is a key driver of changes in mRNA abundance in scd6Δedc3Δ cells.

(A) Hierarchical clustering analysis of log2ΔmRNA values conferred by the indicated mutations vs. WT for 741 of the 1052 mRNAs up-or down-regulated in scd6Δedc3Δ vs. WT cells for which RNA-Seq data was obtained in all five strains and with log2ΔmRNA values >-5 and <5, conducted as in Fig. 2C, showing Spearman coefficients and P values for indicated correlations. (B) Notched box-plots of log2ΔmRNA between the indicated mutants vs. WT for the 554 non-iESR mRNAs up-regulated by scd6Δedc3Δ vs. WT (shown in Fig. 1B). (C) Ratios of capped to total mRNA abundance in TPMs (Relative C/T) in WT or dhh1Δ cells plotted for all mRNAs or the 554 non-iESR mRNA_ up_s6,e3 or 526 non-rESR mRNA_down_s6,e3 transcripts dysregulated by scd6Δedc3Δ vs. WT. (D) Notched box-plots of the codon-protection index (CPI) for all mRNAs or for the sets of mRNAs up-or down-regulated by scd6Δedc3Δ vs. WT, including or excluding ESR transcripts, as indicated. (E) Notched box-plots showing log2 changes in absolute mRNA abundance from ERCC spike-in normalized RNA-Seq (left) or absolute Rpb1 occupancies averaged over the CDSs from S. pombe chromatin spike-in normalized Rpb1 ChIP-Seq (right) in scd6Δedc3Δ vs. WT cells for all mRNAs or the 554 or 526 non-ESR mRNAs up-or down-regulated, respectively, by scd6Δedc3Δ vs. WT.

Additional support for the conclusion that Scd6/Edc3 target mRNAs for decapping came from evidence that the mRNAs up-regulated in the scd6Δedc3Δ mutant tend to accumulate in WT cells as decapped isoforms. It is known that following decapping by Dcp1/Dcp2 mRNAs frequently undergo 5’ to 3’ decay co-translationally, with Xrn1 following behind the last translating ribosome loaded prior to decapping, and such decapped intermediates associated with ribosomes account for ∼12% of the mRNAs in WT cells (Pelechano et al. 2015). We reasoned that mRNAs preferentially targeted by Scd6/Edc3 for decapping and attendant degradation by Xrn1 should exhibit a greater than average proportion of decapped intermediates in WT cells. To test this, we interrogated our previous cap analysis of gene expression (CAGE) data providing the abundance of all capped mRNA 5’ ends and compared it to the abundance of total RNA transcripts determined by RNA-Seq conducted in parallel on biological replicates of mRNA samples prepared from the WT and dhh1Δ strains described above (Vijjamarri et al. 2023a). Transcript numbers per million reads (TPMs) from CAGE (C) and RNA-Seq (T) were determined and C/T ratios calculated as a proxy for the proportion of capped molecules for each transcript (File S1). (Because the CAGE and RNA-Seq data were normalized separately, the C/T ratios are relative, not absolute, proportions of capped transcripts.) Importantly, the C/T ratios are lower in WT cells for the 448 non-iESR mRNAs up-regulated in the scd6Δedc3Δ double mutant (from Fig. 1B) compared to all expressed mRNAs for which CAGE data was obtained (Fig. 3C, cols. 1 & 5). By contrast, the C/T ratios are higher than average for the non-rESR mRNAs that are down-regulated in relative abundance in scd6Δedc3Δ cells, as expected for an unusually low degree of decapping in WT cells (Fig. 3C, cols 3 & 5).

Importantly, these C/T ratios were elevated above their values in WT cells to nearly the same level for all groups of mRNAs, in the manner expected from eliminating Dhh1-stimulated decapping (Fig. 3C, cf. cols. 2,4,6). These results are consistent with the notion that impaired Dhh1-stimulated decapping and attendant 5’ to 3’ degradation is an important driver of increased mRNA abundance in the scd6Δedc3Δ mutant.

Independent evidence that derepression of mRNA abundance in scd6Δedc3Δ cells results from reduced decapping was provided by analyzing the codon protection indices (CPI) of the mRNA_up_s6,e3 transcripts, an indicator of co-translational decay by Xrn1. Decapped degradation intermediates exhibit three-nucleotide periodicity generated by precise Xrn1 cleavage up to the last translating ribosome at the 5’ end of mRNA, and the CPI quantifies the prevalence of such intermediates for each mRNA (Pelechano et al. 2015). The results in Fig. 3D show that the mRNA_up_s6,e3 transcripts (including or excluding ESR mRNAs) exhibit higher than average median CPIs, indicating a greater than average involvement of decapping and co-translational degradation by Xrn1 in their decay, whereas mRNA_dn_s6,e3 transcripts exhibit lower than average CPI values, consistent with an alternative degradation pathway controlling their abundance.

To determine whether increased transcription contributes to increased abundance of any mRNAs derepressed by pat1Δ or dhh1Δ, we performed ChIP-Seq analysis of Rpb1 to measure RNA Polymerase II (Pol II) occupancies averaged across the CDS of every gene. We obtained highly reproducible results for the three biological replicates analyzed for each strain (Fig. S2D & S4A). To quantify absolute changes in Pol II occupancies, S. pombe chromatin was added as a spike-in to each S. cerevisiae chromatin sample prior to immunoprecipitation of Rpb1. To measure absolute changes in mRNA abundance, we re-analyzed the RNA-Seq data taking into account the recovery of External RNA Controls Consortium (ERCC) transcripts that had been spiked-in prior to the replicate total RNA samples prior to preparation of the cDNA libraries. The results for ERCC-normalized RNA-Seq were highly reproducible among the replicates (Fig. S2C).

Considering all expressed mRNAs, the spike-in normalized RNA-Seq data (File S3) revealed a 46% increase in median total mRNA abundance in the scd6Δedc3Δ mutant vs. WT that was accompanied by only a 4% increase in median Rpb1 occupancy in the double mutant (Figs. 3E_alt, cols. 3 & 6), indicating that a reduced rate of mRNA decay associated with little change in transcription rate across the transcriptome leads to a net increase in total mRNA levels in the double mutant. Interrogating the two groups of non-ESR mRNAs found to be up- or down-regulated by scd6Δedc3 revealed a 2.79-fold increase in spike-in normalized median mRNA abundance for the non-iESR mRNA_up_s6,e3 group that was associated with only a 1.2-fold increase in normalized Rpb1 occupancies, indicating a small contribution of increased transcription to the elevated mRNA abundance of these transcripts in the double mutant. The down-regulated non-rESR mRNA_dn_s6,e3 transcripts showed 12% and 7% decreases in spike-in normalized median mRNA abundance and Rpb1 occupancies, respectively, indicating a larger contribution of transcriptional changes to the diminished transcript levels of these mRNAs in the double mutant. It should be noted that spike-in normalization of the RNA-Seq data reveals that the group of mRNAs up-regulated in the scd6Δedc3Δ mutant show larger increases in absolute compared to relative transcript levels (2.79-fold vs. 1.84-fold median increases) but show smaller absolute vs. relative reductions of the group of down-regulated transcripts (0.88-fold vs. 0.57-fold) (Fig. S4B, cf. cols. 1 vs. 4 and 2 vs. 5). This reflects the fact that the absolute changes conferred by scd6Δedc3 for most transcripts involve the increased abundance expected from reduced decapping/decay in cells lacking Scd6/Edc3 (Fig. S4B, col. 3).

Codon nonoptimality has been linked with Dhh1-mediated mRNA decay partly by demonstrating that the sTAI values of mRNAs, which quantify their overall codon optimality (Sabi and Tuller 2014), are inversely correlated with the changes in mRNA abundance observed in dhh1Δ versus WT cells (Radhakrishnan et al. 2016). Similarly, analyzing the mRNA changes conferred by the scd6Δedc3Δ mutation for all non-ESR transcripts reveals a small but statistically significant negative correlation with sTAI values (Pearson r of -0.085, P = 2 x 10-9) similar to that observed for the dhh1Δ mutation (r = -0.071, P = 6 x 10-6) (Fig. S5A, cyan vs. orange), indicating a tendency for mRNAs with lower codon optimality/sTAI scores to show greater increases in relative abundance in response to scd6Δedc3Δ or dhh1Δ. However, the group of 591 non-iESR mRNAs most significantly up-regulated by the scd6Δ/edc3Δ mutations (from Fig. 1B) have a median sTAI value (0.35) nearly identical to that of all nonESR mRNAs, which is also the case for the group of 1018 mRNAs derepressed by the pat1Δ or dhh1 mutations (Vijjamarri et al. 2023a) (Fig. S5B). The same results were obtained for other metrics of codon optimality, tAI and average CSC (Fig. S5B), suggesting that poor codon optimality is not a key property defining the mRNAs repressed in abundance most extensively by these decapping activators in WT cells. It has also been proposed that competition between translation initiation and mRNA decay, rather than codon optimality and elongation, is a major determinant of mRNA stability in yeast (Muhlrad et al. 1995; LaGrandeur and Parker 1999; Schwartz and Parker 1999; Chan et al. 2018). However, the mRNAs derepressed in the scd6Δ/edc3Δ or pat1Δ/dhh1 mutants have slightly greater than average median translational efficiencies (TEs) in WT cells (Fig. S5C), as determined by ribosome profiling experiments described below. Thus, mRNA features besides slow rates of translation initiation or elongation likely dictate preferential targeting by these four decapping activators.

Scd6 and Edc3 function redundantly and cooperate more extensively with Dhh1 than Pat1 in controlling the translation of individual mRNAs

The changes in mRNA abundance were highly correlated with changes in RPF abundance for the scd6Δedc3Δ mutant compared to WT for all expressed transcripts, with an r value of 0.81 (Fig. S2F), providing strong mutual validation of the RNA-Seq and Ribo-Seq data for these strains. Nevertheless, the fact that this correlation is weaker than that observed between biological replicates of RNA-Seq or Ribo-Seq data for each strain (r values >0.95, Fig. S2A-B), suggests that certain mRNAs exhibit altered translational efficiencies in the double mutant. To identify specific mRNAs showing evidence of translational control by Scd6 and Edc3, we used DESeq2 to analyze our Ribo-Seq data on the edc3Δ, scd6Δ and scd6Δedc3Δ mutants, comparing each to WT. We identified 42 mRNAs showing TE increases of >1.41-fold at FDR < 0.10 in the scd6Δ mutant, but only one such mRNA in the edc3Δ strain. Importantly, a larger number of mRNAs (N=184) were identified in the scd6Δedc3Δ double mutant (Fig. 4A), suggesting that Scd6 and Edc3 have overlapping functions in repressing the translation of particular mRNAs. Indeed, a group of 169 mRNAs that are translationally derepressed exclusively in the scd6Δedc3Δ double mutant (TE_up_s6,e3) exhibit only modest TE increases in the two single mutants compared to their marked derepression in the double mutant (Fig. 4A-B sector (iii)). SPI1, encoding a cell wall protein, is a representative transcript exhibiting increased TE only in the double mutant, displaying increased mRNA abundance coupled with an even greater increase in RPF abundance, both exclusively in the double mutant, to confer the observed TE increase of 3.2-fold (Fig. 4C). As expected, the 28 mRNAs showing substantial TE increases only in scd6Δ cells show no increase in median TE in the edc3Δ strain (Fig. 4A-B sector (i)), and thus appear to be translationally repressed by Scd6 alone.

Most mRNAs translationally up-regulated in scd6Δedc3Δ cells exhibit Scd6/Edc3 functional redundancy for repressing TE and show correlated changes in TE and protein abundance.

(A) Venn diagram of overlap between the 184 and 42 mRNAs in the TE_up groups identified in the scd6Δedc3Δ vs. WT or scd6Δ vs. WT comparisons, respectively. (B) Notched box-plots of log2ΔTE values between the indicated mutants vs. WT for the mRNAs belonging to sectors (i) or (iii) of the diagram in (A). (C) Gene browser image for SPI1 presented as in Fig. S3C, except also giving the TE changes for each mutant vs. WT on the lower right. (D) Density scatterplot of log2ΔRPF values measured by ribosome profiling vs. log2Δprotein values measured by TMT-MS for 4339 mRNAs for which data were obtained in both analyses, indicating the Pearson correlation coefficient (r) and P-value of the correlation. (E-F) Notched box-plots of log2Δprotein values from TMT-MS analysis between the scd6Δedc3Δ mutant vs. WT for the 843 and 839 mRNAs belonging to the RPF_up or RPF_down groups, respectively (D), or the 184 and 152 TE_up or TE_down mRNA groups (E) determined for the scd6Δedc3Δ mutant vs. WT, or for all mRNAs, for which TMT-MS data was obtained.

We asked next whether the increased ribosome occupancies observed in the scd6Δedc3Δ double mutant are generally associated with increased synthesis of the encoded proteins. To this end, we carried out TMT mass spectrometry (TMT-MS) of total cellular proteins to obtain ratios of peptide abundance in the scd6Δedc3Δ double and WT strain mutant for >4000 different proteins, obtaining highly correlated results from three biological replicates (Fig. S2E). Importantly, we observed significant correlations between these changes in protein abundance and changes in ribosome occupancies (RPFs) for mRNAs across the translatome (Fig. 4D). Moreover, the groups of mRNAs showing significantly increased or decreased RPFs in the scd6Δedc3Δ double mutant vs. WT (>1.5-fold, FDR<0.05, Ribo_up and Ribo_dn) likewise exhibit increased or decreased median protein abundance determined by TMT-MS (Fig. 4E). These results suggest that increased ribosome occupancies measured by Ribo-Seq, which could occur by increases in mRNA, TE, or both, are generally associated with increased synthesis of the encoded proteins in the scd6Δedc3Δ double mutant.

It was possible that translational repression by Edc3/Scd6 generally occurs by slowing the rate of elongation, leading to increased ribosome densities (RPF/mRNA ratios, ie. calculated TEs) in WT cells and decreased TE values in the scd6Δedc3Δ mutant that would be associated with increased protein expression in the mutant. In this scenario, changes in TEs would be inversely associated with changes in protein expression. At odds with this possibility, the groups of mRNAs showing increased or decreased TEs in the double mutant vs. WT (TE_up_s6,e3 and TE_dn_s6,e3 transcripts defined above) also exhibit increased or decreased protein expression in scd6Δedc3Δ vs. WT cells (Fig. 4F). These findings imply that Scd6/Edc3 generally influence translation at the level of initiation rather than elongation.

We previously identified a group of 274 mRNAs whose TEs are derepressed by the same criteria employed here (>1.41-fold at FDR < 0.10) in either the pat1Δ, dhh1Δ, or pat1Δdhh1Δ mutants versus WT (Vijjamarri et al. 2023a). Interestingly, we found a significant overlap between this group of mRNAs and those translationally derepressed in either the single or double scd6Δ/edc3Δ mutants identified here (Fig. 5A). The 76 mRNAs common to both groups show considerably larger TE increases in the scd6Δedc3Δ double mutant compared to the 136 transcripts whose TE was up-regulated only in the scd6Δ/edc3Δ mutants (Fig. 5A-B sector (ii) vs. sector (i)). The former 76 mRNAs also show marked TE increases in the dhh1Δ, pat1Δ, and pat1Δdhh1Δ mutants (Fig. 5B(ii)), indicating that all four decapping activators are required for efficient translational repression of these mRNAs in WT cells. The 136 mRNAs significantly derepressed only in the scd6Δ/edc3Δ mutants show a moderate increase in median TE in dhh1Δ cells, but little response to pat1Δ (Fig. 5B(i)). The 198 mRNAs substantially derepressed only in the pat1Δ/dhh1Δ mutants show only a small TE increase in the scd6Δedc3Δ strain, a somewhat greater response to the pat1Δ vs. dhh1Δ single mutation, and cumulative TE increases in the pat1Δdhh1Δ double mutant (Fig. 5B(iii)). In summary, similar to our findings for repression of mRNA abundance, efficient translational repression of certain mRNAs requires the combined functions of Pat1, Dhh1, and either Scd6 or Edc3, whereas other mRNAs are translationally repressed primarily by either Edc3/Scd6 or Pat1, with appreciable contributions from Dhh1 for both of these latter groups.

Most mRNAs translationally up-regulated in the scd6Δedc3Δ mutant are also translationally derepressed by dhh1Δ or pat1Δ.

(A) Venn diagram of overlap between all 212 mRNAs translationally derepressed by scd6Δ, edc3Δ, or scd6Δedc3Δ (defined in Fig. 4A) or all 274 mRNAs translationally derepressed by dhh1Δ, pat1Δ, or pat1Δdhh1Δ vs. WT identified previously (Vijjamarri et al. 2023a). (B) Notched box-plots of log2ΔTE values between the indicated mutants vs. WT for the mRNAs belonging to the specified sectors of the diagram in (A). (C) Hierarchical clustering analysis of log2ΔTE values conferred by the indicated mutations vs. WT for 222 of the 336 mRNAs translationally up- or down-regulated in scd6Δedc3Δ vs. WT cells for which RNA-Seq and Ribo-Seq data were obtained in all five strains and with log2ΔTE values >-5 and <5 conducted as in Fig. 2C, including the Spearman coefficients (ρ) and P values for the indicated correlations. (D) Notched box-plots of log2ΔTE values between the indicated mutants vs. WT for the 54 mRNAs showing >2-fold TE increases conferred by both scd6Δedc3Δ and pat1Δdhh1Δ mutations vs. WT. (E) Relative Dhh1 occupancies from the Dhh1 RIP-seq experiments of Miller et al (2018) for the 212 and 274 mRNAs identified as TE_up in the scd6Δ, edc3Δ, or scd6Δedc3Δ mutants, or the dhh1Δ, pat1Δ, or pat1Δdhh1Δ mutants, vs. WT, respectively (cols 1-2), or for the 591 and 1018 mRNAs identified as mRNA_up in either the scd6Δ, edc3Δ, or scd6Δedc3Δ mutants, or the dhh1Δ, pat1Δ, or pat1Δdhh1Δ mutants, vs. WT, respectively (cols. 3-4).

Greater cooperation of Scd6/Edc3 with Dhh1 vs. Pat1 in controlling translation is also revealed by clustering analysis of TE changes in different mutants for the mRNAs exhibiting significantly increased or decreased TEs in the scd6Δ/edc3Δ mutants, which reveals a greater similarity between TE changes conferred by scd6Δedc3Δ vs. dhh1Δ (ρ = 0.82) compared to scd6Δedc3Δ vs. pat1Δ (ρ = 0.68) or scd6Δedc3Δ vs. pat1Δdhh1Δ (ρ = 0.69) cells (Fig. 5C). However, as noted above, the group showing the strongest derepression of TEs in the scd6Δedc3Δ double mutant also tends to be highly derepressed in all three of the other mutants lacking Dhh1 or Pat1 (bracketed mRNAs at the top of Fig. 5C). This behavior is illustrated in Fig. 5D for a group of 54 mRNAs whose TEs are derepressed by 2.0-fold or more in both the scd6Δedc3Δ and pat1Δdhh1Δ double mutants, displaying marked TE increases in the scd6Δedc3Δ, pat1Δ, dhh1Δ, and pat1Δdhh1Δ double mutants, but not in the scd6Δ and edc3Δ single mutants. Thus, although Scd6/Edc3 appear to cooperate more extensively with Dhh1 than with Pat1 in translational control, the subset of 50-60 mRNAs exhibiting the strongest repression by Scd6/Edc3 or Pat1/Dhh1 require the concerted functions of Pat1, Dhh1, and either Scd6 or Edc3 to achieve their strong translational repression in WT cells.

We wondered whether the mRNAs translationally repressed by the decapping activators are also targeted for degradation by these factors. To assess this, we examined the mRNA changes in the four mutants for the groups of mRNAs defined above exhibiting TE derepression in the scd6Δ/edc3Δ or pat1Δ/dhh1Δ mutants (sectors (i) to (iii) of Fig. 5A). The group of mRNAs showing increased TEs only in the pat1Δ/dhh1Δ strains show little change in mRNA abundance in these mutants (Fig. S6A(iii)), indicating selective repression of translation vs degradation of these mRNAs. Similarly, the mRNAs translationally up-regulated only in the scd6Δ/edc3Δ mutants show little change in mRNA abundance in the scd6Δedc3Δ strain, while showing modest derepression in the pat1Δ/dhh1Δ mutants (Fig. S6A(i)). In contrast, the group of transcripts showing increased TEs in all of the mutants also show marked increases in mRNA abundance in all four mutants (Fig. S6A(ii)). Thus, coupled repression of translation and abundance by the decapping factors occurs only for the subset of transcripts exhibiting concerted translational repression by the four decapping activators. In contrast, the much larger groups of transcripts defined above showing increased mRNA abundance in the mutants (described in Fig. 2A) exhibit little change in median TE in all four mutants (Fig. S6B), suggesting that most mRNAs targeted for enhanced turnover do not exhibit translational repression of the fraction of undegraded transcripts still detectable in WT cells.

We next asked how well the mRNAs translationally repressed by Scd6/Edc3 are translated in WT cells. The group of mRNAs translationally repressed by Scd6/Edc3 acting in concert with Pat1/Dhh1 tend to be poorly translated in WT cells on rich medium, having a substantially lower median relative TE compared to the TE of all mRNAs (0.24 vs. 0.95), as assessed using ribosome profiling results for the WT strain (Fig. S6C, col. 2), thus resembling the mRNAs translationally repressed exclusively by Dhh1/Pat1 (col. 3). In contrast, the subset of transcripts translationally repressed exclusively by Edc3/Scd6 are generally well translated, exhibiting a ∼2.5-fold greater than average median TE in WT cells on rich medium (Fig. S6C, col. 1). Consistent with this distinction, we found by gene ontology (GO) analysis that products of the 136 mRNAs translationally repressed exclusively by Scd6/Edc3 are enriched for cytoplasmic or mitochondrial ribosomal proteins (P = 1 x 10-7), whereas products of the 76 transcripts repressed by Scd6/Edc3 and Pat1/Dhh1 acting in concert are enriched for factors involved in fermentation (P = 2 x 10-7), carbohydrate metabolism (P = 1 x 10-6) or metabolism of non-protein amino acids (P = 3 x 10-6). Thus, Scd6/Edc3 translationally repress distinct groups of mRNAs depending on whether Dhh1/Pat1 participate in the repression.

We showed previously (Zeidan et al. 2018) that mRNAs up-regulated by dhh1Δ and thus deemed to be targeted by Dhh1 for enhanced turnover are enriched for Dhh1 protein association in vivo as judged by RIP-Seq analysis of yeast mRNAs using antibodies against Dhh1 (Miller et al. 2018), thus providing evidence for a direct role of Dhh1 in stimulating decapping/degradation of these transcripts. The same was observed for the mRNAs up-regulated by pat1Δ, consistent with widespread cooperation between Dhh1 and Pat1 in repressing mRNA abundance (Vijjamarri et al. 2023a). Interestingly, here we observed greater than average Dhh1 occupancies for all four mRNA groups defined above (in Figs. 2A or 5A) that are up-regulated in transcript abundance or TE by the scd6Δ/edc3Δ or dhh1Δ/pat1Δ mutations (Fig. 5E). This finding is consistent with the fact that Dhh1 contributes to repressing the abundance or TE of the majority of transcripts in these groups (Figs. 2B & 5B). It is possible that Dhh1 is recruited to nearly all of these mRNAs in a complex with Dcp1:Dcp2 and other decapping activators (He et al. 2022), but makes differential contributions to their degradation or translational repression.

Scd6/Edc3 repress post-transcriptionally repress mRNAs encoding enzymes of respiration and catabolism of non-glucose carbon sources in rich medium

Recently, we reported that mRNAs up-regulated in abundance or TE in the dcp2Δ, pat1Δ and dhh1Δ mutants are enriched for mitochondrial proteins that function in Ox. Phos. (Vijjamarri et al. 2023a; Vijjamarri et al. 2023b). GO analysis led to the same finding here for the non-iESR mRNAs up-regulated in abundance by the scd6Δ/edc3Δ mutations, including many of the same categories of mitochondrial functions enriched among the mRNAs derepressed by the dhh1Δ/pat1Δ mutations (Fig. S7A-B, green entries). GO analysis of genes showing increased ribosome occupancies (RPFs), indicating either increased mRNA abundance or TE, confirms that the scd6Δ/edc3Δ mutations derepress the translation of Ox. Phos. gene transcripts (Fig. S7C-D). Examining a collection of Ox. Phos. genes functioning in electron transport, the TCA cycle or mitochondrial ATP synthase for changes in mRNA abundance, RPFs, and TEs reveals that the scd6Δedc3Δ and pat1Δdhh1Δ double mutations up-regulate expression of these genes primarily at the level of mRNA abundance, with small additional increases in TE (Fig. 6A). Western blot analysis revealed increased expression of four Ox. Phos. proteins (Qcr8, Atp20, Idh1, and Sdh4), two proteins (Cox14 and Cox20) involved in cytochrome c oxidase assembly, and mitochondrial cytochrome b2 (Cyb2) required for lactate utilization in scd6Δedc3Δ cells, relative to Gcd6 examined as loading control (Fig. 6B-C). Results similar to those in Fig. 6B-C were obtained for the same set of mitochondrial proteins in the dhh1Δ and pat1Δ mutants (Vijjamarri et al. 2023a) and for a subset of these proteins in dcp2Δ cells (Vijjamarri et al. 2023b). Importantly, we also observed increased expression of Cox2, one of three mitochondrially encoded subunits of mitochondrial cytochrome c oxidase, terminal enzyme of the mitochondrial ETC, whose expression correlates well with mitochondrial activity (see https://elifesciences.org/reviewed-preprints/90293). Except for dcp2Δ, all the decapping mutants tested had significantly increased amounts of Cox2 protein (Fig 6D), suggesting an increase in ETC activity in these mutants. Derepression of mRNA and RPF abundance in scd6Δedc3Δ and pat1Δdhh1Δ double mutants also occurred for enzymes of the glyoxylate cycle (Fig. 6E), which catalyze certain reactions of the TCA cycle in the cytoplasm to support gluconeogenesis during respiratory growth on two-carbon compounds; and increased expression of one such enzyme, Cit2, was confirmed by Western analysis (Figs. 6B-C). Both Ox. Phos. and the glyoxylate cycle normally operate at low levels in yeast growing with abundant glucose, as in our experiments, suggesting that Scd6/Edc3 cooperate with Dhh1/Pat1 to help suppress these pathways in glucose-replete cells.

Scd6/Edc3 post-transcriptionally repress proteins involved in respiration and suppress mitochondrial membrane potential in rich medium.

(A) Log2 changes in mRNA, RPFs, or TE conferred by the indicated double mutations vs. WT for 53 nuclear genes encoding mitochondrial proteins involved directly in oxidative phosphorylation. (B-C) Western blot analysis of 9 mitochondrial proteins and Gcd6 (examined as loading control) in WT and scd6Δedc3Δ strains, cultured in duplicate in YPD medium to OD600 of ∼0.6-0.8. WCEs were extracted under denaturing conditions and aliquots corresponding to 1X or 2X amounts of WCE were loaded in successive lanes for the two replicate cultures. Immune complexes were visualized with enhanced chemiluminescence (B). Signals for each protein were quantified, normalized to the corresponding signals for Gcd6 in the same extract and expressed relative to the resulting values for WT cells. Mean values and standard errors are plotted (C). (D) Western blot analysis of Cox2 in strains of the indicated genotypes in cells cultured as in (B). Cox2 signal intensity was normalized to total Coomassie-stained protein and the resulting relative Cox2 protein levels from three biological replicates were averaged and plotted. (E) Log2 changes in mRNA, RPFs, or TE conferred by the indicated double mutations vs. WT for 22 genes encoding enzymes of the glyoxylate cycle or fatty acid metabolism. (F) Expression of the CYC1-lacZ reporter on plasmid pLG265, lacking UAS1 and containing the optimized version of UAS2, UAS2UP1, in the WT strain grown on SC-Ura medium containing either 2% glucose or 3% glycerol/2% ethanol as carbon sources (i), or in WT and the indicated mutant strains on SC-Ura with 2% glucose (ii). β-galactosidase activity (nmoles of o-nitrophenyl-β-D-galactopyranoside (ONPG) cleaved per min per mg of total protein) was measured in whole cell extracts for 3 biological replicates of each strain and the mean values were normalized to the mean activity measured in WT grown with glucose as carbon source. **, P-value <0.01 from student’s t-test; ns, not significant. (G) Measurements of mitochondrial membrane potential. WT cells or transformants of the scd6Δedc3Δ mutant containing the EDC3 plasmid pLfz614-7 or empty vector were cultured in SC-Ura to mid-log phase. TMRM (500 nM) was added and incubated for 30 min before samples were collected and washed once with deionized water. ΔΨm was determined by measuring TMRM fluorescence intensity using flow cytometry. Data are presented in arbitrary fluorescence intensity units per OD600. 2-way ANOVA was used for statistical analysis and data are given as mean values ± SD (n=3) (****p<0.0001).

Consistent with post-transcriptional control of Ox. Phos. genes, the median relative RPF levels were up-regulated in the scd6Δedc3Δ mutant substantially more than the increases in relative Pol II occupancies observed at the cognate genes by ChIP-Seq analysis of Rpb1 (Fig. S8B, cols. 1-2). Moreover, the transcription factors responsible for induction of Ox. Phos. genes, the Hap2/Hap3/Hap4/Hap5 complex, were not activated in scd6Δedc3Δ cells: expression of a CYC1-lacZ reporter known to be activated by this Hap complex (Forsburg and Guarente 1989) was not elevated in the scd6Δ/edc3Δ mutants vs. WT in glucose-containing medium (Fig. 6F(ii)) but showed the expected induction in WT cells grown with glycerol/ethanol versus glucose (6F(i)) (Broach 2012). These findings support the notion that reduced decapping stabilizes Ox. Phos. gene transcripts in scd6Δedc3Δ cells.

In addition to Ox. Phos. genes, GO analysis revealed enrichment for genes involved in utilization of alternative carbon sources among those up-regulated at the level of mRNA or RPFs in the scd6Δ/edc3Δ mutants (Fig. S7A-D), as we observed recently for dhh1Δ,pat1Δ mutants (Vijjamarri et al. 2023a). Consistent with this, a group of 83 carbon catabolite repressed (CCR) genes, known to be glucose-repressed or activated by transcription factors Adr1 or Cat8 (Young et al. 2003; Tachibana et al. 2005), exhibit increased translation (RPFs) largely through increased mRNA abundance in both scd6Δedc3Δ and pat1Δdhh1Δ double mutants (Fig. S8A). This group includes enzymes for β-oxidation of fatty acids in addition to the glyoxylate cycle, which allow cells to synthesize precursors that can feed into gluconeogenesis or amino acid biosynthesis, or produce acetyl-CoA and generate NADH by respiration when growing on non-fermentable carbon sources (Young et al. 2003). The CCR genes exhibit larger increases in RPFs compared to Pol II occupancies at the cognate genes in scd6Δedc3Δ vs. WT cells (Fig. S8B, cols. 3-4), consistent with post-transcriptional repression of these genes by Scd6/Edc3. Supporting this, expression of an ADH2-lacZ reporter transcriptionally induced by activated Adr1 (Sloan et al. 1999) displayed the expected large induction in our WT strain cultured with glycerol/ethanol versus glucose as carbon source (Fig. S8C(i)), but its expression was actually reduced in the scd6Δedc3Δ mutant versus WT in glucose-grown cells (Fig. S8C(ii)).

Functional evidence that Scd6/Edc3 repress oxidative phosphorylation

We examined the effects of eliminating Scd6/Edc3 on mitochondrial electron transport by measuring mitochondrial membrane potential (ΔΨm) generated by the ETC using the probe tetramethylrhodamine (TMRM)—a cationic fluorescent dye that accumulates in mitochondria as a function of ΔΨm. Quantifying dye fluorescence by flow cytometry revealed increased TMRM fluorescence in the scd6Δedc3Δ mutant containing an empty vector compared to both the isogenic WT strain and the mutant complemented by WT EDC3 (Fig. 6G). These results are consistent with increased mitochondrial ETC activity in glucose-grown scd6Δedc3Δ cells.

To determine whether derepression of Ox. Phos. and other glucose-repressed genes in the decapping activator mutants alters cellular metabolites, we used targeted, quantitative LC-MS/MS based approaches (see Methods) to quantify levels of 147 polar metabolites in the isogenic scd6Δedc3Δ, dhh1Δ, pat1Δ, dcp2Δ and WT strains described above, cultured in YPD medium. Principle component analysis revealed clustering of results from biological replicates in the manner expected for reproducible differences in metabolite levels among different strains, with results for the scd6Δedc3Δ mutant most closely resemble those for dhh1Δ, which in turn were more similar to the results for pat1Δ vs. the dcp2Δ mutant or WT strains (Fig. 7A). This conclusion was borne out by cluster and correlation analyses of the changes in metabolites between each mutant compared to WT, with the strongest correlation observed for scd6Δedc3Δ vs. dhh1Δ, followed by dhh1Δ vs. pat1Δ (Fig. 7B). Considering the subset of 46 metabolites up-regulated in any two of the four mutants again showed greatest similarity between changes conferred by scd6Δedc3Δ vs. dhh1Δ followed by dhh1Δ vs. pat1Δ (Fig. 7C). These findings mirror the results from RNA-seq and ribosome profiling in which the up-regulation of mRNA levels or translation was most similar between the scd6Δedc3Δ and dhh1Δ mutants. Pathway analysis of the 46 up-regulated metabolites (at https://www.metaboanalyst.ca/MetaboAnalyst/Secure/pathway) revealed a significant enrichment for metabolites of both the TCA and glyoxylate cycles (Fig. 7D), with five of six TCA cycle intermediates detected (fumarate, malate, α-ketoglutarate, cis-aconitate and citrate) being elevated in the decapping mutants (Fig. 7E). These results are consistent with the possibility that up-regulation of Ox. Phos. proteins (Fig. 6A-D) and ETC function (Fig. 6F) in the decapping mutants leads to increased flux from glucose towards the TCA cycle.

To determine unambiguously if there was indeed increased flux through the TCA cycle coming from glucose breakdown, we utilized a pulse label of 13C6-glucose and then traced the 13C label incorporation from glucose breakdown into intermediates of the TCA cycle. Specifically, WT and mutant cells were grown in high glucose, pulsed with 13C6 glucose, and relative label incorporation into TCA cycle intermediates was estimated for each carbon molecule derived from glucose (as indicated schematically in Fig. 8A) eight min following the pulse. In all mutants, 13C label incorporation in all TCA cycle intermediates was significantly higher than in WT cells (Fig. 8A), demonstrating that carbon flux from glucose into the TCA cycle was elevated in the decapping mutants. These results, along with increased ETC activity (Fig. 6G), indicate that mitochondria are functionally derepressed in glucose in these mutants. If so, we might expect the mutants to show increased ATP production from respiration versus glycolysis. Measuring total ATP levels in glucose-grown cells revealed that, although all of the decapping mutants except pat1Δ showed some reduction in total levels of ATP per cell (Fig. 8B), the proportion of ATP produced from respiration, and thus eliminated by sodium azide treatment (by inhibiting the ETC), is elevated in all of the decapping mutants (Fig. 8C).

Eliminating decapping activators or decapping enzyme confers similar changes in polar metabolites.

(A) Principal component analysis of the levels of 147 metabolites in biological replicates of each strain. (B-C) Hierarchical clustering analysis of log2 changes in all 147 metabolites analyzed (B) or the 46 metabolites up-regulated in any two of the four mutants (C) conferred by the indicated mutations vs. WT, including the Spearman coefficients (ρ) and P values for the indicated correlations. (D) Results of pathway analysis of the 46 up-regulated metabolites described in (C) conducted at https://www.metaboanalyst.ca/MetaboAnalyst/Secure/pathway. Red ovals depict groups of metabolites significantly enriched among the set of 46 compounds, with P-value <0.05. (E-F) Log2 changes in levels of TCA cycle intermediates (E) or amino acids (F) conferred by the indicated mutations vs. WT.

Elimination of decapping activators or decapping enzyme up-regulates respiration, increasing flux from glucose into TCA cycle intermediates and proportion of ATP produced by Ox. Phos.

(A) Three biological replicates of cells of each indicated genotype were cultured in YP with 2% unlabeled glucose, shifted to YP with 1% unlabeled glucose for 20 min, and pulsed with 13C6 labelled glucose (at final concentration of 1%) for 8 min, followed by extraction of metabolites and quantification of the indicated TCA cycle intermediates by mass spectrometry. In the diagrams, green circles signify the labelled carbon atom and the notation M+1, M+2, etc. indicate the mass increase in the molecules due to the labelled carbon. The depicted labelling pattern of metabolites reflects on cycle of the TCA cycle, resulting in mass additions of M+1 and M+2; however, across multiple cycles, a broader range of metabolite species with different mass additions will emerge. The metabolite signal intensities in all samples are expressed relative to that determined for the first replicate of the WT strain. (B-C) Measurements of ATP levels and proportions of total ATP impaired by azide inhibition of ETC activity. Cells cultured in YPD were treated or untreated with sodium azide for 30 min prior to harvesting. ATP levels were determined in extracts and normalized to OD600 units of cells for three biological replicates each of treated and untreated cell aliquots. Mean values for each strain are plotted in (B) and relative fractions of ATP in untreated samples retained following azide treatment (ATP_untreated)-(ATP_Azide)/ATP_untreated), normalized to the values determined for WT are plotted in (C) with results from a student’s t-test indicated with asterisks: **, P<0.005; *, P<0.05.

Interestingly, amino acids are also up-regulated in the decapping mutants (Fig. 7D), particularly in pat1Δ cells (Fig. 7F), as well as intermediates in amino acid biosynthesis and amino acid derivatives (Fig. 7D). The expression of amino acid biosynthetic enzymes is unaffected or reduced in the mutants however (Fig. S8D), suggesting that increased amino acid abundance results from metabolic reprogramming rather than increased biosynthetic capacity. One possibility is that accumulation of the TCA cycle intermediate α-ketoglutarate leads to increased production of glutamate and glutamine, which are precursors in all amino acid biosynthetic pathways (Ljungdahl and Daignan-Fornier 2012). The two 3-phosphotrioses generated in glycolysis (glyceraldehyde 3-phosphate and dihydroxyacetone phosphate) are also elevated in the mutants, possibly owing to increased glyoxylate shunt function, which can lead to precursors of these intermediates and stimulate synthesis of amino acids serine and glycine (Ljungdahl and Daignan-Fornier 2012) (Fig. 7D, Methane metabolism category. The increases in fructose 1,6-bisphosphate, fructose 6-phosphate, glucose 6-phosphate, and UDP-glucose in all four mutants (Fig. 7D, Starch/Sucrose metabolism category) all suggest substantial metabolic rewiring indicative of glucose derepression and usage of alternative carbon sources. In addition to amino acids, pyrimidine nucleotides are up-regulated in the mutants (Fig. 7D, Pyrimidine metab.), which might be driven by the elevated glutamine levels or increased glycolytic flux towards the pentose phosphate pathway (Ljungdahl and Daignan-Fornier 2012).

Discussion

RNA-seq analysis of single and double scd6Δ and edc3Δ mutants has revealed that Scd6 and Edc3 have largely overlapping functions in repressing mRNA abundance, as the presence of either protein alone is sufficient for nearly WT levels of most mRNAs found derepressed in the double mutant (Fig. S3A & Fig. 1B-C(ii)-(iii)). The mRNAs dysregulated in the scd6Δedc3Δ mutant are significantly enriched for ESR mRNAs; however, the majority are not ESR transcripts and are most likely repressed more directly by Scd6/Edc3 (Fig. 1A-B). Scd6/Edc3 redundancy in regulating mRNA abundance is consistent with their similarities in sequence and domain structure, including shared FDF motifs that interact competitively with Dhh1 (Tritschler et al. 2008; Tritschler et al. 2009) and N-terminal LSm domains that compete for binding motifs in the Dcp2 CTT (Fromm et al. 2012), as well as the synthetic growth defect produced by deleting both genes simultaneously (Decourty et al. 2008) (Fig. S1A). Interestingly, a set of 37 mRNAs appears to be repressed exclusively by Edc3 with little contribution from Scd6, or with even an opposing stimulatory effect of Scd6, on their abundance (Fig. 1B-C(i)). This last group includes the canonical Edc3 targets identified previously, YRA1 and RPS28B (He et al. 2022) and, interestingly, is enriched for genes belonging to the GO category “mitochondrion” (22/37 genes) and the Ox. Phos. categories of electron transport and mitochondrial ATP synthesis (11/37 genes: ATP3 ATP16 ATP5 TIM11 ATP17 COX4 QCR9 CYC1 COX12 ATP18 COX5A). The fact that deleting SCD6 in the double mutant frequently diminishes the up-regulation of these transcripts conferred by deleting EDC3 alone (Fig. 1C(i)) might indicate that eliminating both Scd6 and Edc3 together enables a distinct degradation pathway that compensates for loss of Edc3-stimulated decapping/decay.

Several lines of evidence support the conclusion that most of the changes in mRNA abundance in the scd6Δedc3Δ double mutant result from impaired decapping and attendant 5’-3’ degradation by Xrn1. ChIP-seq analysis of Rpb1 shows that Pol II occupancies in the coding sequences increase by a much smaller amount compared to the increased transcript levels for non-iESR mRNAs derepressed in scd6Δ/edc3Δ cells (Fig. 3E), indicating a minor contribution of increased transcription to their up-regulation. Consistent with mRNA turnover via decapping, most of these transcripts are up-regulated similarly by scd6Δedc3Δ and deletion of DCP2 (Fig. 3A-B), they exhibit a heightened proportion of decapped isoforms in WT cells that is eliminated by deleting DHH1 (Fig. 3C), and they show evidence of co-translational 5’-3’ decay of decapped intermediates by Xrn1 (Fig. 3D); none of which was observed for the mRNAs down-regulated in scd6Δ/edc3Δ cells. Similar findings were reported previously for sets of mRNAs up-regulated by dhh1Δ or pat1Δ (Vijjamarri et al. 2023a), consistent with the widespread involvement of Dhh1 and Pat1 in controlling the levels of mRNAs preferentially targeted by Scd6/Edc3.

Supporting this last assertion, most (∼70%) of the mRNAs repressed in abundance by Edc3/Scd6 are also repressed by Dhh1 and Pat1, and their efficient repression in WT cells involves independent contributions of similar magnitude by Pat1, Dhh1 and either Edc3 or Scd6 (Fig. 2A-B(ii)). Another large set of mRNAs is repressed predominantly by Pat1 with lesser comparable contributions by Dhh1 and Scd6/Edc3 (Fig. 2A-B(iii)). Our finding of extensive cooperation among these four decapping factors in repressing a common set of mRNAs is consistent with our recent finding that ∼55% of the mRNAs derepressed in the dcp2Δ mutant, and thus targeted by Dcp2 for enhanced degradation, tend to be derepressed in the pat1Δdhh1Δ and scd6Δedc3Δ mutants, whereas the remaining 45% are generally derepressed by upf1Δ instead. This finding suggested a major bifurcation of Dcp2-activation by either Pat1/Dhh1/Scd6/Edc3 or the Upf factors responsible for NMD (Vijjamarri et al. 2023b).

Closer examination of the effects of individual mutations on mRNA levels revealed a greater overlap between the mRNAs dysregulated by dhh1Δ and scd6Δ/edc3Δ versus those altered by pat1Δ (Fig. 2C-E), consistent with the aforementioned FDF motifs in both Scd6 and Edc3 that interact with Dhh1, and with evidence for distinct complexes of the decapping enzyme Dcp1:Dcp2 containing Dhh1 and Edc3 or Scd6 but lacking Pat1. It also supports the recent suggestion that Edc3 and Scd6 act interchangeably to recruit Dhh1 to the Edc3-interaction site in the Dcp2 CTT to stimulate turnover of several Dhh1 target mRNAs (He et al. 2022). Our RNA-seq results support this last proposal in revealing that all four Dhh1 target mRNAs examined there (EDC1, SDS23, HXT6, and HSP12) are derepressed in the scd6Δedc3Δ double mutant while showing little derepression in either single mutant (Fig. S9), as expected for redundant targeting of Dhh1 to the Dcp2-CTT by Scd6/Edc3. Finding here that this behavior applies to most mRNAs up-regulated in the dhh1Δ mutant (Fig. 2E), we propose that redundant Scd6/Edc3 targeting is the predominant mechanism of Dhh1-enhanced mRNA degradation in the yeast transcriptome (Fig. S10).

Despite strong evidence for the involvement of decapping in mRNA turnover directed by either Scd6/Edc3 or Pat1/Dhh1, it appears that neither enrichment for non-optimal codons nor low translational efficiencies are principal determinants of preferential targeting of mRNAs by these proteins (Fig. S5A-B). There are probably exceptions to these generalizations among the hundreds of transcripts repressed by these factors; nor can we rule out the involvement of clusters of non-optimal codons in transcripts of average overall codon optimality that generate a queue of slowly elongating ribosomes that triggers decapping. Nevertheless, it seems likely that other sequences or properties of the transcripts most strongly repressed by Scd6/Edc3 or Dhh1/Pat1 frequently underlie their preferential targeting by these factors for decapping/decay. Sequence-specific RNA-binding proteins that actively recruit these decapping activators is an attractive potential mechanism (Fig. S10). For example, there is evidence that Dhh1 is recruited to mRNAs in association with the Ccr4-NOT deadenylase via RNA-binding protein Puf5 (Goldstrohm et al. 2006).

Ribosome profiling of the single and double scd6Δ and edc3Δ mutants provided evidence that Scd6 and Edc3 also have highly overlapping functions in repressing the translation of ∼200 mRNAs, whose TE values are up-regulated substantially only in the double mutant (Fig. 4A-B(iii)). The observation that changes in TE (RPF density) generally correlate with changes in protein abundance conferred by scd6Δedc3Δ (Fig. 4E) implies that Scd6/Edc3 generally repress translation at the stage of initiation rather than elongation. Many mRNAs translationally repressed by Scd6/Edc3 additionally require Dhh1, and the subset most strongly repressed also requires Pat1 for efficient translational repression (Fig. 5B(ii) and 5D). This last group of mRNAs is translated at very low levels in WT cells on rich medium (Fig. S6C), which might reflect the concerted action of all four decapping activators in suppressing their translation. The transcript abundance of these mRNAs is also derepressed in each of the mutants (Fig. S6A(ii)), indicating enhanced degradation coupled with translational repression by the four decapping activators. One possibility is that a complex of Dcp1:Dcp2 containing Scd6/Edc3, Pat1/Dhh1, or different combinations of these factors (He et al. 2022), is recruited to these mRNAs and impedes recruitment of PICs to diminish translation in parallel with stimulating decapping and 5’-3’ degradation by Xrn1 (Fig. S10(i) & (ii)-(iv)).

The additional sets of mRNAs identified here that are translationally repressed exclusively by Scd6/Edc3 or Pat1/Dhh1 (Fig. 5A(i) & (iii)) show much less evidence of transcript degradation promoted by these factors (Fig. S6A(i) & (iii)), suggesting that inhibition of PIC recruitment for translation initiation occurs without decapping, or that it involves decapping uncoupled from degradation by Xrn1 (Fig. S10 (ii)-(iii)). In the latter scenario, the decapped mRNAs would be unable to recruit the cap-binding initiation factors necessary for PIC recruitment and hence persist in WT cells as translationally-inert degradation intermediates. Eliminating the decapping activators would then diminish decapping and eliminate these fractions of poorly translated transcript isoforms to confer the increased TE values we observed for the cognate genes in the decapping mutants. This hypothetical mechanism of translational repression via decapping is consistent with the lower than average capped to total mRNA (C/T) ratios observed in WT cells for the mRNAs translationally repressed by all four decapping activators or by Pat1/Dhh1 (Fig. S6D (ii)-(iii)), indicating their high rates of decapping and accumulation of decapped isoforms in WT cells. Previously, we suggested that such translational repression via decapping could account for a subset of poorly translated mRNAs exhibiting TE increases in both dcp2Δ and dhh1Δ mutants and showing Dhh1-dependent lower than average C/T ratios in WT (Vijjamarri et al. 2023b). In contrast, the group of mRNAs identified here that are translationally repressed exclusively by Scd6/Edc3 display higher than average C/T ratios in WT (Fig. S6D(i)), which is more compatible with the inhibition of PIC recruitment independently of decapping (Fig. S10(iii)). Translational repression without decapping or decay could also occur if transcripts bound by the decapping complexes are sequestered from the translation initiation machinery in RNA granules in a state where mRNA decapping or degradation by Xrn1 occurs inefficiently (Fig S9(iv)). Finally, it is important to note that mRNA decay occurs apparently without translational control for the much larger groups of ∼600 and ∼1000 transcripts showing derepressed mRNA abundance but little evidence of TE changes in the scd6Δedc3Δ and pat1Δdhh1Δ mutants, respectively (Fig. S6B). It is unclear whether these mRNAs escape translational repression and are targeted exclusively for decapping and degradation (Fig. S10(i)), or if they are simply degraded too fast to allow translational repression to be detected by ribosomal profiling of steady-state mRNAs.

Recently, we reported that Pat1 and Dhh1 functionally collaborate with the decapping enzyme to repress genes and cellular pathways that are normally repressed in rich medium (Vijjamarri et al. 2023a; Vijjamarri et al. 2023b). Here we found evidence that Scd6 and Edc3 broadly cooperate with Pat1/Dhh1 in this post-transcriptional control, acting redundantly to down-regulate the abundance, translation, or both, of mRNAs encoding numerous proteins involved in respiration or utilization of alternative carbon sources (Figs. 6A-C, S6C, S7A-B), without affecting transcription of the cognate genes Fig. 6F, S7A-C). This derepression includes multiple nucleus-encoded mitochondrial proteins that function in Ox. Phos. (Fig. S7C-D & Fig. 6A-B), including mitochondrially encoded subunit II of ETC component cytochrome c oxidase (Fig. 6D). Evidence that derepression of Ox. Phos. mRNAs/proteins in the scd6Δedc3Δ mutant is sufficient to elevate respiration in glucose-replete cells came from our finding of elevated mitochondrial membrane potential in the scd6Δedc3Δ mutant, extending similar findings we made previously for dcp2Δ, dhh1Δ and pat1Δ mutants (Vijjamarri et al. 2023a; Vijjamarri et al. 2023b). Consistent with this, our metabolomics analysis here revealed elevated steady-state levels of TCA cycle intermediates in all four decapping mutants (Fig. 7D), and metabolic flux experiments on cells pulse-labeled with 13C6-glucose demonstrated increased carbon incorporation from glucose breakdown into TCA cycle intermediates in the same mutant strains (Fig. 8A). Analysis of ATP levels further revealed increased proportions of ATP derived from Ox. Phos. in the decapping mutants (Fig. 8B). Together, these findings provide strong evidence that mitochondrial respiration is inappropriately elevated in glucose-replete cells of the decapping activator mutants. It is well established that the enzymes involved in respiration and catabolism of non-glucose carbon sources are repressed at the transcriptional level in glucose-replete yeast (Zaman et al. 2008). Our findings demonstrate that Scd6/Edc3 collaborate with Dhh1/Pat1 to provide an additional layer of post-transcriptional control of these enzymes through enhanced decapping/decay or translational repression to help ensure their complete suppression in glucose-replete cells.

eLife Assessment

This useful study highlights the largely redundant role of the decapping activators Edc3 and Scd6 in orchestrating post-transcriptional programs to modulate metabolic responses to nutrients in yeast. The authors provide solid evidence for their conclusions, employing a variety of mutants in conjunction with a battery of transcriptome-wide analyses. This study could be further strengthened by direct biochemical validation of the functional interactions observed by systems biology approaches.

Reviewer #1 (Public review):

Summary:

mRNA decapping and decay factors play critical roles in post-transcriptionally regulating gene expression. Here, Kumar and colleagues investigate how deleting two yeast decapping enhancer proteins (Edc3 and Scd6), either alone or in tandem, influences the transcriptome. Using RNA-Seq and ribosome profiling, they come to the conclusion that these factors generally act in a redundant fashion, with a mutant lacking both proteins showing an increased abundance of select mRNAs. As these upregulated transcripts are also upregulated in mutants lacking the decapping enzyme, Dcp2, and show no increases in transcription of their cognate genes, they come to the conclusion that this is at the level of mRNA decapping and decay. Their ribosome profiling data also led them to conclude that Scd6 and Edc3 display functional redundancy and cooperativity with Dhh1/Pat1 in repressing the translation of specific transcripts. Finally, as their data suggest that Scd6 and Edc3 repress mRNAs coding for proteins involved in cellular respiration, as well as proteins involved in the catabolism of alternative carbon sources, they go on to show that these decapping activators play a role in repressing oxidative phosphorylation.

Strengths:

Overall, this manuscript is well-written and contains a large amount of high-quality data and analyses. At its core, it helps to shed light on the overlapping roles of Edc3 and Scd6 in sculpting the yeast transcriptome.

Weaknesses:

(1) While the data presented makes conclusions about mRNA stability based on corresponding ChIP-Seq analyses and analyzing other mutants (e.g. Dcp2 knockout), at no point is mRNA stability actually ever directly assessed. This direct assessment, even for select transcripts, would further strengthen their conclusions.

(2) Scd6 and Edc3 show a high level of functional redundancy, as demonstrated by the double mutant. As these proteins form complexes with other decapping factors/activators, I'm curious if depleting both proteins in the double mutant destabilizes any of these other factors. Have the authors ever assessed the levels of other key decapping factors in the double mutants (i.e. Dhh1, Pat1, Dcp2...etc)? I wonder if depleting both proteins leads to a general destabilization of key complexes. It would also be interesting to see if depleting Edc3 or Scd6 leads to a concomitant increase in the other protein as a compensatory mechanism.

(3) While not essential, it would be interesting if the authors carried out add-back experiments to determine which domain within Scd6/Edce3 plays a critical role in enforcing the regulation that they see. Their double mutant now puts them in a perfect position to carry out such experiments.

Reviewer #2 (Public review):

Summary:

This manuscript by Kumar and Zhang presents compelling evidence that Edc3 and Scd6 decapping activators present a high degree of redundancy that can only be overcome by a double mutant. In addition, the authors provide strong evidence of these complexes in regulating starvation-induced pathways as evidenced by measurements of mitochondrial membrane potential, metabolomics, and analysis of the flux of Krebs cycle intermediates.

Strengths:

Kumar and Zhang et al provide multiple sources of evidence of the direct mechanism of Edc3 and Scd6 function, by using and comparing different approaches such as mRNA-seq, ribosome occupancies, and translational efficiencies. By extensive analysis, the authors show that this complex can also regulate genes outside the Environmental Stress Response (non-iESR), which are significantly up-regulated in all three mutants. Remarkably, the gene ontology analysis of these non-iESR genes identifies enrichment for mitochondrial proteins that are implicated in the Krebs cycle. Overall, this study adds novel mechanistic insight into how nutrients control gene expression by modulating decapping and translational repression.

Weaknesses:

The authors show very nicely in Figure S1A that growth phenotypes from scd6Δedc3∆ can be rescued by transformation of EDC3 (pLfz614-7) or SCD6 (pLfz615-5). The manuscript might benefit from using these rescue strategies in the analysis performed (e.g. RNA-seq, ribosome occupancies, and translational efficiencies). Also, these rescue assays could provide a good platform to further characterise the protein-protein interactions between Edc3, Scd6, and Dhh1.

Reviewer #3 (Public review):

Summary:

In this paper, Kumar et al aimed to investigate the roles of two decapping activators, Edc3 and Scd6, in regulating mRNA decay and translation in yeast. Previous research suggested limited individual roles for these proteins in mRNA decay. The authors hypothesized that Edc3 and Scd6 act redundantly and explored how these proteins interact with two other factors involved in mRNA decapping and translational repression, with Dhh1 and Pat1, particularly in response to nutrient availability. The study aims to identify mRNAs targeted by these proteins for degradation and translation repression and assess their impact on metabolic pathways including mitochondrial function and alternative carbon source utilization.

Strengths:

The paper has several strengths including the comprehensive approach taken by the authors using multiple experimental techniques (RNA-seq, ribosome profiling, Western blotting, TMT-MS, and polysome profiling) to examine both mRNA abundance and translation efficiency, thereby providing multiple lines of evidence to support their conclusions. The authors demonstrate clear redundancy of the factors by using single and double mutants for Edc3 and Scd6 and their global approach enables an understanding of these factors' roles across the yeast transcriptome. The work connects post-transcriptional processes to nutrient-dependent gene regulation, providing insights into how cells adapt to changes in their environment. The authors demonstrate the redundant roles of Edc3 and Scd6 in mRNA decapping and translation repression. Their RNA-seq and ribosome profiling results convincingly show that many mRNAs are derepressed only in the double mutants, confirming their hypothesis of redundancy. Furthermore, the functional cooperation between Edc3/Scd6 and Dhh1/Pat1 in regulating specific metabolic pathways, like mitochondrial function and carbon source utilization, is supported by the data. The results therefore support the authors' conclusions that these decapping factors work together to fine-tune gene expression in response to nutrient availability.

Weaknesses:

The limitations of the study include the use of indirect evidence to support claims that Edc3 and Scd6 recruit Dhh1 to the Dcp2 complex, which is inferred from correlations in mRNA abundance and ribosome profiling data rather than direct biochemical evidence. Also, there is limited exploration of other signals as the study is focused on glucose availability, and it is unclear whether the findings would apply broadly across different environmental stresses or metabolic pathways.

Nonetheless, the study provides new insights into how mRNA decapping and degradation are tightly linked to metabolic regulation and nutrient responses in yeast. The RNA-seq and ribosome profiling datasets are valuable resources for the scientific community, providing quantitative information on the role of decapping activators in mRNA stability and translation control.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

Strengths:

Overall, this manuscript is well-written and contains a large amount of high-quality data and analyses. At its core, it helps to shed light on the overlapping roles of Edc3 and Scd6 in sculpting the yeast transcriptome.

Weaknesses:

(1) While the data presented makes conclusions about mRNA stability based on corresponding ChIP-Seq analyses and analyzing other mutants (e.g. Dcp2 knockout), at no point is mRNA stability actually ever directly assessed. This direct assessment, even for select transcripts, would further strengthen their conclusions.

We appreciate the reviewer’s concern but wish to emphasize that we conducted ChIP-Seq analysis of RNA Polymerase II occupancies in the CDSs of all genes, known to be a reliable indicator of transcription rate, and found only small increases in Pol II occupancies that cannot account for the increased transcript levels of the cohort of mRNAs up-regulated in the scd∆6edc3∆ double mutant (Fig. 3E). This provides strong evidence that increased transcription is not the main driver of increased mRNA abundance in this mutant. Bolstering this conclusion, we showed that the Hap2/Hap3/Hap4/Hap5 complex of transcription factors responsible for induction of Ox. Phos. genes was not activated in scd6Δedc3Δ cells in glucose medium (Fig. 6F(ii)); nor was the Adr1 activator of CCR genes activated (Fig. S9C(i)), ruling out transcriptional induction of their target genes in glucose-replete scd6Δ/edc3Δ cells and instead favoring reduced degradation as the mechanism underlying derepression of Ox. Phos. and CCR gene transcripts in this mutant. In Fig. 3B, we further showed that the majority of mRNAs up-regulated in the scd6Δedc3Δ double mutant are also derepressed by dcp2Δ, and in Fig. 3D that the mRNAs up-regulated in scd∆6edc3∆ cells exhibit a higher than average codon protection index (CPI) indicating a heightened involvement of decapping and co-translational degradation by Xrn1 in their decay. To provide additional support for our conclusion, we have conducted new experiments to measure the abundance of capped mRNAs genome-wide by CAGE sequencing of total mRNA in both WT and scd∆6edc3∆ cells. As established previously, normalizing CAGE TPMs to total mRNA TPMs determined by RNA-Seq, dubbed the C/T ratio, provides a reliable measure of the capped proportion of each transcript. The new data presented in Fig. 3C indicate that the mRNAs up-regulated in the scd∆6edc3∆ mutant have significantly lower than average C/T ratios in WT cells, whereas the C/T ratios for the down-regulated transcripts are higher than average, and that these differences between the two groups and all expressed mRNAs are diminished in the scd∆6edc3∆ double mutant. These are the results expected if the up-regulated mRNAs are selectively targeted for decapping in WT cells dependent on Edc3/Scd6, whereas the downregulated mRNAs are targeted by Edc3/Scd6 less than the average transcript. In the original version of the paper, we came to the same conclusion by analyzing our previous CAGE data for the dhh1∆ mutant for the same transcripts dysregulated scd∆6edc3∆ cells, now presented as supportive data in Fig. S3F. Finally, we added the fact that among all four Dhh1 target mRNAs examined in the previous study of He et al. (2022) and found here to be up-regulated selectively in the scd6∆edc3∆ double mutant (Fig. S10), two of them (SDS23 and HXT6) were shown directly to have longer half-lives in dhh1∆ vs. WT cells by He et al. (2018). Hence, the combined evidence is compelling that selective up-regulation of particular mRNAs in the scd∆6edc3∆ mutant results from diminished decapping/decay rather than enhanced transcription; and we feel that the additional supporting evidence that would be provided by measuring half-lives of a small group of up-regulated transcripts would not justify the considerable effort required to do so. Moreover, the standard approach for such experiments of impairing transcription with an inhibitor of Pol II or a Pol II Ts- mutation has been criticized because of the known buffering (suppression) of mRNA decay rates in response to impaired transcription.

(2) Scd6 and Edc3 show a high level of functional redundancy, as demonstrated by the double mutant. As these proteins form complexes with other decapping factors/activators, I'm curious if depleting both proteins in the double mutant destabilizes any of these other factors. Have the authors ever assessed the levels of other key decapping factors in the double mutants (i.e. Dhh1, Pat1, Dcp2...etc)? I wonder if depleting both proteins leads to a general destabilization of key complexes. It would also be interesting to see if depleting Edc3 or Scd6 leads to a concomitant increase in the other protein as a compensatory mechanism.

We thank the reviewer for this insight. Examining our Ribo-Seq and TMT-MS data revealed that Dhh1 expression and steady-state abundance are increased ~2-fold in the scd6∆edc3∆ strain, indicating that the up-regulation of many of the same mRNAs by scd6∆edc3∆ and dhh1∆ does not result indirectly from reduced levels of Dhh1 in the scd6∆edc3∆ mutant. The predicted increased in Dhh1 expression might signify a compensatory response to the absence of Scd6/Edc3. We also observed an ~40% reduction in Dcp2 translation (RPFs) and mRNA abundance in the scd6∆edc3∆ strain, which might contribute to the up-regulation of mRNAs dysregulated in this mutant. However, our new immunoblot analyses revealed no significant reduction in steady-state Dcp2 levels in scd6∆edc3∆ cells (Input lanes in Figs. 3F and S4C(i)-(ii)). Moreover, our previous finding that the majority of mRNAs subject to NMD, up-regulated by both upf1∆ and dcp2∆, are not upregulated by scd6∆edc3∆ implies that Dcp2 abundance in scd6∆edc3∆ cells is adequate for normal levels of NMD and favors a direct role for Scd6/Edc3 in accelerating degradation of most transcripts up-regulated in this mutant. We have added these points to the DISCUSSION.

(3) While not essential, it would be interesting if the authors carried out add-back experiments to determine which domain within Scd6/Edce3 plays a critical role in enforcing the regulation that they see. Their double mutant now puts them in a perfect position to carry out such experiments.

We agree with the reviewer that our scd6∆edc3∆ strain provides an opportunity to dissect the Scd6 and Edc3 proteins to determine which domains and motifs of each protein are most critically required for their functions in activating mRNA decay. However, if conducted thoroughly, this would entail an extensive analysis requiring a combination of genetics, biochemistry and genomics. Considering the large amount of data already presented in 43 and 34 panels of main and supplementary figures, respectively, we feel that these additional experiments would be conducted more appropriately as a stand-alone follow-up study.

Reviewer #2 (Public review):

Weaknesses:

The authors show very nicely in Figure S1A that growth phenotypes from scd6Δedc3∆ can be rescued by transformation of EDC3 (pLfz614-7) or SCD6 (pLfz615-5). The manuscript might benefit from using these rescue strategies in the analysis performed (e.g. RNA-seq, ribosome occupancies, and translational efficiencies). Also, these rescue assays could provide a good platform to further characterise the protein-protein interactions between Edc3, Scd6, and Dhh1.

We responded to this point immediately above in responding to Rev. #1.

Reviewer #3 (Public review):

Weaknesses:

The limitations of the study include the use of indirect evidence to support claims that Edc3 and Scd6 recruit Dhh1 to the Dcp2 complex, which is inferred from correlations in mRNA abundance and ribosome profiling data rather than direct biochemical evidence.

While the reviewer makes a valid point, it is important to note that the greater correlations between effects of scd6∆edc3∆ with those conferred by dhh1∆ vs. pat1∆ also extended to changes in metabolites (Fig. 7A-C). To provide more direct evidence that Edc3 and Scd6 recruit Dhh1 to the Dcp2 complex, we have now conducted co-immunoprecipitation experiments (presented in new Figs. 3F and S5) demonstrating that association of Dhh1 with Dcp2 is diminished in the scd6∆edc3∆ double mutant but not in either scd6∆ or edc3∆ single mutant, thus providing biochemical support for our proposal.

Also, there is limited exploration of other signals as the study is focused on glucose availability, and it is unclear whether the findings would apply broadly across different environmental stresses or metabolic pathways. Nonetheless, the study provides new insights into how mRNA decapping and degradation are tightly linked to metabolic regulation and nutrient responses in yeast. The RNA-seq and ribosome profiling datasets are valuable resources for the scientific community, providing quantitative information on the role of decapping activators in mRNA stability and translation control.

While not disputing the facts of this comment, we think it is unjustified to label as a weakness that our study focused on glucose-grown cells considering the large amount of new data and insights made possible by our multi-omics approach, presented in >70 separate figure panels and nine supplementary datafiles, which the reviewer has characterized as being valuable to the scientific community. Parallel studies in non-preferred carbon or nitrogen sources are underway and represent large-scale investigations in their own right, for which the current dataset in glucose-replete cells provides the critical reference condition.

Reviewer #1 (Recommendations for the authors):

The authors made a note that a set of 37 mRNAs is repressed exclusively by Edc3 with little contribution by Scd6, a list that includes the RPS28B mRNA. Edc3 has been previously reported to promote the decay of this mRNA in a deadenylation-independent fashion by binding to an element in its 3'UTR (PMIDs 15225544, 24492965). Can the authors comment on whether Edc3 may be binding to similar elements in the 3'UTRs of these transcripts in their shortlist? This could be an interesting topic matter for discussion as well.

While an interesting idea, this seems unlikely because the 3’UTR sequence in RPS28B mRNA was shown to bind Rps28 protein itself to confer heightened decapping and decay dependent on Edc3 in a negative autoregulatory loop that exerts tight control over Rps28 protein levels. It would be surprising if Edc3mediated repression of the other 36 mRNAs would involve Rps28 as none of them encode cytoplasmic ribosomal proteins. Nevertheless, we searched for a conserved motif among the 3’UTRs of the 37 mRNAs using the MEME suite and found enrichment for motifs identified for RNA binding proteins Hrp1 and Nab2 and two novel motifs, but none of these motifs could be recognized within in the Rps28 autoregulatory loop. We have chosen not to comment on these findings in the revised manuscript to avoid lengthening it unnecessarily with inconclusive observations.

Reviewer #2 (Recommendations for the authors):

The authors show very nicely in Figure S1A that growth phenotypes from scd6Δedc3∆ can be rescued by the transformation of EDC3 (pLfz614-7) or SCD6 (pLfz615-5). The manuscript might benefit from using these rescue strategies on the analysis performed (e.g. RNA-seq, ribosome occupancies, and translational efficiencies); or expressing truncated mutants of EDC3 (pLfz614-7) or SCD6 (pLfz615-5), to show that they can act as dominant negative competitors, either on the binding to Dhh1 and Dcp2.

We addressed this comment above in our response to this Reviewer.

Reviewer #3 (Recommendations for the authors):

(1) Labels such as "mRNA_up_s6,e3" are not defined in figures or the text. I suggest clearer sample labeling throughout.

The labels had been defined at first mention in the RESULTS but are now indicated there more explicitly, as well as in the legend to Fig. 1.

(2) In Figure 1D it is surprising that the mRNA profile has a peak in the 5' UTR. I would expect to see such a peak in ribosome footprinting data. Is it possible these are incorrectly labeled?

The figure is correctly labeled. Generally, one does not expect to see RPFs in the 5’UTR region unless there is an efficiently translated uORF, which appears not to be the case for MDH2.

In general, the information in this panel and C is inadequate. None of the numbers are clearly explained in the figure legend or in the figure.

We had cited the legend to Fig. S3C for details of all such gene browser images but have now inserted this information into the Fig. 1D legend, at the first occurrence of such data in the regular figures.

(3) Figures 1C and 1D are in the wrong order.

Corrected.

(4) Figure 2D is a very complicated Venn Diagram. I suggest using UpSet plots as an alternative to Venn diagrams to more clearly convey overlaps between sets.

We provided additional explanatory text in the Fig. 2D legend to facilitate understanding.

(5) The use of the same color scheme to represent different sets in panels of the same figure is a source of confusion. E.g. the cyan in Figures 2A, 2D, and 2E indicates unrelated categories, but one would think they are related.

The use of the same cyan color in these three figure panels actually does designate results for the same set of 591 mRNAs up-regulated in the three mutants. The application of the color schemes is now mentioned explicitly in Figs. 1, 2, and S3.

(6) Reporting of p-values = 0 in figures is not useful.

Corrected.

(7) The whole manuscript is extremely long which reduces the overall impact. For example, the introduction is six pages long. I suggest reducing redundant text and being more concise to enhance readability.

We tried to streamline the text wherever possible, in particular shortening the Introduction by two pages.

(8) Many abbreviations are used throughout the text that are not introduced the first time they are used.

Corrected throughout.

(9) The ERCC normalization is unclear. Were the spike-ins added before cell lysis to allow estimation of per-cell RNA counts or to the extracted RNA? If added to extracted RNA rather than cells it is not clear to me how the claim can be made regarding increased mRNA abundance in the mutants.

We thank the reviewer for this comment. As we explained in the Methods, 2.4 µl of 1:100 diluted ERCC RNA Spike-In Control Mix 1 was added to 1.2 µg of each total RNA sample prior to cDNA library preparation. Because the majority of total mRNA is comprised of rRNA, this normalization yields the abundance of each mRNA relative to rRNA. Owing to repression of rESR mRNAs encoding ribosomal proteins and biogenesis factors in the scd6∆edc3∆ strain (Fig. S3D), the ribosome content per cell is expected to be reduced in this mutant vs. WT. We showed previously that the isogenic dcp2∆ mutant that elicits an ESR response of similar magnitude, showed a 30% reduction in bulk ribosomal subunits per cell compared to same WT strain examined here {Vijjamarri, 2023 #7866}. Assuming a similar reduction in ribosome abundance in the scd6∆edc3∆ mutant, the changes in mRNA per cell conferred by the scd6∆edc3∆ mutation are expected to be 0.7-fold of the ERCCnormalized values given in Fig. 3E, yielding fold-changes of 2.00 and 0.62 for the mRNA_up and mRNA_dn, groups, respectively, which still differ substantially from the corresponding changes in normalized Rpb1 occupancies of 1.2 and 0.93, respectively. We have added this new analysis to the text of RESULTS.

(10) The use of the terms "up-regulated" and "derepressed" throughout is confusing. Both refer to observed increased abundance of mRNAs, but they imply different causes which are never clearly defined.

We changed all occurrences of “derepressed” to “up-regulated”.

Version 2

Abstract

Abstract

Degradation of many yeast mRNAs involves decapping by the Dcp1:Dcp2 complex. Previous studies on decapping activators Edc3 and Scd6 suggested their limited roles in mRNA decay. RNA-seq analysis of mutants lacking one or both proteins revealed that Scd6 and Edc3 have largely redundant activities in targeting numerous mRNAs for degradation that are masked in the single mutants. These transcripts are frequently targeted by decapping activators Dhh1 and Pat1, and the collective evidence suggests that Scd6/Edc3 act interchangeably to recruit Dhh1 to Dcp2. Ribosome profiling shows that redundancy between Scd6 and Edc3 and their functional interactions with Dhh1 and Pat1 extend to translational repression of particular transcripts, including a cohort of poorly translated mRNAs displaying interdependent regulation by all four factors. Scd6/Edc3 also participate with Dhh1/Pat1 in post-transcriptional repression of proteins required for respiration and catabolism of alternative carbon sources, which are normally expressed only in limiting glucose. Simultaneously eliminating Scd6/Edc3 increases mitochondrial membrane potential and elevates metabolites of the tricarboxylic acid and glyoxylate cycles typically observed only during growth in low glucose. Thus, Scd6/Edc3 act redundantly, in parallel with Dhh1 and in cooperation with Pat1, to adjust gene expression to nutrient availability by controlling mRNA decapping and decay.

Introduction

Degradation of mRNA is a key aspect of gene expression that can be regulated in response to nutrient availability, cell stress, and developmental pathways in eukaryotic cells and further serves to eliminate defective mRNAs. A major pathway of cytoplasmic mRNA turnover involves truncating the poly(A) tail by the Ccr4/Not and Pan2/Pan3 complexes, followed by removal of the m7G cap by the Dcp1/Dcp2 decapping complex and 5’ to 3’ exonucleolytic degradation by Xrn1. The decapping complex is activated by factors that interact with low-complexity sequence motifs in the C-terminal tail (CTT) of the catalytic subunit Dcp2, including Edc3, Scd6, DEAD-box helicase Dhh1, and Pat1, which also interact with one another extensively. There is evidence that Pat1 is recruited to oligoadenylate tails remaining on mRNAs following partial deadenylation, in association with the Lsm1-Lsm7 complex, and activates decapping via interactions with other decapping activators and with the Dcp2 CTT itself (Parker 2012) (He and Jacobson 2023)

Genome-wide analysis of mRNA abundance (RNA-seq) in yeast mutants lacking Dhh1, Pat1, or both factors revealed functional cooperation by Pat1 and Dhh1 (He et al. 2018), with a large fraction being up-regulated only in the pat1Δdhh1Δ double mutant (Vijjamarri et al. 2023a). The majority of mRNAs up-regulated by dhh1Δ or pat1Δ mutations are likewise up-regulated in dcp2Δ cells (He et al. 2018) and exhibit greater than average proportions of decapped mRNAs in WT cells but not in dhh1Δ or pat1Δ cells (Vijjamarri et al. 2023a), consistent with Dhh1/Pat1 targeting mRNAs for degradation via decapping. While more than half of the mRNAs up-regulated by dcp2Δ are up-regulated by dhh1Δ and/or pat1Δ, a large fraction are targeted primarily by the Upf factors instead responsible for nonsense-mediated mRNA decay (NMD) (Vijjamarri et al. 2023b) (Celik et al. 2017).

The cumulative contributions of Dhh1 and Pat1 to mRNA decay is consistent with their independent interactions with distinct segments of the Dcp2 CTT (He et al. 2018), which appears to be direct for Pat1 but bridged by Edc3 or Scd6 for Dhh1. There is also evidence for distinct decapping complexes containing either Dhh1 or Pat1 in addition to Xrn1, Edc3 or Scd6 (He et al. 2022). However, Dhh1 occupancy (Miller et al. 2018), tends to be elevated for mRNAs up-regulated by either dhh1Δ (23) or pat1Δ (Vijjamarri et al. 2023a), consistent with Dhh1 contributing to degradation of mRNAs targeted by Pat1. Moreover, Pat1 binding to the Dcp2 CTT was required for degradation of certain Dhh1-targeted mRNAs when Dhh1 recruitment to the CTT was compromised (He et al. 2022).

There is evidence that Edc3 is a common constituent of decapping complexes containing Xrn1 and one or more of the decapping activators Dhh1, Pat1, Scd6, and Upf1 (He et al. 2022). It is surprising therefore that Edc3 has been implicated in targeting only two transcripts for degradation, YRA1 and RPS28B (Badis et al. 2004; Dong et al. 2007). Edc3 shares sequence similarity with decapping activator Scd6 and both proteins contain an FDF motif shown to interact competitively with the Dhh1 homolog in animal systems (Tritschler et al. 2008; Tritschler et al. 2009). Edc3 and Scd6 also share N-terminal LSm domains that compete for binding to helical-leucine-rich (HLM) motifs in the CTT of fission yeast Dcp2 (Fromm et al. 2012). Analyzing effects of deleting the Edc3 interaction site in the S. cerevisiae Dcp2-CTT on levels of several Dhh1-repressed mRNAs suggested that recruitment of Dhh1 to Dcp2 can be mediated interchangeably by Edc3 or Scd6 bound to the same site in the Dcp2 CTT (He et al. 2022). These findings, plus the fact that deleting S. cerevisiae SCD6 and EDC3 simultaneously confers a synthetic growth defect (Decourty et al. 2008), suggest that yeast Scd6 and Edc3 function redundantly in targeting specific mRNAs for decapping and degradation.

Although tethering yeast Dhh1 or Scd6 enhances degradation of reporter mRNAs via Dcp1/Dcp2 (Carroll et al. 2011; Sweet et al. 2012; Zeidan et al. 2018), it is not well understood how these factors are targeted to specific native mRNAs. Dhh1 has been implicated in accelerating degradation associated with non-optimal codons in yeast mRNAs (Presnyak et al. 2015), being required for the rapid turnover conferred by suboptimal codons inserted in reporter mRNAs (Sweet et al. 2012). A queue of slowly elongating ribosomes upstream from non-optimal codons can be recognized by Dhh1, and overexpressing or tethering Dhh1 evokes ribosome stalling at non-optimal codons. Moreover, Dhh1 association and Dhh1-dependent repression of mRNA abundance both correlate with codon non-optimality across the yeast transcriptome (Radhakrishnan et al. 2016). However, the mRNAs most highly up-regulated in dhh1Δ and pat1Δ cells are not enriched for suboptimal codons, suggesting that other features are responsible for their preferential targeting by Dhh1 or Pat1 for decapping/decay (Vijjamarri et al. 2023a).

In addition to enhancing mRNA decay, there is evidence that Pat1, Dhh1, and Scd6 can repress translation. Tethering Dhh1 or Scd6 represses translation of reporter mRNAs in dcp2Δ cells where the tethered transcripts cannot be decapped and degraded (Carroll et al. 2011; Sweet et al. 2012; Zeidan et al. 2018). Deletion of Dhh1 and Pat1 simultaneously eliminated loss of bulk polysomes evoked by nutrient starvation and also increased initiation rates of certain mRNAs (Holmes et al. 2004) (Coller and Parker 2005) (Arribere et al. 2011). Supporting a direct role in repressing translation, overexpressing Dhh1 or Pat1 in nonstarved cells evoked polysome disassembly and reduced the initiation rate of specific mRNAs; and addition of Dhh1 (Coller and Parker 2005) or N-terminally truncated Pat1 (Nissan et al. 2010) to yeast extracts inhibited bulk translation and 48S preinitiation complex (PIC) assembly in vitro. Ribosome profiling studies of dhh1Δ, pat1Δ and pat1Δdhh1Δ mutants identified hundreds of genes whose transcripts are translationally down-regulated or activated by Dhh1 or Pat1 in nutrient-replete cells (Jungfleisch et al. 2017) (Radhakrishnan et al. 2016; Zeidan et al. 2018), which frequently involves cooperation between Dhh1 and Pat1 (Vijjamarri et al. 2023a).

Recently, we showed that Pat1 and Dhh1 function with the decapping enzyme in rich medium to repress the abundance or translation of numerous mRNAs encoding proteins required specifically in media containing an alternative carbon or nitrogen source (Vijjamarri et al. 2023a; Vijjamarri et al. 2023b). These include mitochondrial proteins involved in oxidative phosphorylation (Ox. Phos.) and diverse other proteins known to be transcriptionally repressed by carbon or nitrogen catabolite repression in rich medium.

In this study, we used a multi-omics approach to determine whether Edc3 and Scd6 have largely redundant functions in targeting mRNAs for decapping and attendant degradation, and whether they functionally cooperate with Pat1 or Dhh1 in repressing the abundance or translation of specific mRNAs. We identified a large cohort of mRNAs that are up-regulated in an scd6Δedc3Δ double mutant but not in either single mutant lacking only Scd6 or Edc3, without a commensurate increase in transcription of the cognate genes. These transcripts display a strikingly similar pattern of up-regulation in the dhh1Δ mutant in the manner predicted if Edc3/Scd6 redundantly recruit Dhh1 to Dcp2 for activation of decapping (He et al. 2022). We further observed functional redundancy between Scd6 and Edc3 and extensive cooperation with Dhh1/Pat1 in repressing translation of particular mRNAs, with evidence for interdependent repression by all four decapping factors. Importantly, Edc3/Scd6 contribute to post-transcriptional repression of proteins required for catabolism of non-preferred carbon or nitrogen sources on rich medium, acting collectively to enhance glucose repression, maintain low-level mitochondrial electron transport and reduce levels of tricarboxylic acid (TCA) and glyoxylate cycle intermediates in glucose-replete cells.

Methods

Yeast strains and plasmids

Unless otherwise indicated, the following yeast strains were employed for all experiments: WT strain HFY114 (W303: MATa ade2-1 ura3-1 his3-11,15 trp1-1 leu2-3,112 can1-100) (He et al. 2003), scd6Δ strain SYY2352 (MATa ade2-1 ura3-1 his3-11,15 trp1-1 leu2-3,112 can1-100 scd6Δ::kanMX6) (He and Jacobson 2015), edc3Δ strain FZY862 (MATa ade2-1 ura3-1 his3-11,15 trp1-1 leu2-3,112 can1-100 edc3Δ::kanMX4), and scd6Δedc3Δ strain FZY858 (MATa ade2-1 ura3-1 his3-11,15 trp1-1 leu2-3,112 can1-100 scd6Δ::hphMX4 edc3Δ::kanMX4).

Strains FZY858 and FZY862 were generated by replacing chromosomal EDC3 in strains FZY855 and HFY114, respectively, with the edc3Δ::kanMX4 allele amplified by PCR from the chromosomal DNA of strain 255 (MATa his3Δ1 leu2Δ0 met15Δ0 ura3Δ0 edc3Δ::kanMX4) obtained from Research Genetics. DNA sequences up to 310 bp upstream and 340 bp downstream of the EDC3 coding sequence were included in the amplified fragment used for transformation of FZY855 and HFY114 to G418-resistance. FZY855 was derived from SYY2352 by transforming with marker-swap plasmid pAG32 (Goldstein and McCusker 1999) to replace scd6Δ::kanMX6 with scd6Δ::hphMX6, selecting for resistance to hygromycin and screening for loss of G418-resistance. The presence of all deletion alleles was verified by PCR analysis of chromosomal DNA.

Strains H5695, H5696, H5697, and H5698 strains, employed for co-immunoprecipitation experiments, were generated by PCR-amplification of a DHH1-TAP-HIS3MX fragment from strain F2262 using primers AKV224 and AKV225, containing 474 bp upstream and 450 bp downstream of the DHH1 stop codon, that was used to transform strains HFY114, FZY862, SYY2352, and FZY858, respectively, to His⁺. A complete list of yeast strains employed is given in Table S1.

Plasmids pLfz614-7 and pLfz635-5 contain the EDC3 CDS with 420 bp upstream and 490 bp downstream, on a ∼2.6 kb fragment, and plasmids pL615-5 and pLfz636-1 contain the SCD6 CDS with 500 bp upstream and 180 bp downstream on a ∼1.7 kb fragment, amplified by PCR from yeast genomic DNA and inserted between the XhoI and EcoRI sites of YCplac33 or YCplac111, respectively. Plasmid pAK133 containing DCP2-3HA was produced by a multi-step PCR amplification scheme. First, the DCP2 CDS with 920 nucleotides upstream of the start codon and 587 nucleotides downstream of the stop codon was amplified from genomic DNA of WT strain HFY114. Sequences encoding the 3XHA tag were introduced immediately upstream of the DCP2 stop codon using primer pairs AKV372/AKV373 and AKV374/AKV375 during the first round of PCR amplification; a second round of PCR followed using primers AKV372 and AKV375 to amplify a complete DCP2-3HA fragment also containing sequences homologous to the ends of plasmid YCplac33 linearized by digestion with BamHI. The PCR product was fused into BamHI-digested YCplac33 using the Takara In-Fusion cloning kit (#638956) according to the manufacturer’s instructions. Primers AKV372 and AKV375 were designed using the Takara on-line primer design tool to ensure optimal homologous recombination during the cloning process. The inserted yeast DNA fragments were verified by sequencing in their entirety. A complete list of plasmids employed is given in Table S2 and the list of primers in Table S3.

Cell spotting growth assays

Yeast transformants harboring plasmids containing EDC3, SCD6, or empty vector were grown to mid-logarithmic phase at 30°C in liquid synthetic complete medium (SC) without uracil (SC-U). Cultures were diluted to OD600 of 1.0 and 10-fold serial dilutions were spotted on agar medium of the same composition and incubated at 30°C or 37°C for 2 days.

Polysome profiling

Polysome profiling was conducted as described previously (Zeidan et al. 2018). In brief, strains were cultured in YPD medium at 30°C to mid-logarithmic phase (OD600 of ∼0.5-0.6). Fifteen A260 units of WCEs were resolved on a 10–50% sucrose gradient by centrifugation at 35,000 rpm. Gradients were fractionated with continuous scanning at 260 nm. Areas under the A260 tracings of polysome and monosome peaks were calculated using ImageJ software and used to calculate polysome to monosome (P/M) ratios in mutants and WT.

Ribosome footprint profiling

Duplicate cultures of yeast strains HFY114 (WT W303), SYY2352 (scd6Δ), FZY862 (edc3Δ), and FZY858 (scd6Δedc3Δ) were cultured in YPD medium at 30°C to OD600 ∼0.6-0.7. Yeast cells were quickly filtered, frozen in liquid nitrogen and stored at -80°C. Cells were lysed in a freezer mill and ribosome protected mRNA fragments (RPF) were isolated and used for cDNA library preparation as described previously (Vijjamarri et al. 2023b). Single-end 100 bp Illumina sequencing was performed by the National Heart, Lung and Blood Institute (NHLBI) DNA Sequencing and Genomics Core facility (Bethesda, MD).

RNA sequencing with spike-in normalization and CAGE analysis

The same lysates used for RPF library preparation were used to prepare RNA-Seq libraries after adding ERCC RNA Spike-In Control Mix 1 (Thermo Fisher Scientific, Cat. # 4456740). Total RNA was extracted from cell lysates using QIAzol Lysis reagent (Qiagen Cat. # 79306) and miRNeasy Mini Kit (Qiagen, Cat. 217004). Twenty µg of total RNA of each sample was subjected to RNase-free DNase I (Roche, Cat. # 04716728001) treatment and then processed with the RNA Clean and Concentrator Kit (Zymo, Cat. R1018). An aliquot of 2.4 µl of 1:100 diluted ERCC RNA Spike-In Control Mix 1 was added to 1.2 µg of each RNA sample and submitted to the NHLBI DNA Sequencing and Genomics Core facility (Bethesda, MD) for cDNA library preparation and Illumina sequencing. CAGE sequencing and RNA-Seq were conducted in parallel on two biological replicates of total RNA isolated from the same WT and scd6Δedc3Δ strains, cultured under the same conditions as employed for ribosomal profiling, exactly as described previously (Vijjamarri et al. 2023a).

Rpb1 ChIP-Seq with spike-in normalization

Triplicates of wild type and scd6Δedc3Δ yeast strains were cultured in rich YPD medium at 30°C to OD600=0.6∼0.8. Chromatin extracts were prepared from formaldehyde cross-linked cells as described previously (Qiu et al. 2016). S. pombe chromatin was added to each chromatin sample as a spike-in control prior to immunoprecipitation with Rpb1 antibodies, and ChIP-Seq DNA libraries were prepared as described previously (Zheng et al. 2023) subjected to 50 bp paired-end Illumina sequencing by the NHLBI DNA Sequencing and Genomics Core facility (Bethesda, MD) and analyzed as previously described (Zheng et al. 2023).

Western blot analysis

For the results in Fig. 6B-C, WCEs were prepared by trichloroacetic acid (TCA) extraction as previously described (Reid and Schatz 1982) and immunoblot analysis was conducted as described previously (Nanda et al. 2009). After electroblotting to nitrocellulose membranes (Bio-Rad 1620094), membranes were probed with antibodies against Atp20, Cox14, Pet10, Qcr8, Sdh4 (kindly provided by Dr. Nikolaus Pfanner), Cyb2 (kindly provided by Dr. Thomas Fox), Idh1 (Abnova, PAB19472), Cit2 (Antibodies-online.com, ABIN4889057), and Gcd6 (Bushman et al. 1993). Secondary antibodies employed were HRP-conjugated anti-rabbit (Cytiva, NA9340V), anti-mouse IgG (Cytiva, NA931V) and anti-goat IgG (Abnova, PAB29101). Detection was performed using enhanced chemiluminescence (ECL) Western Blotting Detection Reagent (Cytiva, RPN2016) and the Azure 200 gel imaging biosystem. NIH Image J was employed to analyze images for quantification. For analysis of Cox2 levels in Fig. 6D, total proteins were TCA-extracted as described previously (Rashida et al. 2021) and subjected to Western blot analysis as recently described (Niphadkar et al. 2024) using mouse monoclonal antibody MTCO2 (4B12A5) from Invitrogen.

TMT-MS analysis of global protein abundance

TMT-MS analysis was conducted as described previously (Vijjamarri et al. 2023a) with the following modifications. Replicate cultures of WT, edc3Δ, scd6Δ and scd6Δedc3Δ strains were cultured in YPD medium for ∼3 doublings to OD600 of ∼0.6, and harvested by centrifugation for 5 min at 3000 x g. Cells were resuspended in nuclease-free water, collected by centrifugation, and stored at -80°C. WCEs were prepared in freshly prepared 8M Urea, 25 mM triethylammonium-bicarbonate (TEAB; Thermo Scientific, 90114) by washing the cell pellets once and resuspending again in the extraction buffer, then vortexing with glass beads in the cold room. Lysates were clarified by centrifugation at 13,000 x g for 30 min and the quality of extracted proteins was assessed following SDS-PAGE using GelCode™ Blue Stain (Thermo Scientific, 24592) and quantified with the Pierce™ BCA Protein Assay Kit (Thermo Scientific, 23225). Lysates were stored at -80°C. Sample preparation, TMT-MS/MS (Zecha et al. 2019) and data analysis were performed at the IDeA National Resource for Quantitative Proteomics.

Co-immunoprecipitation analysis

Transformants of DHH1-TAP strains were cultured in 50 ml YPD at 30°C to OD₆₀₀ ≈ 0.6–0.8, harvested by centrifugation, and washed with 1 ml lysis buffer (LB) containing 0.5 mM PMSF. Cell pellets were resuspended in 500 µl LB supplemented with 0.5 mM PMSF and 1x protease inhibitor cocktail (Roche Cat # 5056489001) and lysed by vortexing in 200 µl glass beads (8 cycles: 30 sec on, 45 sec off) on ice in a cold room. Lysates were clarified by centrifugation (15,000 rpm, 20 min, 4°C), and protein concentrations determined using the Bradford assay. Approximately 1 µg of lysate was incubated with IgG sepharose beads pre-treated with 5% BSA for 2 h at 4°C. Beads were washed five times with LB containing 0.5 mM DTT. Bound complexes were eluted with 500 µl LB + 0.5 mM DTT containing 2 µl TEV protease (1 mg/ml, 10 U/µl; Invitrogen 12575-015) at room temperature for 2 h. Eluates were precipitated with 10% cold TCA on ice for 10 min, pelleted by centrifugation (13,000 rpm, 10 min, 4°C), washed with cold acetone, air-dried, resuspended in 50 µl 2× SDS-PAGE sample buffer, neutralized with 50 µl unadjusted 1 M Tris, and boiled for 5 min. Samples were subjected to immunoblot analysis as described above using rabbit polyclonal TAP antibodies (Thermo Fisher Scientific Cat# CAB1001) and mouse monoclonal antibodies against HA (12CA5) (Roche 10522600) for detecting TAP and HA tags.

Measuring mitochondrial membrane potential

Precultures were grown in SC-Ura (to select for the URA3 plasmids) to OD600 of ∼3.0 and used to inoculate YPD medium at OD600 of 0.2. Cells were grown to OD600 of ∼0.6-0.8, incubated with 500 nM TMRM for 1 h and subjected to flow cytometry to measure dye fluorescence in individual cells as described previously (Vijjamarri et al. 2023b). In control samples, 50 µM FCCP was added to dissipate the membrane potential and reveal non-specific background fluorescence. Results are presented in arbitrary fluorescence units normalized to OD600 of the cultures.

Analysis of polar metabolites of intermediary metabolism

Yeast cell culturing

Four replicate 15 mL cultures in YEPD medium were prepared for each strain by inoculating with saturated overnight cultures to OD600 of 0.1, culturing with shaking at 30°C, and harvesting at OD600 of 0.6 by centrifugation in conical 15mL tubes for 3 min at 3000 x g in an Avanti J-HC centrifuge pre-cooled to -10°C. Cell pellets were resuspended in 1.8 mL of ice-cold PBS and centrifuged in 2 mL screw-cap tubes in a refrigerated microfuge for 30 s. Supernatants were decanted, tube rims blotted on tissue paper, and cell pellets frozen in dry ice/ethanol for 5 min and stored at -80C. Frozen cell pellets were shipped on dry ice to the NYU Metabolomics Core Resource Laboratory, where all subsequent analyses described below were conducted.

Extraction of metabolites

Frozen cell pellets were thawed on wet ice. Extraction buffer, consisting of 80% methanol (Fisher Scientific) and 500 nM metabolomics amino acid mix standard (Cambridge Isotope Laboratories, Inc.), was prepared and placed on dry ice. Samples were extracted by mixing cell pellets with extraction buffer at 10mg/mL (determined by sample OD measurements) in 2.0 mL screw-cap vials containing ∼100 µL of disruption beads (Research Products International, Mount Prospect, IL). Each sample was homogenized for 10 cycles on a bead blaster homogenizer (Benchmark Scientific, Edison, NJ), with each cycle consisting of 30 sec homogenization at 6 m/s followed by a 30 sec pause. Samples were centrifuged at 21,000 g for 3 min at 4°C. Aliquots of 450 µL were transferred to 1.5 mL tubes, dried in a Speedvac (Thermo Fisher, Waltham, MA), reconstituted in 50 µL of Optima LC/MS grade water (Fisher Scientific, Waltham, MA), sonicated for 2 min, and centrifuged at 21,000 g for 3 min at 4°C. Aliquots of 20 µL were transferred to LC vials containing glass inserts for analysis, and the remaining samples stored at -80°C.

LC-MS/MS with the hybrid metabolomics method

Samples were subjected to an LC-MS analysis to detect and quantify known peaks. Extraction of polar metabolites was carried out on each sample based on a previously described method (Jones et al. 2014). LC was conducted using a Millipore™ ZIC-pHILIC (2.1 x150 mm, 5 μm) column coupled to a Dionex Ultimate 3000™ system with gradient elution conducted at 25°C with a flow rate of 100 μL/min using the following buffersA) 10 mM ammonium carbonate in water, pH 9.0, and B) neat acetonitrile. The gradient profile was as follows; 80-20% B (0-30 min), 20-80% B (30-31 min), 80-80% B (31-42 min). Injection volume was set to 2 μL for all analyses with a 42 min total run time per injection.

MS was conducted by coupling the LC system to a Thermo Q Exactive HF™ mass spectrometer operating in heated electrospray ionization mode (HESI). Method duration was 30 min with a polarity switching data-dependent Top 5 method for both positive and negative modes. Spray voltage for both positive and negative modes was 3.5kV and capillary temperature was set to 320 °C with a sheath gas rate of 35, aux gas of 10, and max spray current of 100 μA. The full MS scan for both polarities utilized 120,000 resolution with an AGC target of 3e6 and a maximum IT of 100 ms, and the scan range was from 67-1000 m/z. Tandem MS spectra for both positive and negative mode used a resolution of 15,000, AGC target of 1e5, maximum IT of 50 ms, isolation window of 0.4 m/z, isolation offset of 0.1 m/z, fixed first mass of 50 m/z, and 3-way multiplexed normalized collision energies (nCE) of 10, 35, 80. The minimum AGC target was 1e4 with an intensity threshold of 2e5. All data were acquired in profile mode.

Relative quantification and statistical analyses of polar metabolites

The resulting ThermoTM RAW files were converted to SQLite format using an in-house python script to enable downstream peak detection and quantification. The available MS/MS spectra were first searched against the NIST17 MS/MS (Simon-Manso et al. 2013), METLIN (Smith et al. 2005) and respective Decoy spectral library databases using an in-house data analysis python script adapted from our previously described approach for metabolite identification false discovery rate control (FDR) (Wang et al. 2018; Wang et al. 2020). Metabolite peaks were extracted based on the theoretical m/z of the expected ion type, e.g., [M+H]+, with a 15 part-per-million (ppm) tolerance and a ± 0.2 min peak apex retention time tolerance within an initial retention time search window of ±0.5 min. For all the group-wise comparisons, t-tests were performed using the Python SciPy (1.5.4) (Virtanen et al. 2020) library to test for differences and generate statistics for the downstream analyses. For the pairwise t-tests, any metabolite with a p-value < 0.01 was considered significantly regulated (up- or down-) for prioritization in the subsequent analyses.

Coverage of a library of polar metabolites of major pathways of intermediary metabolism allowed detection of 130 of the 147 metabolites examined in at least 4 samples, and 91 detected in all 20 samples after background threshold correction. Instrument performance was assessed using the internal standards added to the samples during metabolite extraction and instrument mass accuracy was within tolerance (-2.0 ppm), LC column performance was stable (0.16 min RT range) and internal standard response variability was 13% across the samples. The resulting data were analyzed by principal components analysis (Fig. 7A), by constructing heatmaps with unsupervised hierarchical clustering of the imputed matrix values utilizing the R library pheatmap (1.0.12). GraphPad Prism 9 (9.4.1, GraphPad Software, San Diego, CA), and volcano plots (generated utilizing R library script Manhattanly (0.2.0)), in addition to the statistical comparisons summarized in Supplementary File S7.

13C6-glucose metabolic flux measurements

Three replicate cultures of each strain were grown in YP medium with 2% unlabeled glucose to OD600 of 0.6-0.7 and then shifted to YP with 1% unlabeled glucose and cultured for 20 min. 13C6-labelled glucose (1%) was then added and growth was continued for 8 min. Cells were collected, and metabolites were extracted, derivatized and subjected to mass spectrometry as described previously, with the addition of additional masses for assessing the label incorporation into the specific labelled intermediates that were analyzed (Walvekar et al. 2018).

ATP Measurements

Cells were cultured in YPD to OD600 of 0.6-0.7 and treated or untreated with 5 mM sodium azide for 30 min at 30°C. Five OD600 units of cells were harvested and resuspended in 300 µL ice-cold 5% trichloroacetic acid (TCA), incubated on ice for 15 min, and diluted 50-fold in 20 mM Tris-HCl (pH 7.5) to 0.1 % TCA, after which 10 µL was mixed with 90 µL of ATP Determination reaction mixture (Cat. # A22066, Thermo-Fisher Scientific) and incubated at room temperature for 5 min. Units of Firefly Luciferase were measured in a Berthold Centro XS3 LB 960 Luminometer, and ATP levels were calculated from an ATP standard curve generated using the same assay.

Additional data visualization and statistical analyses

Notched box-plots were constructed using a web-based tool at http://shiny.chemgrid.org/boxplotr/. In all such plots, the upper and lower boxes contain the 2nd and 3rd quartiles and the band gives the median. If the notches in two plots do not overlap, there is roughly 95% confidence that their medians differ. Density scatter plots, Venn diagrams, significance testing of gene set overlaps in Venn diagrams using the hypergeometric distribution, hierarchical clustering analysis, and gene ontology (GO) analysis all were conducted as described previously (Vijjamarri et al. 2023a).

Results

Evidence that Scd6 and Edc3 functionally cooperate to control the abundance of many individual mRNAs

To determine whether Scd6 and Edc3 function redundantly in post-transcriptional control of gene expression, we constructed a scd6Δedc3Δ double mutant isogenic to the scd6Δ and edc3Δ single mutants we examined previously (Zeidan et al. 2018). Only the double mutant exhibits a marked slow-growth (Slg-) phenotype on synthetic complete medium (SC), which was largely complemented by introducing either SCD6 or EDC3 on a single copy plasmid (Supplemental Fig. S1A). Analysis of polysome assembly revealed a ∼40% reduction in ratio of polysomes to monosomes (P/M) in the scd6Δedc3Δ double mutant, whereas the single mutants showed little (scd6Δ) or no (edc3Δ) reduction in bulk translation by this assay (Fig. S1B). These results suggest that Scd6 and Edc3 act redundantly to carry out one or more functions required for WT levels of bulk translation and cell growth in nutrient replete cells.

To examine the effects of the scd6Δ, edc3Δ, and scd6Δedc3Δ mutations on the abundance and translation of individual mRNAs, we conducted RNA-Seq and ribosome profiling (Ribo-Seq) of the mutant and WT strains following growth in liquid rich medium (YPD) at 30°C (processed data compiled in supplementary File S1). Ribo-Seq entails deep-sequencing of ribosome-protected fragments (RPFs, or ribosome footprints), and cycloheximide was added to the lysates to arrest elongating ribosomes on the mRNA following cell breakage. The ratio of RPF sequencing reads summed over the coding sequences (CDS) to the total mRNA reads from RNA-Seq for the corresponding transcript provides a measure of translational efficiency (TE) for each mRNA (Ingolia et al. 2009). The ribosome profiling and RNA-Seq results between two biological replicates for each strain were highly reproducible with Pearson correlation coefficients (r) ranging between 0.95-1.0 for different pairwise comparisons of replicates (Fig. S2A-B). We employed DESeq2 (Love et al. 2014) to identify statistically significant differences in relative mRNA abundance, RPF abundance, or TE for all expressed mRNAs between WT and mutant strains (see Methods for details).

Analysis of the RNA-Seq results identified 81 mRNAs that were significantly up-regulated in the edc3Δ mutant vs. WT by >1.5-fold at a false discovery rate (FDR) of <0.05 (dubbed mRNA_up_e3 transcripts; Fig. S3A), and 123 mRNAs reduced in abundance by edc3Δ by the same criteria (dubbed mRNA_dn_e3, Fig. S3B). Only 14 mRNAs were up-regulated and only 34 down-regulated by the scd6Δ single mutation (respectively, mRNA_up_s6 and mRNA_dn_s6, Fig. S3A-B). Importantly, many more mRNAs were dysregulated by the scd6Δedc3Δ double mutation: 741 in the mRNA_up_s6,e3 group and 793 in the mRNA_dn_s6,e3 group (Fig. S3A-B), indicating that the two factors have highly redundant functions in controlling mRNA abundance.

Many yeast mutants with Slg- phenotypes, including pat1Δ, dhh1Δ, and dcp2Δ deletion mutants, exhibit altered expression of most mRNAs belonging to the Environmental Stress Response (ESR) (O’Duibhir et al. 2014), which includes ∼300 induced (iESR) and ∼600 down-regulated (rESR) mRNAs dysregulated in WT cells by various stresses (Gasch et al. 2000). In keeping with its Slg- phenotype, the scd6Δedc3Δ mutation conferred a marked reduction in median expression of rESR mRNAs, and increased expression of the iESR mRNAs, which exceeded in magnitude the changes observed for the slowest growing yeast deletion mutants analyzed previously (O’Duibhir et al. 2014) (Fig. S3C-D). The two single mutations, by contrast, conferred much smaller changes in ESR mRNAs (Fig. S3C-D). (In all box plots, when notches do not overlap between adjacent boxes, their two medians differ with 95% confidence; and when notches do not overlap 0 in log2 plots, the median differs significantly from that of all mRNAs, which is invariably close to 1.0.) Consistent with these results, the transcripts up-regulated in the double mutant are enriched for iESR mRNAs (Fig. 1A), suggesting that the 187 iESR transcripts up-regulated in this strain are responding indirectly to cell stress. The remaining 75% of mRNAs up-regulated in scd6Δedc3Δ cells are not iESR mRNAs however (Fig. 1A), suggesting that their increased abundance arises from eliminating Edc3/Scd6 functions in mRNA decay. Below, we excluded the ESR mRNAs from analyses of mRNA changes in an effort to focus on the transcripts controlled directly by Scd6 and Edc3.

Most mRNAs up-regulated in the scd6Δedc3Δ mutant are not iESR transcripts and exhibit Scd6/Edc3 functional redundancy in repression of transcript abundance.

(A) Venn diagram of overlap between the 741 mRNAs up-regulated in the scd6Δedc3Δ mutant vs. WT (mRNA_up_s6,e3) and the 283 induced ESR (iESR) mRNAs, indicating fold-enrichment and P value of overlap determined by the hypergeometric distribution. (B) Venn diagram of overlaps involving all 591 non-iESR mRNAs up-regulated in abundance by scd6Δ (6 Non-iESR mRNA_up_s6 transcripts), edc3Δ (64 mRNA_up_e3 transcripts) or scd6Δedc3Δ mutations (554 Non-iESR mRNA_up_s6,e3 transcripts). (C) Notched box-plot analyses of log2 changes in mRNA abundance (log2ΔmRNA) determined by DESeq2 analysis between the indicated mutants vs. WT for mRNAs belonging to the specified sectors of the Venn diagram in (B). The numbers of mRNAs in each group for which data were obtained are indicated immediately above the x-axis; un-logged median values are indicated for each column at the bottom. (D) Gene browser image for MDH2 showing the mRNA (top 16 tracks) and RPF (bottom 16 tracks) reads measured by parallel RNA-Seq and ribosome profiling analyses for two biological replicates of WT and the indicated mutants with fold-changes in mRNA or RPFs between mutant and WT indicated to the right of each track. In panels A-D, results for the scd6Δ, edc3Δ, and scd6Δedc3Δ mutations, abbreviated as Δs6, Δe3, and Δs6,Δe3, are shown in grey, white or light grey, and cyan, respectively.

Considering only mRNAs not governed by the ESR, we identified 591 non-iESR mRNAs significantly up-regulated in any of the three mutants (Fig. 1B; eg. Non-iESR mRNA_up_s6,e3 designating the 554 mRNAs up-regulated in the double mutant). Examining the ΔRNA values for the majority fraction of the Non-iESR transcripts up-regulated only in the double mutant reveals little change in median abundance in each single mutant but strong up-regulation in the double mutant (Figs. 1B & 1C, sector (iii)), as expected for redundant repressive functions of Scd6 and Edc3. MDH2, encoding an enzyme of the glyoxylate cycle, exemplifies a non-iESR transcript up-regulated in both mRNA and RPF abundance exclusively in the double mutant (Fig. 1D). The small fraction of 27 mRNAs significantly up-regulated in both the edc3Δ single mutant and double mutant exhibits only slightly elevated median abundance in the scd6Δ single mutant, and only slightly greater up-regulation in the double mutant vs. edc3Δ single mutant (Fig. 1B & 1C, sector (ii)), indicating a minimal repressive contribution by Scd6. Interestingly, the small set of 37 mRNAs significantly up-regulated only in the edc3Δ single mutant shows reduced rather than increased abundance in the scd6Δ single mutant, and lower up-regulation in the double mutant vs. the edc3Δ single mutant (Fig. 1B & C, sector (i)), suggesting that Scd6 enhances rather than represses these mRNAs, especially in edc3Δ cells.

Evidence that Dhh1 and Pat1 functionally cooperate with Scd6/Edc3 in repressing mRNA abundance

We recently identified a group of 1018 non-iESR mRNAs up-regulated by either pat1Δ, dhh1Δ, or pat1Δdhh1Δ mutations (Vijjamarri et al. 2023a). Importantly, this group is highly enriched for the 591 mRNAs up-regulated by scd6Δ, edc3Δ, or scd6Δedc3Δ mutations (Fig. 2A). Indeed, ∼70% of the transcripts up-regulated by scd6Δ/edc3Δ are also up-regulated in one of the three pat1Δ/dhh1Δ mutants (Fig. 2A, sector (ii)), showing comparable up-regulation in the scd6Δedc3Δ, pat1Δ, and dhh1Δ mutants, but little change in the scd6Δ and edc3Δ single mutants (Fig. 2B(ii)). These mRNAs generally exhibit the greatest up-regulation in the pat1Δdhh1Δ double mutant, indicating cumulative contributions of Dhh1 and Pat1 to their repression, in contrast to the largely redundant roles played by Scd6 and Edc3 in their repression. As expected, the smaller group of 180 mRNAs significantly up-regulated exclusively by scd6Δ/edc3Δ (sector (i) of Fig. 2A) shows strong up-regulation only in the scd6Δedc3Δ double mutant; but they are appreciably up-regulated by dhh1Δ while being largely unaffected by pat1Δ (Fig. 2B, sector (i)). Finally, the majority group of 607 mRNAs significantly up-regulated by only pat1Δ or dhh1Δ display the largest increases in the two mutants lacking Pat1, but smaller and similar increases in response to the scd6Δ/edc3Δ and dhh1Δ mutations (Figs. 2A & 2B, sector (iii)). Overall, these findings suggest that Dhh1 and Pat1 both contribute to repressing the majority of mRNAs repressed redundantly by Scd6 and Edc3, with Dhh1 contributing more extensively than Pat1.

Most mRNAs up-regulated in the scd6Δedc3Δ mutant are also up-regulated by dhh1Δ and pat1Δ.

(A) Venn diagram of overlap between all 591 non-iESR mRNAs up-regulated in abundance by the scd6Δ, edc3Δ, or scd6Δedc3Δ mutations (from Fig. 1B) and 1018 non-iESR mRNAs up-regulated by the dhh1Δ, pat1Δ, or pat1Δdhh1Δ mutations identified previously (Vijjamarri et al. 2023a). (B) Notched box-plots of log2ΔmRNA between the indicated mutants vs. WT for mRNAs in the three sectors specified in (A). (C) Hierarchical clustering analysis of log2ΔmRNA values conferred by the indicated mutations vs. WT for 784 of the 794 mRNAs up- or down-regulated in dhh1Δ vs. WT cells for which RNA-Seq data was obtained in all four strains and with log2ΔmRNA values >-5 and <5, conducted with R heatmap.2 function from R ‘gplots’ library, using default hclust hierarchical clustering algorithm. Spearman coefficients (ρ) and associated P values are given for the indicated correlation analyses. (D) Venn diagram of overlaps between the 591 non-iESR mRNAs up-regulated by the scd6Δ/edc3Δ mutations vs. WT (from Fig. 1B) and the indicated 431 and 590 non-iESR mRNAs up-regulated by dhh1Δ or pat1Δ vs. WT, respectively, identified previously (Vijjamarri et al. 2023a). The 220, 78, and 293 mRNAs found exclusively in only one of the three sets are indicated in the cyan, red-orange, and yellow sectors, respectively while the 127, 71 and 53 mRNAs shared by two of the three sets and the 173 mRNAs shared by all three sets, are indicated in the corresponding regions of overlap. (E) Notched box-plots of log2ΔmRNA between the indicated mutants vs. WT for the 431 non-iESR mRNAs up-regulated by dhh1Δ vs. WT shown in panel (D). In panels A, B, D & E, results for mRNAs up-regulated by the scd6Δ, edc3Δ, and scd6Δedc3Δ mutations are shown in grey, white, and cyan, respectively, while those for mRNAs up-regulated by the pat1Δ, dhh1Δ, and pat1Δdhh1Δ mutations are shown in yellow, orange, and red, respectively.

Further evidence for this last point comes from a k-means clustering analysis of mRNA changes in different mutants for the group of 431 non-iESR mRNAs up-regulated by dhh1Δ, which revealed a stronger correlation between the changes conferred by scd6Δ/edc3Δ and dhh1Δ (ρ = 0.90) than between pat1Δ with dhh1Δ (ρ = 0.74) or scd6Δ/edc3Δ (ρ = 0.71) (Fig. 2C). Consistent with this, the scd6Δ/edc3Δ mutations up-regulate a considerably larger proportion of the mRNAs significantly up-regulated by dhh1Δ (∼70%) vs. those up-regulated by pat1Δ (∼41%) (Fig. 2D). Furthermore, the majority of mRNAs up-regulated by dhh1Δ exhibit redundant repression by Scd6 and Edc3, showing marked up-regulation comparable to that given by dhh1Δ itself only in the scd6Δ/edc3Δ double mutant (Fig. 2E). These findings are consistent with the involvement of Edc3/Scd6 in the degradation of most mRNAs targeted by Dhh1, which are more variably and less extensively repressed by Pat1.

Evidence that Scd6/Edc3 repress mRNA abundance by enhancing decapping/degradation and facilitating Dhh1 association with Dcp2

We next examined whether the mRNAs significantly down-regulated by Edc3/Scd6 are also regulated by the decapping enzyme Dcp1/Dcp2. Supporting this, cluster analysis of mRNA changes revealed that the majority of non-iESR mRNAs up-regulated in the scd6Δedc3Δ double mutant are also up-regulated by dcp2Δ (Fig. 3A, blue colors in cols. 1 & 3), with a strong correlation between the abundance changes conferred by these mutations relative to WT (ρ=0.74). In addition, dcp2Δ increased the median abundance of this group of mRNAs similarly to that given by scd6Δedc3Δ (Fig. 3B, cols. 1 & 4). These findings implicate decapping by Dcp1:Dcp2 as an important driver of the repression of mRNA levels directed by Scd6/Edc3. The cluster analysis in Fig. 3A once again reveals greater similarity between the mRNA changes conferred by scd6Δedc3Δ and those given by dhh1Δ (ρ = 0.90) versus pat1Δ (ρ = 0.69); and pat1Δ also confers a smaller median reduction than dhh1Δ of the Scd6/Edc3-down-regulated mRNAs (Fig. 3B).

Evidence that impaired decapping not increased transcription drives changes in mRNA abundance in scd6Δedc3Δ cells and that Edc3/Scd6 redundantly promote Dhh1 association with Dcp2.

(A) Hierarchical clustering analysis of log2ΔmRNA values conferred by the indicated mutations vs. WT for 741 of the 1052 mRNAs up- or down-regulated in scd6Δedc3Δ vs. WT cells for which RNA-Seq data was obtained in all five strains and with log2ΔmRNA values >-5 and <5, conducted as in Fig. 2C, showing Spearman coefficients and P values for indicated correlations. (B) Notched box-plots of log2ΔmRNA between the indicated mutants vs. WT for the 554 non-iESR mRNAs up-regulated by scd6Δedc3Δ vs. WT (shown in Fig. 1B). (C) Ratios of capped to total mRNA abundance in TPMs (Relative C/T) in WT or scd6Δedc3Δ cells plotted for all 5391 mRNAs, the 498 non-iESR mRNA_ up_s6,e3, or 429 non-rESR mRNA_down_s6,e3 transcripts dysregulated by scd6Δedc3Δ vs. WT. (D) Notched box-plots of the codon-protection index (CPI) for all mRNAs or for the sets of mRNAs up- or down-regulated by scd6Δedc3Δ vs. WT, including or excluding ESR transcripts, as indicated. (E) Notched box-plots showing log2 changes in absolute mRNA abundance from ERCC spike-in normalized RNA-Seq (left) or absolute Rpb1 occupancies averaged over the CDSs from S. pombe chromatin spike-in normalized Rpb1 ChIP-Seq (right) in scd6Δedc3Δ vs. WT cells for all mRNAs or the 554 or 526 non-ESR mRNAs up- or down-regulated, respectively, by scd6Δedc3Δ vs. WT. (F) Co-immunoprecipitation analysis of Dhh1-Dcp2 association in yeast cell extracts. Transformants of DHH1-TAP strains H5695 (WT), H5696 (edc3Δ), H5697 (scd6Δ), and H5698 (edc3Δ scd6Δ) and the parental untagged strains all harboring single-copy plasmid pAK133 expressing HA-tagged Dcp2 were cultured in SC-Ura at 30°C and whole cell extracts were incubated with IgG sepharose beads to purify Dhh1-TAP and associated proteins. Washed beads were eluted by boiling in SDS loading buffer and eluates and input extracts were resolved in parallel by SDS-PAGE and subjected to immunoblot analysis to detect the tagged proteins. Immune complexes were visualized with enhanced chemiluminescence.

Additional support that Scd6/Edc3 target mRNAs for decapping came from evidence that mRNAs up-regulated in the scd6Δedc3Δ mutant tend to accumulate in WT cells as decapped isoforms. Following decapping by Dcp1/Dcp2 mRNAs frequently undergo 5’ to 3’ decay co-translationally, with Xrn1 following behind the last translating ribosome loaded prior to decapping, and such decapped intermediates account for ∼12% of all mRNAs in WT cells (Pelechano et al. 2015). We reasoned that mRNAs preferentially targeted by Scd6/Edc3 for decapping and attendant degradation by Xrn1 should exhibit a greater than average proportion of decapped intermediates in WT cells. To test this, we conducted cap analysis of gene expression (CAGE) to quantify the abundances of all capped mRNA 5’ ends and compare them to total mRNA abundances determined by RNA-Seq conducted in parallel on biological replicates prepared from the WT and scd6Δedc3Δ strains described above. Transcript numbers per million reads (TPMs) from CAGE (C) and RNA-Seq (T) were determined and C/T ratios calculated as a proxy for the proportion of capped molecules for each transcript (Vijjamarri et al. 2023a) (File S1). (Because the CAGE and RNA-Seq data were normalized separately, the C/T ratios are relative, not absolute, proportions of capped transcripts.) Importantly, the C/T ratios are lower in WT cells for the 498 non-iESR mRNAs up-regulated in the scd6Δedc3Δ double mutant compared to all expressed mRNAs (Fig. 3C, cols. 1 & 5) in the manner expected for mRNAs preferentially targeted for decapping in WT cells. By contrast, the C/T ratios are higher than average for the non-rESR mRNAs that are down-regulated in relative abundance in scd6Δedc3Δ cells, as expected for an unusually low degree of decapping in WT (Fig. 3C, cols 3 & 5). Importantly, these C/T ratios were elevated in scd6Δedc3Δ cells to nearly the same level for all three groups of mRNAs, which exceeds the ratios found in WT cells, in the manner expected from eliminating Scd6/Edc3-stimulated decapping (Fig. 3C, cf. cols. 2,4,6). Similar results were obtained for these same groups of dysregulated mRNAs by analyzing our previous CAGE data obtained for the dhh1Δ mutant (Vijjamarri et al. 2023a) (Fig. S3E). These findings are consistent with the notion that impaired decapping dependent on Scd6/Edc3 and Dhh1 is an important driver of increased mRNA abundance in the scd6Δedc3Δ mutant.

Independent evidence for this last conclusion was provided by analyzing the codon protection indices (CPI) of mRNAs up-regulated by scd6Δedc3Δ, an indicator of co-translational decay by Xrn1. Decapped degradation intermediates exhibit three-nucleotide periodicity generated by precise Xrn1 cleavage up to the last translating ribosome at the 5’ end of mRNA, and the CPI quantifies the prevalence of such intermediates for each mRNA (Pelechano et al. 2015). Importantly, the mRNA_up_s6,e3 transcripts (whether including or excluding ESR mRNAs) exhibit higher than average median CPIs, indicating a greater than average involvement of decapping and co-translational degradation by Xrn1 in their decay, whereas mRNA_dn_s6,e3 transcripts exhibit lower than average CPI values, consistent with an alternative degradation pathway controlling their abundance scd6Δedc3Δ (Fig. 3D).

To determine whether increased transcription contributes to increased abundance of mRNAs up-regulated by scd6Δedc3Δ, we performed ChIP-Seq analysis of Rpb1 to measure RNA Polymerase II (Pol II) occupancies averaged across the CDS of every gene, obtaining highly reproducible results across replicates (Fig. S2D & S4A). To quantify absolute changes in Pol II occupancies, S. pombe chromatin was added to each S. cerevisiae chromatin sample prior to immunoprecipitation. To measure absolute changes in mRNA abundance, we re-analyzed the RNA-Seq data taking into account the recovery of External RNA Controls Consortium (ERCC) transcripts that were added to each total RNA sample prior to preparation of cDNA libraries, yielding highly reproducible normalized values among replicates (Fig. S2C). Interrogating the non-ESR mRNAs up- or down-regulated by scd6Δedc3 revealed a 2.79-fold increase in median normalized mRNA abundance for the non-iESR mRNA_up_s6,e3 group that was associated with only a 1.2-fold increase in normalized Rpb1 occupancies (Fig. 3E, cols. 1 & 4). The down-regulated non-rESR mRNA_dn_s6,e3 transcripts showed 12% and 7% decreases in spike-in normalized median mRNA abundance and Rpb1 occupancies, respectively, (Fig. 3E, cols. 2 & 5). Because the majority of total mRNA is rRNA, the ERCC normalization effectively yields the abundance of each mRNA relative to rRNA, and the ribosome content per cell is expected to be reduced in the scd6Δedc3Δ mutant owing to repression of rESR mRNAs encoding ribosomal proteins and biogenesis factors (Fig. S3C). We showed previously that the isogenic dcp2Δ mutant displays an ESR response of similar magnitude and a 30% reduction in ribosomal subunits per cell compared to same WT examined here (Vijjamarri et al. 2023b). Assuming a similar reduction in ribosome abundance in scd6Δedc3Δ cells, the absolute changes in mRNA per cell conferred by scd6Δedc3Δ are expected to be 0.7-fold of the ERCC-normalized values given in Fig. 3E, yielding fold-changes of 2.0 and 0.62 for the mRNA_up_s6,e3 and mRNA_dn_s6,e3 groups, respectively. Because these predicted changes in mRNA per cell still differ substantially from the corresponding changes in normalized Rpb1 occupancies of 1.2 and 0.93, respectively, there is only a small contribution of altered transcription to the altered abundance of transcripts dysregulated in the double mutant.

Our findings above that the scd6Δedc3Δ and dhh1Δ mutations up-regulate highly similar sets of mRNAs (Fig. 2C) is consistent with a previous proposal that recruitment of Dhh1 to Dcp2 can be mediated interchangeably by Edc3 or Scd6 bound to the same site in the Dcp2 CTT (He et al. 2022). To test this model we asked whether association of Dhh1 with Dcp2 in cell extracts is impaired only when both Edc3 and Scd6 are eliminated. Indeed, co-immunoprecipitation of HA-tagged Dcp2 with TAP-tagged Dhh1 was reduced in scd6Δedc3Δ cells but not in either single mutant in replicate experiments (Fig. 3F, lanes 2-4; Fig. S5(i)-(ii)). The residual association in the double mutant suggests that Dhh1 can interact with Dcp2 either directly or via Pat1 in addition to the interaction mediated by Scd6/Edc3, which might explain why a small fraction of mRNAs are up-regulated by dhh1Δ but not by scd6Δedc3Δ (Fig. 2C).

Slow rates of translation initiation or elongation generally do not dictate preferential decapping/decay by the decapping activators

Codon non-optimality has been linked with Dhh1-mediated mRNA decay partly by demonstrating that the sTAI values of mRNAs, which quantify their overall codon optimality (Sabi and Tuller 2014), are inversely correlated with the changes in mRNA abundance observed in dhh1Δ versus WT cells (Radhakrishnan et al. 2016). Similarly, analyzing the mRNA changes conferred by the scd6Δedc3Δ mutation for all non-ESR transcripts reveals a small but statistically significant negative correlation with sTAI values (Pearson r of -0.085, P = 2 x 10-9) similar to that observed for the dhh1Δ mutation (r = -0.071, P = 6 x 10-6) (Fig. S6A, cyan vs. orange), indicating a tendency for mRNAs with lower codon optimality/sTAI scores to show greater increases in relative abundance in response to scd6Δedc3Δ or dhh1Δ. However, the 591 non-iESR mRNAs up-regulated by the scd6Δ/edc3Δ mutations (from Fig. 1B) have a median sTAI value (0.35) nearly identical to that of all nonESR mRNAs, which is also the case for the group of 1018 mRNAs up-regulated by the pat1Δ or dhh1 mutations (Vijjamarri et al. 2023a) (Fig. S6B). Similar results were obtained for other metrics of codon optimality, tAI and average CSC (Fig. S6B), suggesting that poor codon optimality is not a key property defining the mRNAs repressed most extensively by these decapping activators. It has also been proposed that competition between translation initiation and mRNA decay, rather than codon optimality and elongation, is a major determinant of mRNA stability in yeast (Muhlrad et al. 1995; LaGrandeur and Parker 1999; Schwartz and Parker 1999; Chan et al. 2018). However, the mRNAs up-regulated in the scd6Δ/edc3Δ or pat1Δ/dhh1 mutants have slightly greater than average median translational efficiencies (TEs) in WT cells (Fig. S6C), as determined by ribosome profiling experiments described below. Thus, neither slow rates of translation initiation nor pausing during elongation at non-optimal codons appears to dictate preferential targeting of the mRNAs most strongly repressed by these four decapping activators.

Scd6 and Edc3 act redundantly and cooperate more extensively with Dhh1 than Pat1 in controlling translation of individual mRNAs

The changes in mRNA abundance were highly correlated with changes in RPF abundance for the scd6Δedc3Δ mutant compared to WT for all expressed transcripts, with an r value of 0.81 (Fig. S2F), providing strong mutual validation of the RNA-Seq and Ribo-Seq data for these strains. Nevertheless, the fact that this correlation is weaker than that observed between biological replicates of RNA-Seq or Ribo-Seq data (r values >0.95, Fig. S2A-B), suggests that certain mRNAs exhibit altered translational efficiencies in the double mutant. Indeed, DESeq2 analysis of the Ribo-Seq data identified 184 mRNAs showing TE increases of >1.41-fold at FDR < 0.10 in the scd6Δedc3Δ strain (dubbed TE_up_s6,e3) but only 42 and one such mRNA in the scd6Δ and edc3Δ single mutants, respectively (Fig. 4A), suggesting that Scd6 and Edc3 have overlapping functions in repressing the translation of particular mRNAs. Supporting this, a group of 169 mRNAs translationally up-regulated exclusively in the scd6Δedc3Δ double mutant exhibit only modest TE increases in the two single mutants (Fig. 4A-B, sector (iii)). SPI1, encoding a cell wall protein, is a representative transcript displaying increased mRNA abundance coupled with an even greater increase in RPF abundance to confer a TE increase of ∼3-fold only in the double mutant (Fig. 4C). As expected, the 28 mRNAs showing substantial TE increases only in scd6Δ cells show no increase in median TE in the edc3Δ strain (Fig. 4A-B, sector (i)), and thus appear to be translationally repressed by Scd6 alone.

Most mRNAs translationally up-regulated in scd6Δedc3Δ cells exhibit Scd6/Edc3 functional redundancy for repressing TE and show correlated changes in TE and protein abundance.

(A) Venn diagram of overlap between the 184 and 42 mRNAs in the TE_up groups identified in the scd6Δedc3Δ vs. WT or scd6Δ vs. WT comparisons, respectively. (B) Notched box-plots of log2ΔTE values between the indicated mutants vs. WT for the mRNAs belonging to sectors (i) or (iii) of the diagram in (A). (C) Gene browser image for SPI1 presented as in Fig. 1D, except also giving the TE changes for each mutant vs. WT on the lower right. (D) Density scatterplot of log2ΔRPF values measured by ribosome profiling vs. log2Δprotein values measured by TMT-MS for 4339 mRNAs for which data were obtained in both analyses, indicating the Pearson correlation coefficient (r) and P-value of the correlation. (E-F) Notched box-plots of log2Δprotein values from TMT-MS analysis between the scd6Δedc3Δ mutant vs. WT for the 843 and 839 mRNAs belonging to the Ribo_up or Ribo_down groups, respectively (D), or the 184 and 152 TE_up or TE_down mRNA groups (E) determined for the scd6Δedc3Δ mutant vs. WT, or for all mRNAs, for which TMT-MS data was obtained.

We explored next whether the increased ribosome occupancies observed in the scd6Δedc3Δ mutant are associated with increased protein synthesis by conducting TMT mass spectrometry (TMT-MS) of total cell proteins to obtain ratios of peptide abundance in the scd6Δedc3Δ strain vs. WT for >4000 different proteins (Fig. S2E). Importantly, significant correlations exist between changes in protein abundance and RPFs across the translatome (Fig. 4D). Moreover, mRNAs showing increased or decreased RPFs in scd6Δedc3Δ vs. WT cells (>1.5-fold, FDR<0.05, Ribo_up and Ribo_dn) likewise exhibit increased or decreased median protein abundances determined by TMT-MS (Fig. 4E). These results suggest that increased ribosome occupancies measured by Ribo-Seq, which could occur by increases in mRNA, TE, or both, are generally associated with increased synthesis of the encoded proteins in scd6Δedc3Δ cells.

It was possible that translational repression by Edc3/Scd6 generally occurs by slowing elongation, leading to increased ribosome densities (RPF/mRNA ratios, ie. calculated TEs) in WT cells and decreased TE values in scd6Δedc3Δ cells that would be associated with increased protein expression. In this scenario, changes in TEs would be inversely associated with changes in protein expression. Instead, mRNAs showing increased or decreased TEs in the double mutant (TE_up_s6,e3 and TE_dn_s6,e3 transcripts defined above) also exhibit increased or decreased protein expression in scd6Δedc3Δ vs. WT cells (Fig. 4F), implying that Scd6/Edc3 generally influence translation at the initiation step.

We previously identified 274 mRNAs whose TEs are up-regulated in the pat1Δ, dhh1Δ, or pat1Δdhh1Δ mutants by the same criteria employed here (>1.41-fold at FDR < 0.10) (Vijjamarri et al. 2023a). Interestingly, these mRNAs overlap significantly with those translationally up-regulated in the single or double scd6Δ/edc3Δ mutants (Fig. 5A). The 76 mRNAs common to both groups show larger TE increases in the scd6Δedc3Δ mutant compared to the 136 transcripts up-regulated exclusively in the scd6Δ/edc3Δ mutants (Figs. 5A-B, sector (ii) vs. sector (i)). These 76 mRNAs also show marked TE increases in the dhh1Δ, pat1Δ, and pat1Δdhh1Δ mutants (Fig. 5B(ii)), indicating that all four decapping activators are required for efficient translational repression of these mRNAs in WT cells. The 136 mRNAs significantly up-regulated only in the scd6Δ/edc3Δ mutants show a moderate increase in median TE in dhh1Δ cells, but little response to pat1Δ (Fig. 5B(i)). The 198 mRNAs up-regulated only in the pat1Δ/dhh1Δ mutants show only a small TE increase in the scd6Δedc3Δ strain and cumulative TE increases in the pat1Δdhh1Δ double mutant (Fig. 5B(iii)). In summary, similar to our findings for repression of mRNA abundance, efficient translational repression of certain mRNAs requires the combined functions of Pat1, Dhh1, and either Scd6 or Edc3, whereas other mRNAs are translationally repressed by either Edc3/Scd6 or Pat1 with appreciable contributions from Dhh1.

Most mRNAs translationally up-regulated in the scd6Δedc3Δ mutant are also translationally up-regulated by dhh1Δ or pat1Δ.

(A) Venn diagram of overlap between all 212 mRNAs translationally up-regulated by scd6Δ, edc3Δ, or scd6Δedc3Δ (defined in Fig. 4A) or all 274 mRNAs translationally up-regulated by dhh1Δ, pat1Δ, or pat1Δdhh1Δ vs. WT identified previously (Vijjamarri et al. 2023a). (B) Notched box-plots of log2ΔTE values between the indicated mutants vs. WT for the mRNAs belonging to the specified sectors of the diagram in (A). (C) Hierarchical clustering analysis of log2ΔTE values conferred by the indicated mutations vs. WT for 222 of the 336 mRNAs translationally up- or down-regulated in scd6Δedc3Δ vs. WT cells for which RNA-Seq and Ribo-Seq data were obtained in all five strains and with log2ΔTE values >-5 and <5 conducted as in Fig. 2C, including the Spearman coefficients (ρ) and P values for the indicated correlations. (D) Notched box-plots of log2ΔTE values between the indicated mutants vs. WT for the 54 mRNAs showing >2-fold TE increases conferred by both scd6Δedc3Δ and pat1Δdhh1Δ mutations vs. WT. (E) Relative Dhh1 occupancies from the Dhh1 RIP-seq experiments of Miller et al (2018) for the 212 and 274 mRNAs identified as TE_up in the scd6Δ, edc3Δ, or scd6Δedc3Δ mutants, or the dhh1Δ, pat1Δ, or pat1Δdhh1Δ mutants, vs. WT, respectively (cols 1-2), or for the 591 and 1018 mRNAs identified as mRNA_up in either the scd6Δ, edc3Δ, or scd6Δedc3Δ mutants, or the dhh1Δ, pat1Δ, or pat1Δdhh1Δ mutants, vs. WT, respectively (cols. 3-4).

Greater cooperation of Scd6/Edc3 with Dhh1 vs. Pat1 in controlling translation is also revealed by clustering analysis of TE changes in different mutants for the mRNAs exhibiting altered TEs in the scd6Δ/edc3Δ mutants, with greater similarity between TE changes conferred by scd6Δedc3Δ vs. dhh1Δ (ρ = 0.82) compared to scd6Δedc3Δ vs. pat1Δ (ρ = 0.68) or scd6Δedc3Δ vs. pat1Δdhh1Δ (ρ = 0.69) cells (Fig. 5C). However, as noted above, the group showing the strongest up-regulation of TEs in the scd6Δedc3Δ double mutant tends to be highly up-regulated in all three of the other mutants lacking Dhh1 or Pat1 (bracketed mRNAs at the top of Fig. 5C), as in Fig. 5D for a group of 54 mRNAs whose TEs are up-regulated by 2.0-fold or more in both the scd6Δedc3Δ and pat1Δdhh1Δ double mutants. Thus, although Scd6/Edc3 appear to cooperate more extensively with Dhh1 than Pat1 in translational control, the subset of 50-60 mRNAs exhibiting the strongest repression by Scd6/Edc3 or Pat1/Dhh1 require the concerted functions of Pat1, Dhh1, and either Scd6 or Edc3 to achieve their strong translational repression in WT cells. To assess whether mRNAs translationally repressed by the decapping activators are also targeted for degradation by these factors, we examined mRNA changes for the groups of mRNAs exhibiting TE up-regulation in the scd6Δ/edc3Δ or pat1Δ/dhh1Δ mutants (sectors (i) to (iii) of Fig. 5A). The mRNAs showing increased TEs only in the pat1Δ/dhh1Δ strains show little change in mRNA abundance in these mutants (Fig. S7A(iii)), indicating selective repression of translation vs degradation. Similarly, mRNAs translationally up-regulated only in the scd6Δ/edc3Δ mutants show little change in mRNA abundance in the scd6Δedc3Δ strain and only modest up-regulation in the pat1Δ/dhh1Δ mutants (Fig. S6A(i)). In contrast, the transcripts showing increased TEs in all of the mutants also show markedly increased abundance in the same mutants (Fig. S6A(ii)). Thus, coupled repression of translation and abundance by the decapping factors occurs only for the subset of transcripts exhibiting concerted translational repression by all four factors. In contrast, the much larger groups of transcripts defined above showing increased mRNA abundance in the mutants (described in Fig. 2A) exhibit little change in median TE in all four mutants (Fig. S6B), suggesting that most mRNAs targeted for enhanced turnover do not exhibit translational repression of the fraction of undegraded transcripts remaining in WT cells.

We next examined the translational efficiencies of mRNAs translationally repressed by Scd6/Edc3 by interrogating the Ribo-Seq data for WT cells on rich medium. The mRNAs translationally repressed by Scd6/Edc3 in concert with Pat1/Dhh1 tend to be poorly translated in WT, having a substantially lower median relative TE compared to all mRNAs (0.24 vs. 0.95) (Fig. S6C, col. 2), thus resembling the mRNAs translationally repressed exclusively by Dhh1/Pat1 (col. 3). In contrast, the transcripts translationally repressed exclusively by Edc3/Scd6 are generally well translated, exhibiting a ∼2.5-fold greater than average median TE in WT cells (Fig. S6C, col. 1). Consistent with this distinction, gene ontology (GO) analysis revealed that products of the 136 mRNAs translationally repressed exclusively by Scd6/Edc3 are enriched for cytoplasmic or mitochondrial ribosomal proteins (P = 1 x 10-7), whereas products of the 76 transcripts repressed by Scd6/Edc3/Pat1/Dhh1 are enriched for factors involved in fermentation (P = 2 x 10-7), carbohydrate metabolism (P = 1 x 10-6) or metabolism of non-protein amino acids (P = 3 x 10-6). Thus, Scd6/Edc3 translationally repress distinct groups of mRNAs depending on whether Dhh1/Pat1 participate in the repression.

We showed previously (Zeidan et al. 2018) that mRNAs up-regulated by dhh1Δ are enriched for Dhh1 protein association in vivo as judged by RIP-Seq analysis of yeast mRNAs using antibodies against Dhh1 (Miller et al. 2018), thus providing evidence for a direct role of Dhh1 decapping/degradation of these transcripts. The same observation was made for the mRNAs up-regulated by pat1Δ, consistent with widespread cooperation between Dhh1 and Pat1 in repressing mRNA abundance (Vijjamarri et al. 2023a). Interestingly, here we observed greater than average Dhh1 occupancies for all four mRNA groups defined above (in Figs. 2A or 5A) that are up-regulated in transcript abundance or TE by the scd6Δ/edc3Δ or dhh1Δ/pat1Δ mutations (Fig. 5E). This finding is consistent with the fact that Dhh1 contributes to repressing the abundance or TE of the majority of transcripts in these groups (Figs. 2B & 5B). It is possible that Dhh1 is recruited to most of these mRNAs in a complex with Dcp1:Dcp2 and other decapping activators (He et al. 2022), but contributes differentially to their degradation or translational repression.

Scd6/Edc3 post-transcriptionally repress mRNAs encoding enzymes for respiration and catabolism of non-glucose carbon sources in rich medium

Recently, we reported that mRNAs up-regulated in abundance or TE in the dcp2Δ, pat1Δ and dhh1Δ mutants are enriched for mitochondrial proteins that function in Ox. Phos. (Vijjamarri et al. 2023a; Vijjamarri et al. 2023b). GO analysis led to the same finding here for the non-iESR mRNAs up-regulated in abundance by the scd6Δ/edc3Δ mutations, including many of the same categories of mitochondrial functions enriched among the mRNAs up-regulated by dhh1Δ/pat1Δ mutations (Fig. S8A-B, green type). GO analysis of genes showing increased ribosome occupancies (RPFs), indicating either increased mRNA abundance or TE, confirms that the scd6Δ/edc3Δ mutations derepress the translation of Ox. Phos. gene transcripts (Fig. S8C-D). Examining a collection of Ox. Phos. genes functioning in electron transport, the TCA cycle or mitochondrial ATP synthase reveals that the scd6Δedc3Δ and pat1Δdhh1Δ double mutations up-regulate expression of these genes primarily at the level of mRNA abundance, with small additional increases in TE (Fig. 6A). Western blot analysis revealed increased expression of four Ox. Phos. proteins (Qcr8, Atp20, Idh1, and Sdh4), two proteins (Cox14 and Cox20) involved in cytochrome c oxidase assembly, and mitochondrial cytochrome b2 (Cyb2) required for lactate utilization in scd6Δedc3Δ cells, relative to Gcd6 examined as loading control (Fig. 6B-C). Results similar to these were obtained for the same mitochondrial proteins in the dhh1Δ and pat1Δ mutants (Vijjamarri et al. 2023a) and for a subset in dcp2Δ cells (Vijjamarri et al. 2023b). Importantly, we also observed increased expression of Cox2, a mitochondrially encoded subunit of cytochrome c oxidase, terminal enzyme of the mitochondrial ETC, whose expression correlates with mitochondrial activity (Vengayil et al. 2024). Except for dcp2Δ, all the decapping mutants had significantly increased Cox2 protein (Fig 6D), suggesting an increase in ETC activity in these mutants. Up-regulation of mRNA and RPF abundance in scd6Δedc3Δ and pat1Δdhh1Δ double mutants also occurred for enzymes of the glyoxylate cycle (Fig. 6E), which catalyze certain reactions of the TCA cycle in the cytoplasm to support gluconeogenesis during respiratory growth on two-carbon compounds; and the increased expression of one such enzyme, Cit2, was confirmed by Western analysis (Figs. 6B-C). Both Ox. Phos. and the glyoxylate cycle normally operate at low levels in yeast growing with abundant glucose, as in our experiments, suggesting that Scd6/Edc3 cooperate with Dhh1/Pat1 to help suppress these pathways in glucose-replete cells.

Consistent with post-transcriptional control of Ox. Phos. genes, the median relative RPF levels were up-regulated in the scd6Δedc3Δ mutant substantially more than the increases in relative Pol II occupancies observed at the cognate genes by ChIP-Seq analysis of Rpb1 (Fig. S9B, cols. 1-2). Moreover, the transcription factors responsible for induction of Ox. Phos. genes, the Hap2/Hap3/Hap4/Hap5 complex, were not activated in scd6Δedc3Δ cells: expression of a CYC1-lacZ reporter activated by this complex (Forsburg and Guarente 1989) was not elevated in the scd6Δ/edc3Δ mutants in glucose-containing medium (Fig. 6F(ii)) but showed the expected induction in WT cells grown with glycerol/ethanol versus glucose (6F(i)) (Broach 2012). These findings support the notion that reduced decapping stabilizes Ox. Phos. gene transcripts in scd6Δedc3Δ cells.

In addition to Ox. Phos., GO analysis of genes up-regulated in mRNA or RPF abundance in the scd6Δ/edc3Δ mutants revealed enrichment for utilization of alternative carbon sources (Fig. S8A-D), as observed recently for dhh1Δ,pat1Δ mutants (Vijjamarri et al. 2023a). Consistent with this, 83 carbon catabolite down-regulated (CCR) genes, known to be glucose-repressed or activated by transcription factors Adr1 or Cat8 (Young et al. 2003; Tachibana et al. 2005), exhibit increased translation (RPFs) largely through increased mRNA abundance in both scd6Δedc3Δ and pat1Δdhh1Δ double mutants (Fig. S9A). These genes encode enzymes for β-oxidation of fatty acids in addition to the glyoxylate cycle, which allow cells to synthesize precursors that feed into gluconeogenesis or amino acid biosynthesis, or produce acetyl-CoA and generate NADH by respiration on non-fermentable carbon sources (Young et al. 2003). The CCR genes exhibit larger increases in RPFs compared to Pol II occupancies at the cognate genes in scd6Δedc3Δ vs. WT cells (Fig. S9B, cols. 3-4), consistent with post-transcriptional repression by Scd6/Edc3. Supporting this, an ADH2-lacZ reporter transcriptionally induced by activated Adr1 (Sloan et al. 1999) displayed the expected large induction in our WT strain cultured with glycerol/ethanol versus glucose as carbon source (Fig. S9C(i)), but was down-regulated by scd6Δedc3Δ in glucose-grown cells (Fig. S8C(ii)).

Functional evidence that Scd6/Edc3 repress oxidative phosphorylation

We examined the effects of eliminating Scd6/Edc3 on mitochondrial electron transport by measuring mitochondrial membrane potential (ΔΨm) generated by the ETC using the probe tetramethylrhodamine (TMRM)—a cationic fluorescent dye that accumulates in mitochondria as a function of ΔΨm. Quantifying dye fluorescence by flow cytometry revealed increased TMRM fluorescence in the scd6Δedc3Δ mutant containing an empty vector compared to both the isogenic WT strain and the mutant complemented by WT EDC3 (Fig. 6G). These results are consistent with increased mitochondrial ETC activity in glucose-grown scd6Δedc3Δ cells.

Scd6/Edc3 post-transcriptionally repress proteins involved in respiration and suppress mitochondrial membrane potential in rich medium.

(A) Log2 changes in mRNA, RPFs, or TE conferred by the indicated double mutations vs. WT for 53 nuclear genes encoding mitochondrial proteins involved directly in oxidative phosphorylation. (B-C) Western blot analysis of 9 mitochondrial proteins and Gcd6 (examined as loading control) in WT and scd6Δedc3Δ strains, cultured in duplicate in YPD medium to OD600 of ∼0.6-0.8. WCEs were extracted under denaturing conditions and aliquots corresponding to 1X or 2X amounts of WCE were loaded in successive lanes for the two replicate cultures. Immune complexes were visualized with enhanced chemiluminescence (B). Signals for each protein were quantified, normalized to the corresponding signals for Gcd6 in the same extract and expressed relative to the resulting values for WT cells. Mean values and standard errors are plotted (C). (D) Western blot analysis of Cox2 in strains of the indicated genotypes in cells cultured as in (B). Cox2 signal intensity was normalized to total Coomassie-stained protein and the resulting relative Cox2 protein levels from three biological replicates were averaged and plotted. (E) Log2 changes in mRNA, RPFs, or TE conferred by the indicated double mutations vs. WT for 22 genes encoding enzymes of the glyoxylate cycle or fatty acid metabolism. (F) Expression of the CYC1-lacZ reporter on plasmid pLG265, lacking UAS1 and containing the optimized version of UAS2, UAS2UP1, in the WT strain grown on SC-Ura medium containing either 2% glucose or 3% glycerol/2% ethanol as carbon sources (i), or in WT and the indicated mutant strains on SC-Ura with 2% glucose (ii). β-galactosidase activity (nmoles of o-nitrophenyl-β-D-galactopyranoside (ONPG) cleaved per min per mg of total protein) was measured in whole cell extracts for 3 biological replicates of each strain and the mean values were normalized to the mean activity measured in WT grown with glucose as carbon source. **, P-value <0.01 from student’s t-test; ns, not significant. (G) Measurements of mitochondrial membrane potential. WT cells or transformants of the scd6Δedc3Δ mutant containing the EDC3 plasmid pLfz614-7 or empty vector were cultured in SC-Ura to mid-log phase. TMRM (500 nM) was added and incubated for 30 min before samples were collected and washed once with deionized water. ΔΨm was determined by measuring TMRM fluorescence intensity using flow cytometry. Data are presented in arbitrary fluorescence intensity units per OD600. 2-way ANOVA was used for statistical analysis and data are given as mean values ± SD (n=3) (****p<0.0001).

To determine whether up-regulation of Ox. Phos. and other glucose-repressed genes in the decapping activator mutants alters cellular metabolites, we used targeted, quantitative LC-MS/MS based approaches (see Methods) to quantify levels of 147 polar metabolites in the isogenic scd6Δedc3Δ, dhh1Δ, pat1Δ, dcp2Δ and WT strains described above, cultured in YPD medium. Principle component analysis revealed clustering of results from biological replicates in the manner expected for reproducible differences in metabolite levels among different strains, with results for the scd6Δedc3Δ mutant most closely resembling those for dhh1Δ, which in turn were more similar to the results for pat1Δ vs. the dcp2Δ mutant or WT strains (Fig. 7A). This conclusion was borne out by cluster and correlation analyses of changes in metabolites between each mutant compared to WT, with the strongest correlation observed for scd6Δedc3Δ vs. dhh1Δ, followed by dhh1Δ vs. pat1Δ (Fig. 7B). Considering the subset of 46 metabolites up-regulated in any two of the four mutants again showed greatest similarity between changes conferred by scd6Δedc3Δ vs. dhh1Δ followed by dhh1Δ vs. pat1Δ (Fig. 7C). These findings mirror the results from RNA-seq and ribosome profiling in which the up-regulation of mRNA levels or translation was most similar between the scd6Δedc3Δ and dhh1Δ mutants. Pathway analysis of the 46 up-regulated metabolites (at https://www.metaboanalyst.ca/MetaboAnalyst/Secure/pathway) revealed a significant enrichment for metabolites of both the TCA and glyoxylate cycles (Fig. 7D), with five of the six TCA cycle intermediates detected (fumarate, malate, α-ketoglutarate, cis-aconitate and citrate) being elevated in the decapping mutants (Fig. 7E). These results are consistent with the possibility that up-regulation of Ox. Phos. proteins (Fig. 6A-D) and ETC function (Fig. 6F) in the decapping mutants leads to increased flux from glucose towards the TCA cycle.

Eliminating decapping activators or decapping enzyme confers similar changes in polar metabolites.

(A) Principal component analysis of the levels of 147 metabolites in biological replicates of each strain. (B-C) Hierarchical clustering analysis of log2 changes in all 147 metabolites analyzed (B) or the 46 metabolites up-regulated in any two of the four mutants (C) conferred by the indicated mutations vs. WT, including the Spearman coefficients (ρ) and P values for the indicated correlations. (D) Results of pathway analysis of the 46 up-regulated metabolites described in (C) conducted at https://www.metaboanalyst.ca/MetaboAnalyst/Secure/pathway. Red ovals depict groups of metabolites significantly enriched among the set of 46 compounds, with P-value <0.05. (E-F) Log2 changes in levels of TCA cycle intermediates (E) or amino acids (F) conferred by the indicated mutations vs. WT.

To confirm increased flux through the TCA cycle coming from glucose breakdown, WT and mutant cells were grown in high glucose, pulsed with 13C6 glucose, and relative 13C label incorporation into TCA cycle intermediates was estimated for each carbon molecule derived from glucose (as indicated schematically in Fig. 8A) eight min following the pulse. In all mutants, 13C label incorporation in all TCA cycle intermediates was significantly higher than in WT cells (Fig. 8A), demonstrating increased carbon flux from glucose into the TCA cycle. These results, along with increased ETC activity (Fig. 6G), indicate that mitochondrial function is up-regulated in these mutants on glucose. If so, we might expect them to show increased ATP production from respiration versus glycolysis. Measuring total ATP levels in glucose-grown cells revealed that, while all decapping mutants except pat1Δ showed reduced ATP per cell (Fig. 8B), the proportion of ATP produced from respiration, and thus eliminated by sodium azide treatment (by inhibiting the ETC), was elevated in all of the mutants (Fig. 8C).

Elimination of decapping activators or decapping enzyme up-regulates respiration, increasing flux from glucose into TCA cycle intermediates and proportion of ATP produced by Ox. Phos.

(A) Three biological replicates of cells of each indicated genotype were cultured in YP with 2% unlabeled glucose, shifted to YP with 1% unlabeled glucose for 20 min, and pulsed with 13C6 labelled glucose (at final concentration of 1%) for 8 min, followed by extraction of metabolites and quantification of the indicated TCA cycle intermediates by mass spectrometry. In the diagrams, green circles signify the labelled carbon atom and the notation M+1, M+2, etc. indicate the mass increase in the molecules due to the labelled carbon. The depicted labelling pattern of metabolites reflects on cycle of the TCA cycle, resulting in mass additions of M+1 and M+2; however, across multiple cycles, a broader range of metabolite species with different mass additions will emerge. The metabolite signal intensities in all samples are expressed relative to that determined for the first replicate of the WT strain. (B-C) Measurements of ATP levels and proportions of total ATP impaired by azide inhibition of ETC activity. Cells cultured in YPD were treated or untreated with sodium azide for 30 min prior to harvesting. ATP levels were determined in extracts and normalized to OD600 units of cells for three biological replicates each of treated and untreated cell aliquots. Mean values for each strain are plotted in (B) and relative fractions of ATP in untreated samples retained following azide treatment (ATP_untreated)-(ATP_Azide)/ATP_untreated), normalized to the values determined for WT are plotted in (C) with results from a student’s t-test indicated with asterisks: **, P<0.005; *, P<0.05.

Interestingly, amino acids are also up-regulated in the decapping mutants (Fig. 7D), particularly in pat1Δ cells (Fig. 7F), as well as intermediates in amino acid biosynthesis and amino acid derivatives (Fig. 7D). The expression of amino acid biosynthetic enzymes is unaffected or reduced in the mutants however (Fig. S9D), suggesting that increased amino acid abundance results from metabolic reprogramming rather than increased biosynthetic capacity. One possibility is that accumulation of the TCA cycle intermediate α-ketoglutarate leads to increased production of glutamate and glutamine, precursors in all amino acid biosynthetic pathways (Ljungdahl and Daignan-Fornier 2012). The two 3-phosphotrioses generated in glycolysis (glyceraldehyde 3-phosphate and dihydroxyacetone phosphate) are also elevated in the mutants, possibly owing to increased glyoxylate shunt function, which might stimulate synthesis of amino acids serine and glycine (Ljungdahl and Daignan-Fornier 2012) (Fig. 7D, Methane metabolism category). The increases in fructose 1,6-bisphosphate, fructose 6-phosphate, glucose 6-phosphate, and UDP-glucose in all four mutants (Fig. 7D, Starch/Sucrose metabolism category) all suggest substantial metabolic rewiring indicative of glucose up-regulation and usage of alternative carbon sources. In addition to amino acids, pyrimidine nucleotides are up-regulated in the mutants (Fig. 7D, Pyrimidine metab.), which might be driven by the elevated glutamine levels or increased glycolytic flux towards the pentose phosphate pathway (Ljungdahl and Daignan-Fornier 2012).

Discussion

RNA-seq analysis of single and double scd6Δ and edc3Δ mutants has revealed that Scd6 and Edc3 have largely overlapping functions in repressing mRNA abundance, as the presence of either protein alone is sufficient for nearly WT levels of most mRNAs found up-regulated in the double mutant (Fig. S3A & Fig. 1B-C(ii)-(iii)). Although the mRNAs dysregulated in the scd6Δedc3Δ mutant are enriched for ESR mRNAs; the majority are not ESR transcripts (Fig. 1A-B) and are most likely repressed more directly by Scd6/Edc3. Functional redundancy of Scd6/Edc3 is consistent with their similarities in sequence and domain structure, including shared FDF motifs that interact competitively with Dhh1 in animals(Tritschler et al. 2008; Tritschler et al. 2009) and N-terminal LSm domains that compete for binding motifs in the Dcp2 CTT of S. pombe (Fromm et al. 2012), as well as the synthetic growth defect produced by deleting both genes simultaneously (Decourty et al. 2008) (Fig. S1A). Interestingly, a set of 37 mRNAs appears to be repressed by Edc3 exclusively (Fig. 1B-C(i)), including the canonical Edc3 targets identified previously, YRA1 and RPS28B (He et al. 2022), and is enriched for genes belonging to the GO category “mitochondrion” (22/37 genes) and the Ox. Phos. categories of electron transport and mitochondrial ATP synthesis (11/37 genes: ATP3 ATP16 ATP5 TIM11 ATP17 COX4 QCR9 CYC1 COX12 ATP18 COX5A). The fact that deleting SCD6 in the double mutant frequently diminishes the up-regulation of these transcripts conferred by deleting EDC3 alone (Fig. 1C(i)) might indicate that eliminating both Scd6 and Edc3 together enables a distinct degradation pathway that compensates for loss of Edc3-stimulated decapping/decay.

Several lines of evidence support the conclusion that most of the changes in mRNA abundance in the scd6Δedc3Δ double mutant result from impaired decapping and attendant 5’-3’ degradation by Xrn1. ChIP-seq analysis of Rpb1 shows that Pol II occupancies in the coding sequences increase by much smaller amounts compared to the increased transcript levels for non-iESR mRNAs up-regulated in scd6Δ/edc3Δ cells (Fig. 3E), indicating a minor contribution of increased transcription to their up-regulation. The transcriptional activators of Ox-Phos. and CCR genes, many of whose transcripts are up-regulated by scd6Δ/edc3Δ, remain largely inert in this mutant, ruling out inappropriate transcriptional induction of their target genes in glucose-replete scd6Δ/edc3Δ cells. Consistent with mRNA turnover via decapping, most of these transcripts are up-regulated similarly by scd6Δedc3Δ and deletion of DCP2 (Fig. 3A-B), exhibit a heightened proportion of decapped isoforms in WT cells that is diminished by deleting SCD6/EDC3 or DHH1 (Fig. 3C & Fig. S3E), and show evidence of co-translational 5’-3’ decay of decapped intermediates by Xrn1 (Fig. 3D); none of which was observed for the mRNAs down-regulated in scd6Δ/edc3Δ cells. Similar findings were reported previously for sets of mRNAs up-regulated by dhh1Δ or pat1Δ (Vijjamarri et al. 2023a), consistent with the widespread involvement of Dhh1 and Pat1 in controlling the levels of mRNAs preferentially targeted by Scd6/Edc3.

Supporting this last assertion, most (∼70%) of the mRNAs repressed in abundance by Edc3/Scd6 are also repressed by Dhh1 and Pat1, and their efficient repression in WT cells involves independent contributions of similar magnitude by Pat1, Dhh1 and either Edc3 or Scd6 (Fig. 2A-B(ii)). Another large set of mRNAs is repressed predominantly by Pat1 with lesser contributions by Dhh1 and Scd6/Edc3 (Fig. 2A-B(iii)). Our finding of extensive cooperation among these four decapping factors is consistent with our recent finding that ∼55% of the mRNAs up-regulated in the dcp2Δ mutant, and thus targeted by Dcp2 for enhanced degradation, tend to be up-regulated in the pat1Δdhh1Δ and scd6Δedc3Δ mutants, whereas the remaining 45% are generally up-regulated by upf1Δ instead. This finding suggested a major bifurcation of Dcp2-activation by either Pat1/Dhh1/Scd6/Edc3 or the Upf factors responsible for NMD (Vijjamarri et al. 2023b).

Closer examination of the effects of individual mutations on mRNA levels revealed a greater overlap between the mRNAs dysregulated by dhh1Δ and scd6Δ/edc3Δ versus those altered by pat1Δ (Fig. 2C-E), consistent with the aforementioned FDF motifs in Scd6 and Edc3 that interact with Dhh1 in animal systems, and with evidence for distinct complexes of the decapping enzyme containing Dhh1 and Edc3 or Scd6 but lacking Pat1. It also supports the recent suggestion that Edc3 and Scd6 act interchangeably to recruit Dhh1 to the Edc3-interaction site in the Dcp2 CTT to stimulate turnover of several Dhh1 target mRNAs (He et al. 2022). Our RNA-seq results support this last proposal in showing that all four Dhh1 target mRNAs examined in that study (EDC1, SDS23, HXT6, and HSP12) are up-regulated in the scd6Δedc3Δ double mutant while changing little in the single mutants (Fig. S10), two of which (SDS23 and HXT6) were shown previously to have longer half-lifes in dhh1Δ vs. WT cells (He et al. 2018). Finding that selective up-regulation in the scd6Δedc3Δ double mutant applies to most mRNAs up-regulated in dhh1Δ cells (Fig. 2E), we propose that redundant Scd6/Edc3 targeting is a widespread mechanism of Dhh1-enhanced mRNA degradation in the yeast transcriptome (Fig. S11). Supporting this, our co-immunoprecipitation analysis revealed diminished Dhh1-Dcp2 association in cell extracts of the scd6Δedc3Δ double mutant but for either single mutant (Fig. 3F).

Examining our Ribo-Seq and TMT-MS data suggests that Dhh1 translation and steady-state abundance are increased ∼2-fold in the scd6Δedc3Δ strain, indicating that up-regulation of many of the same mRNAs by scd6Δedc3Δ and dhh1Δ does not result indirectly from reduced levels of Dhh1 in the scd6Δedc3Δ mutant. The increased Dhh1 expression might signify a compensatory response to the absence of Scd6/Edc3. We also observed an ∼40% reduction in Dcp2 translation and abundance in the scd6Δedc3Δ strain, which might contribute to the up-regulation of mRNAs dysregulated in this mutant. However, our immunoblot analyses revealed no significant reduction in steady-state Dcp2 levels in scd6Δedc3Δ cells (Input lanes in Figs. 3F and S4C(i)-(ii)). Moreover, our previous finding that the majority of mRNAs subject to NMD, up-regulated by both upf1Δ and dcp2Δ, are not up-regulated by scd6Δedc3Δ (Vijjamarri et al. 2023b) implies that Dcp2 abundance in scd6Δedc3Δ cells is adequate for normal levels of NMD. Accordingly, we favor a direct role for Scd6/Edc3 in accelerating degradation of most transcripts up-regulated in the scd6Δedc3Δ mutant.

Despite strong evidence for the involvement of decapping in mRNA turnover directed by either Scd6/Edc3 or Pat1/Dhh1, it appears that neither enrichment for non-optimal codons nor low translational efficiencies are principal determinants of preferential targeting of mRNAs by these proteins (Fig. S6A-C). There are probably exceptions to these generalizations among the hundreds of transcripts repressed by these factors; nor can we rule out the involvement of clusters of non-optimal codons in transcripts of average overall codon optimality that generate a queue of slowly elongating ribosomes that triggers decapping. Nevertheless, it seems likely that other sequences or properties of the transcripts most strongly repressed by Scd6/Edc3 or Dhh1/Pat1 frequently underlie their preferential targeting by these factors for decapping/decay. Sequence-specific RNA-binding proteins that actively recruit these decapping activators is an attractive potential mechanism (Fig. S11). For example, there is evidence that Dhh1 is recruited to mRNAs in association with the Ccr4-NOT deadenylase via RNA-binding protein Puf5 (Goldstrohm et al. 2006). Involvement of RNA-binding proteins would also be consistent with the observed specific targeting of mRNAs whose expression is repressed in response to nutrient abundance.

Ribosome profiling of the single and double scd6Δ and edc3Δ mutants provided evidence that Scd6 and Edc3 also have highly overlapping functions in repressing the translation of ∼200 mRNAs, whose TE values are up-regulated substantially only in the double mutant (Fig. 4A-B(iii)). The observation that changes in TE (RPF density) generally correlate with changes in protein abundance conferred by scd6Δedc3Δ (Fig. 4E) implies that Scd6/Edc3 generally repress translation at the stage of initiation rather than elongation. Many mRNAs translationally repressed by Scd6/Edc3 additionally require Dhh1, and the subset most strongly repressed also requires Pat1 for efficient translational repression (Fig. 5B(ii) and 5D). This last group of mRNAs is translated at very low levels in WT cells on rich medium (Fig. S7C), which might reflect the concerted action of all four decapping activators in suppressing translation. The transcript abundance of these mRNAs is also up-regulated in each of the mutants (Fig. S7A(ii)), indicating enhanced degradation coupled with translational repression by the four decapping activators. One possibility is that a complex of Dcp1:Dcp2 containing Scd6/Edc3, Pat1/Dhh1, or different combinations of these factors (He et al. 2022), is recruited to these mRNAs and impedes recruitment of PICs to diminish translation in parallel with stimulating decapping and 5’-3’ degradation by Xrn1 (Fig. S11(i) & (ii)-(iv)).

The additional sets of mRNAs identified here that are translationally repressed exclusively by Scd6/Edc3 or Pat1/Dhh1 (Fig. 5A(i) & (iii)) show much less evidence of transcript degradation promoted by these factors (Fig. S7A(i) & (iii)), suggesting that inhibition of translation initiation occurs without decapping, or involves decapping uncoupled from degradation by Xrn1 (Fig. S10 (ii)-(iii)). In the latter scenario, the decapped mRNAs would be unable to recruit the cap-binding initiation factors necessary for PIC recruitment and hence persist in WT cells as translationally-inert degradation intermediates. Eliminating the decapping activators would diminish decapping and eliminate these fractions of poorly translated uncapped isoforms to confer the increased TE values we observed for the cognate genes in the decapping mutants. This hypothetical mechanism of translational repression via decapping is consistent with the lower than average capped to total mRNA (C/T) ratios found in WT cells for the mRNAs translationally repressed by all four decapping activators or by Pat1/Dhh1 (Fig. S6D (ii)-(iii)), indicating accumulation of decapped isoforms in WT cells, as observed previously for a subset of mRNAs translationally repressed by Dhh1 (Vijjamarri et al. 2023b). In contrast, the mRNAs that are translationally repressed exclusively by Scd6/Edc3 display higher than average C/T ratios in WT (Fig. S7D(i)), which is more compatible with an inhibition of PIC recruitment independently of decapping (Fig. S11(iii)). Translational repression without decapping could also occur if transcripts bound by decapping complexes are sequestered from the translation initiation machinery in RNA granules where mRNA decapping occurs inefficiently (Fig S11(iv)). Finally, it is important to note that mRNA decay occurs apparently without translational control for the much larger groups of ∼600 and ∼1000 transcripts showing up-regulated mRNA abundance but little evidence of TE changes in the scd6Δedc3Δ and pat1Δdhh1Δ mutants, respectively (Fig. S7B). It is unclear whether these mRNAs escape translational repression and are targeted exclusively for decapping/degradation (Fig. S11(i)) or are simply degraded too fast to allow translational repression to be detected by ribosomal profiling of steady-state mRNAs.

Recently, we reported that Pat1 and Dhh1 functionally collaborate with the decapping enzyme to repress genes and cellular pathways normally repressed in rich medium (Vijjamarri et al. 2023a; Vijjamarri et al. 2023b). Here we found evidence that Scd6 and Edc3 broadly cooperate with Pat1/Dhh1 in this control, acting redundantly to repress the abundance, translation, or both, of mRNAs encoding numerous proteins involved in respiration or utilization of alternative carbon sources (Figs. 6A-C, S8A-D), without affecting transcription of the cognate genes Fig. 6F, S9A-C). Evidence that up-regulation of Ox. Phos. mRNAs/proteins in the scd6Δedc3Δ mutant is sufficient to increase respiration in glucose-replete cells came from our findings of increased mitochondrially-encoded subunit II of ETC component cytochrome c oxidase (Fig. 6D) and elevated mitochondrial membrane potential (Fig. 6G) in the scd6Δedc3Δ mutant, extending similar findings we made previously for dcp2Δ, dhh1Δ and pat1Δ mutants (Vijjamarri et al. 2023a; Vijjamarri et al. 2023b). Consistent with this, our metabolomics analysis revealed elevated steady-state levels of TCA cycle intermediates in all four decapping mutants (Fig. 7D), and increased 13C carbon incorporation from 13C6-glucose breakdown into TCA cycle intermediates in the same mutant strains (Fig. 8A). Analysis of ATP levels further revealed increased proportions of ATP derived from Ox. Phos. in the decapping mutants (Fig. 8B), which together provide strong evidence that mitochondrial respiration is inappropriately elevated in glucose-replete cells of decapping activator mutants. It is well established that the enzymes involved in respiration and catabolism of non-glucose carbon sources are repressed at the transcriptional level in glucose-replete yeast (Zaman et al. 2008). Our findings demonstrate that Scd6/Edc3 collaborate with Dhh1/Pat1 to provide an additional layer of post-transcriptional control of these enzymes through enhanced decapping/decay or translational repression to help ensure their complete suppression when glucose is abundant.